---
editor_options: 
  markdown: 
    wrap: 72
---

# Introduction

[Data mining](https://en.wikipedia.org/wiki/Data_mining) has the goal of
finding patterns in large data sets. The popular data mining textbook
[*Introduction to Data
Mining*](https://www-users.cs.umn.edu/~kumar001/dmbook/) [@Tan2018]
covers many important aspects of data mining. This companion contains
annotated R code examples to complement the textbook. To make following
along easier, we follow the chapters in data mining textbook which is
organized by the main data mining tasks:

2.  [Data] covers types of data and also includes data preparation and
    exploratory data analysis in the chapter.

3.  [Classification: Basic Concepts and Techniques] introduces decision
    trees, model training and evaluation.

4.  [Classification: Alternative techniques] introduces and compares
        methods including rule-based classifiers, nearest neighbor
        classifiers, naive Bayes classifier, logistic regression and
        artificial neural networks.

5.  [Association Analysis: Basic Concepts and Algorithms] covers
        frequent itemset and association rule generation and analysis
        including visualization. Some code covers also topics discussed
        in the textbook's Chapter 6.

6.  [Association Analysis: Advanced Concepts] covers categorical and continuous
        attributes, concept hierarchies, and frequent sequence pattern mining.

7.  [Cluster Analysis] discusses clustering approaches including
        k-means, hierarchical clustering, DBSCAN and how to evaluate
        clustering results.

For completeness, we have added chapters on [Regression\*] and on
[Logistic Regression\*]. Sections in this book followed by an asterisk
contain code examples for methods that are not described in the data
mining textbook.

This book assumes that you are
familiar with the basics of R, how to run R code, and install packages.
The rest of this chapter will provide an overview and point you to where 
you can learn more about R and the used packages.

## Used Software

To use this book you need to have
[R](https://cran.r-project.org/) and [RStudio
Desktop](https://www.rstudio.com/products/rstudio/) installed.

Each book chapter will use a set of packages that must be installed. The
installation code can be found at the beginning of each chapter. Here is
the code to install the packages used in this chapter:

```{r setup_01, message=FALSE}
pkgs <- c('tidyverse')

pkgs_install <- pkgs[!(pkgs %in% installed.packages()[,"Package"])]
if(length(pkgs_install)) install.packages(pkgs_install)
```

```{r setup_01-2, include=FALSE}
source("_common.R")
source("format_pkgs.R")
all_pkgs <- union(all_pkgs, pkgs)
```

The code examples in this book use the R package collection `tidyverse` 
[@R-tidyverse] to manipulate data. Tidyverse also includes the package 
`ggplot2` [@R-ggplot2] for visualization. Tidyverse packages make
working with data in R very convenient. Data analysis and data mining 
reports are typically done by creating 
R Markdown documents. Everything in R is built on top of the 
core R programming language and the packages that are automatically 
installed with R. This is referred to as Base-R.

## Base-R

Base-R is covered in detail in [An Introduction to
R](https://cran.r-project.org/doc/manuals/r-release/R-intro.html). The
most important difference between R and many other programming languages
like Python is that R is functional (with some extensions) and
vectorized.

### Vectors

Scalars do not exist in R. The basic data structure is a vector of real
numbers. Special values like infinity and missing (`NA`) can be used.

```{r }
x <- c(10.4, 5.6, Inf, NA, 21.7) # c combines values
x
y <- seq(from = 0, to = 40, length.out = 5) # create a sequence
y
```

### Vectorized Operations

Operations are vectorized so loops are typically not necessary.

```{r }
x + 1
```

Comparisons are performed element wise and return a `logical` vector
(R's name for Boolean).

```{r }
x > y
```

### Subsetting Vectors

We can select vector elements using the `[` operator like in other
programming languages but the index starts at 1.

```{r}
x[1:3] # elements 1 through 3
x[-1] # all but the first
x[!is.na(x)] # all but the missing values

```

We can also assign values to a selection. For example, to get rid of
infinite values.

```{r}
x[!is.finite(x)] <- NA
x
```

### Functions

Functions work like in many other languages. However, they also operate
vectorized and arguments can be specified positional or by named. An
increment function can be defined as:

```{r }
inc <- function(x, by = 1) { 
    x + by 
  }
```

Calling the increment function on vector `x`.

```{r }
inc(x, by = 2)
```

R has many built-in functions like `min()`, `max()`, `mean()` and
`sum()`.

### Strings

R uses `character` vectors where `character` stands for a string and not
like in other programming language for a single character. R accepts
double or single quotation marks for strings.

```{r }
string <- c("Hello", "Goodbye")
string
```

Strings can be combined using `paste()`.

```{r}
paste(string, "World!")
```

Note that paste is vectorized and the string `"World!"` is used twice to
match the length of `string`. This behavior is called in R *recycling*
and works if one vector's length is an exact multiple of the other
vector's length. The special case where one vector has one element is
particularly often used. We have used it already above in the expression
`x + 1` where one is recycled for each element in `x`.

### Plotting

Basic plotting in Base-R is done by calling `plot()`.

```{r }
plot(x, y)
```

Other plot functions are `pairs()`, `hist()`, `barplot()`. In this book,
we will focus on plots created with the ggplot2 package.

### Objects

Other often used data structures include `list`, `data.frame`, `matrix`,
and `factor`. In R, everything is an object. Objects are printed by
either just using the object's name or put it explicitly into the
`print()` function.

```{r }
x
```

For many objects, a summary can be created.

```{r }
summary(x)
```

Objects have a class.

```{r }
class(x)
```

Some functions are generic, meaning that they do something else
depending on the class of the first argument. For example, `plot()` has
many methods implemented to visualize data of different classes. The
specific function is specified with a period after the method name.
There also exists a default method. For plot, this method is
`plot.default()`. Different methods may have their own manual page, so
specifying the class with the dot can help finding the documentation.

It is often useful to know what information it stored in an object.
`str()` returns a humanly readable string representation of the object.

```{r}
str(x)
```

There is much to learn about R. It is highly recommended to go through
the official
[An Introduction to
R](https://cran.r-project.org/doc/manuals/r-release/R-intro.html) manual.
There is also a good [Base R Cheat Sheet](https://github.com/rstudio/cheatsheets/blob/main/base-r.pdf)
available.

## R Markdown

R Markdown is a simple method to include R code inside a text document
written using markdown syntax. Using R markdown is especially convention to analyze
data and compose a data mining report. RStudio makes creating and
translating R Markdown document easy. Just choose
`File -> New File -> R Markdown...` RStudio will create a small
demo markdown document that already includes examples for code and how
to include plots. You can switch to visual mode if you prefer a What You
See Is What You Get editor.

To convert the R markdown document to HTML, PDF or Word, just click the
Knit button. All the code will be executed and combined with the text
into a complete document.

Examples and detailed documentation can be found on [RStudio's R Markdown
website](https://rmarkdown.rstudio.com/) and in the [R Markdown Cheatsheet](https://rstudio.github.io/cheatsheets/html/rmarkdown.html).

## Tidyverse

`tidyverse` [@R-tidyverse] is a collection of many useful packages that
work well together by sharing design principles and data structures.
`tidyverse` also includes `ggplot2` [@R-ggplot2] for visualization.

In this book, we will use

-   often tidyverse tibbles to replace R's built-in data.frames,
-   the pipe operator `|>` to chain functions together, and
-   data transformation functions like `filter()`, `arrange()`,
    `select()`, `group_by()`, and `mutate()` provided by the tidyverse
    package `dplyr` [@R-dplyr].

A good introduction can be found in the [Section on Data
Transformation](https://r4ds.hadley.nz/data-transform) [@Wickham2023].

Load the tidyverse packages.

```{r}
library(tidyverse)
```

### Tibbles

Here is a short example That analyses the vitamin C content of different
fruits. that will get you familiar with the basic syntax. Data tables in
R are called data.frames. Tidyverse introduces its own version called
tibbles. We create a tibble with the price in dollars per pound and the
vitamin C content in milligrams (mg) per pound for five different fruits.


```{r }
fruit <- tibble(
  name = c("apple", "banana", "mango", "orange", "lime"), 
  price = c(2.5, 2.0, 4.0, 3.5, 2.5), 
  vitamin_c = c(20, 45, 130, 250, 132),
  type = c("pome", "tropical", "tropical", "citrus", "citrus"))
fruit
```

### Transformations

We can modify the table by adding a column with the vitamin C (in mg) 
that a dollar buys using `mutate()`. Then we filter only rows with
fruit that provides more than 20 mg, and finally we arrange the data
rows by the vitamin C per dollar from largest to smallest.

```{r }
affordable_vitamin_c_sources <- fruit |>
  mutate(vitamin_c_per_dollar = vitamin_c / price) |> 
  filter(vitamin_c_per_dollar > 20) |>
  arrange(desc(vitamin_c_per_dollar))

affordable_vitamin_c_sources 
```

The pipes operator `|>` lets you pass the value to the left (often the
result of a function) as the first argument to the function to the
right. This makes composing a sequence of function calls to transform
data much easier to write and read. 
You will often see `%>%` as the pipe operator, especially in examples using 
tidyverse. Both operators work similarly where `|>` is a native R operator
while `%>%` is provided in the extension package `magrittr`.

The code above starts with the fruit
data and pipes it through three transformation functions. The final
result is assigned with `<-` to a variable.

We can create summary statistics of price using `summarize()`.

```{r }
affordable_vitamin_c_sources |> 
  summarize(min = min(price), 
            mean = mean(price), 
            max = max(price))
```

We can also calculate statistics for groups by first grouping the data
wirh `group_by()`. Here 
we produce statistics for fruit types.

```{r }
affordable_vitamin_c_sources |> 
  group_by(type) |>
  summarize(min = min(price), 
            mean = mean(price), 
            max = max(price))
```


Often, we want to apply the same function to multiple columns. This can
be achieved using `across()`.

```{r }
affordable_vitamin_c_sources |> 
  summarize(across(c(price, vitamin_c), mean))
```

`dplyr` syntax and evaluation is slightly different from standard R
which may lead to confusion. One example is that column names can be
references without quotation marks. A very useful reference resource
when working with `dplyr` is the [RStudio Data Transformation Cheatsheet](https://rstudio.github.io/cheatsheets/html/data-transformation.html)
which covers on two pages almost everything you will need and also
contains contains simple example code that you can take and modify for
your use case.

### ggplot2

For visualization, we will use mainly `ggplot2`. The *gg* in `ggplot2`
stands for **Grammar of Graphics** introduced by @Wilkinson2005. The
main idea is that every graph is built from the same basic components:

-   the data,
-   a coordinate system, and
-   visual marks representing the data (geoms).

In `ggplot2`, the components are combined using the `+` operator.

```{r eval=FALSE}
ggplot(data, mapping = aes(x = ..., y = ..., color = ...)) +
  geom_point()
```

Since we typically use a Cartesian coordinate system, `ggplot` uses that
by default. Each `geom_` function uses a `stat_` function to calculate
what is visualizes. For example, `geom_bar` uses `stat_count` to create
a bar chart by counting how often each value appears in the data (see
`? geom_bar`). `geom_point` just uses the stat `"identity"` to display
the points using the coordinates as they are. A great introduction can
be found in the [Section on Data
Visualization](https://r4ds.hadley.nz/data-visualize) [@Wickham2023],
and very useful is [RStudio's Data Visualization Cheatsheet](https://rstudio.github.io/cheatsheets/html/data-visualization.html).

We can visualize our fruit data as a scatter plot.

```{r }
ggplot(fruit, aes(x = price, y = vitamin_c)) + 
  geom_point()
```

It is easy to add more geoms. For example, we can add a regression line
using `geom_smooth` with the method `"lm"` (linear model). We suppress
the confidence interval since we only have 3 data points.

```{r }
ggplot(fruit, aes(x = price, y = vitamin_c)) + 
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)
```

Alternatively, we can visualize each fruit's vitamin C content per
dollar using a bar chart.

```{r }
ggplot(fruit, aes(x = name, y = vitamin_c)) + 
  geom_bar(stat = "identity")
```

Note that `geom_bar` by default uses the `stat_count` function to
aggregate data by counting, but we just want to visualize the value in
the tibble, so we specify the identity function instead.

