---
editor_options: 
  markdown: 
    wrap: 72
---

# Introduction

```{r setup_01, include=FALSE}
pkgs <- sort(c('tidyverse', 'ggplot2'))

lapply(pkgs, function(pkg) {
  if (system.file(package = pkg) == '') install.packages(pkg)
})

all_pkgs <- union(all_pkgs, pkgs)
```

**Packages used for this chapter:** `r format_pkgs(pkgs)`

## Used Software

This companion book assumes that you have
[R](https://cran.r-project.org/) and [RStudio
Desktop](https://www.rstudio.com/products/rstudio/) installed and that
you are familiar with the basics of R, how to run R code and install
packages.

If you are new to R, then working through the official R manual [An
Introduction to
R](https://cran.r-project.org/doc/manuals/r-release/R-intro.html)
[@RCore2021] will get you started. There are many introduction videos
for RStudio and a basic video that shows you how to run code and how to
install packages will suffice.

The code in this book uses `tidyverse` to manipulate data and `ggplot2`
for visualization. A great introduction to these very useful tools can
be found in the freely available web book [R for Data
Science](https://r4ds.had.co.nz) by @Wickham2017.

## Tidyverse

```{r}
library(tidyverse)
```

`tidyverse` [@R-tidyverse] is a collection of many very useful packages
that work well together by sharing design principles and data
structures. `tidyverse` also includes `ggplot2` [@R-ggplot2] for
visualization.

In this book, we will use

-   tidyverse tibbles to replace R's built-in data.frames,
-   the pipe operator `%>%` to chain functions together, and
-   data transformation functions like `filter()`, `arrange()`,
    `select()`, `group_by()`, and `mutate()` provided by the tidyverse
    package `dplyr`.

A good introduction can be found in the [Section on Data
Wrangling](https://r4ds.had.co.nz/wrangle-intro.html) [@Wickham2017],
and a very useful reference resource is the [RStudio Data Transformation
Cheat
Sheet](https://github.com/rstudio/cheatsheets/raw/master/data-transformation.pdf).

Here is a short example. We create a tibble with the price in dollars
per pound and the vitamin C content in milligrams (mg) per pound for
three fruit.

```{r }
fruit <- tibble(
  name = c("apple", "banana", "orange"), 
  price = c(2.5, 2.0, 3.5), 
  vitamin_c = c(20, 45, 250))
fruit
```

Now we add a column with the vitamin C (in mg) that a dollar buys you,
filter only fruit that provides more than 20 mg and then order (arrange)
the data by the vitamin C per dollar from largest to smallest.

```{r }
affordable_vitamin_c_sources <- fruit %>% 
  mutate(vitamin_c_per_dollar = vitamin_c / price) %>% 
  filter(vitamin_c_per_dollar > 20) %>% arrange(desc(vitamin_c_per_dollar))
affordable_vitamin_c_sources 
```

The pipes operator `%>%` lets you compose a sequence of function calls
in a more readable way by passing the value to the left on as the first
argument to the function to the right.

## ggplot2

For visualization, we will use mainly `ggplot2`. The *gg* in `ggplot2`
stands for **The Grammar of Graphics** introduced in @Wilkinson2005. The
main idea is that every graph is built from the same basic components:

-   the data,
-   a coordinate system, and
-   visual marks representing the data (geoms).

In `ggplot2`, the components are combined using the `+` operator.

> `ggplot(data, mapping = aes(x = ..., y = ..., color = ...)) +`
> `geom_point()`

Since we typically use a Cartesian coordinate system, `ggplot` uses that
by default. Each `geom_` function uses a `stat_` function to calculate
what is visualizes. For example, `geom_bar` uses `stat_count` to create
a bar chart by counting how often each value appears in the data (see
`? geom_bar`). `geom_point` just uses the stat `"identity"` to display
the points using the coordinates as they are. A great introduction can
be found in the [Chapter on Data
Visualization](https://r4ds.had.co.nz/data-visualisation.html)
[@Wickham2017], and very useful is [RStudio's Data Visualization Cheat
Sheet](https://github.com/rstudio/cheatsheets/raw/master/data-visualization-2.1.pdf).

We can visualize our fruit data as a scatter plot.

```{r }
ggplot(fruit, aes(x = price, y = vitamin_c)) + 
  geom_point()
```

Alternatively, we can visualize the vitamin C content of each fruit
using bars.

```{r }
ggplot(fruit, aes(x = name, y = vitamin_c)) + 
  geom_bar(stat = "identity")
```

Note that `geom_bar` by default tries to aggregate data by counting, but
we just want to visualize the value already available in the tibble, so
we specify the identity statistic.
