# Introduction


```{r setup_01, include=FALSE}
pkgs <- sort(c('tidyverse', 'ggplot2'))

lapply(pkgs, function(pkg) {
  if (system.file(package = pkg) == '') install.packages(pkg)
})

all_pkgs <- union(all_pkgs, pkgs)
```

**Packages used for this chapter:** `r format_pkgs(pkgs)`

This companion book assumes that you have [R](https://cran.r-project.org/) and [RStudio Desktop](https://www.rstudio.com/products/rstudio/) installed and that you are familiar with the basics of R, how to run R code and install packages.

If you are new to R, then working through [An Introduction to R](https://cran.r-project.org/doc/manuals/r-release/R-intro.html) will get you started.
There are many introduction videos for RStudio and a basic video that shows you how to run code and how to install packages will suffice.

The code in this book uses `tidyverse` to manipulate data and `ggplot2` for visualization.
A great introduction to these very useful tools can be found in the book [R for Data Science](https://r4ds.had.co.nz) [@Wickham2017]. 

```{r}
library(tidyverse)
```

`tidyverse` is a collection of many very useful packages that work well together by 
sharing design principles and data structures. `tidyverse` also includes `ggplot2`.

In this book, we will 
use tidyverse tibbles to replace R's built-in data.frames, 
the pipe operator `%>%` to chain
functions and data transformation functions like `filter()`, `arrange()`, `select()`, and
`mutate()` provided by the tidyverse package `dplyr`. 
A good introduction can be found in the
[Section on Data Wrangling](https://r4ds.had.co.nz/wrangle-intro.html) of [@Wickham2017],
and
a very useful resource is the [RStudio Data Transformation Cheat Sheet](https://github.com/rstudio/cheatsheets/raw/master/data-transformation.pdf). 

For visualization, we will use mainly `ggplot2`.
The _gg_ in `ggplot2` stands for __The Grammar of Graphics__ introduced in 
@Wilkinson2005.
The main idea is that every graph is built from the same basic components:

- the data,
- a coordinate system, and
- visual marks representing the data (geoms).

In `ggplot2`, the components are combined using the `+` operator.

> `ggplot(data, mapping = aes(x = ..., y = ..., color = ...)) +`
> `geom_point() +`
> `coord_cartesian()`

Each `geom_` function uses a `stat_` function to calculate what is visualizes. For example,
`geom_bar` uses `stat_count` to create a bar chart by counting how often each value appears in the data (see `? geom_bar`). `geom_point` just uses the stat `"identity"` to display the points using the coordinates as they are.
A great introduction can be found in the [Chapter on Data Visualization](https://r4ds.had.co.nz/data-visualisation.html) [@Wickham2017],
and very useful is
[RStudio's Data Visualization Cheat Sheet](https://github.com/rstudio/cheatsheets/raw/master/data-visualization-2.1.pdf).
