---
editor_options: 
  markdown: 
    wrap: 72
---

# Introduction

[Data mining](https://en.wikipedia.org/wiki/Data_mining) has the goal of
finding patterns in large data sets. The popular data mining textbook
[*Introduction to Data
Mining*](https://www-users.cs.umn.edu/~kumar001/dmbook/) [@Tan2018]
covers many important aspects of data mining. This companion contains
annotated R code examples to complement the textbook. To make following
along easier, we follow the chapters in data mining textbook which is
organized by the main data mining tasks:

-   

    2.  [Data] covers types of data and also includes data preparation
        and exploratory data analysis in the chapter.

-   

    3.  [Classification: Basic Concepts and Techniques] introduces
        decision trees, model training and evaluation.

-   

    4.  [Classification: Alternative techniques] introduces and compares
        methods including rule-based classifiers, nearest neighbor
        classifiers, naive Bayes classifier, logistic regression and
        artificial neural networks.

-   

    5.  [Association Analysis: Basic Concepts and Algorithms] covers
        frequent itemset and association rule generation and analysis
        including visualization. Some code covers also topics discussed
        in the textbook's Chapter 6.

-   

    7.  [Cluster Analysis] discusses clustering approaches including
        k-means, hierarchical clustering, DBSCAN and how to evaluate
        clustering results.

For completeness, we have added chapters on [Regression\*] and on
[Logistic Regression\*]. Sections in this book followed by an asterisk
contain code examples for methods that are not described in the data
mining textbook.

## Used Software

This companion book assumes that you have
[R](https://cran.r-project.org/) and [RStudio
Desktop](https://www.rstudio.com/products/rstudio/) installed and are
familiar with the basics of R, how to run R code, and install packages.
If you are new to R, working through the official R manual [An
Introduction to
R](https://cran.r-project.org/doc/manuals/r-release/R-intro.html)
[@RCore2021] will get you started. There are many introduction videos
for RStudio available, and a basic video that shows how to run code and
how to install packages will suffice.

Each book chapter will use a set of packages that must be installed. The
installation code can be found at the beginning of each chapter. Here is
the code to install the packages used in this chapter:

```{r setup_01, message=FALSE}
pkgs <- c('tidyverse', 'ggplot2')

pkgs_install <- pkgs[!(pkgs %in% installed.packages()[,"Package"])]
if(length(pkgs_install)) install.packages(pkgs_install)
```

```{r setup_01-2, include=FALSE}
source("_common.R")
source("format_pkgs.R")
all_pkgs <- union(all_pkgs, pkgs)
```

The packages used for this chapter are: `r format_pkgs(pkgs)`

The code examples in this book uses the R packages `tidyverse` to manipulate 
data and `ggplot2` for visualization. These packages make working with
data in R very convenient. 
A great introduction to these useful tools can be
found in the freely available web book [R for Data
Science](https://r4ds.hadley.nz/) by @Wickham2023.
We will give a few examples later in this chapter.

Everything we do in R is built on top of the code R programming language
and the packages that are automatically installed with R. This is typically 
referred to as Base-R. 

## Base-R

Base-R is covered in detail in [An Introduction to R](https://cran.r-project.org/doc/manuals/r-release/R-intro.html).
The most important difference between R and many other programming languages
like Python is that R is functional (with some extensions) and vectorized.

### Vectors

Scalars do not exist in R. The basic data structure is a vector of real numbers.
Special values like infinity and missing (`NA`) can be used.

```{r }
x <- c(10.4, 5.6, Inf, NA, 21.7) # c combines values
x
y <- seq(from = 0, to = 40, length.out = 5) # create a sequence
y
```

### Operations

Operations are vectorized so loops are typically not necessary.

```{r }
x + 1
```

Comparisons are performed element wise and return a `logical` vector (R's name for Boolean). 

```{r }
x > y
```

### Functions

Functions work like in many other languages. However, they also operate 
vectorized and arguments can be specified positional or by named. 
An increment function can be defined as:

```{r }
inc <- function(x, by = 1) { 
    x + by 
  }
```

Calling the increment function on vector `x`.

```{r }
inc(x, by = 2)
```

R has many built-in functions like `min()`, `max()`, `mean()` and `sum()`.

### Subsetting Vectors

We can select vector elements using the `[` operator like in other 
programming languages but the index starts at 1.

```{r}
x[1:3] # elements 1 through 3
x[-1] # all but the first
x[!is.na(x)] # all but the missing values

```

We can also assign values to a selection. For example, to get rid of 
infinite values.
```{r}
x[!is.finite(x)] <- NA
x
```

### Strings

R uses `character` vectors. R accepts double or single quotation marks.

```{r }
string <- c("Hello", "Goodbye")
string
```

Strings can be combined using `paste()`.

```{r}
paste(string, "World!")
```

Note that paste is vectorized and the string `"World!"`
is used twice to match the length of `string`. 
This behavior is called in R _recycling._

### Plotting 

Basic plotting in Base-R is done by calling `plot()`.

```{r }
plot(x, y)
```

Other plot functions are `pairs()`, `hist()`, `barplot()`. In this book,
we will focus on plots created with the ggplot2 package.

### Objects

Other often used data structures include `list`, `data.frame`, `matrix`, and
`factor`. In R, everything is an object. Objects are printed 
by either just using the object's name or put it explicitly 
into the `print()` function.

```{r }
x
```

For many objects, a summary can be created.

```{r }
summary(x)
```

Objects have a class.

```{r }
class(x)
```

Some functions are generic, meaning that they do something else depending 
on the class of the first argument. For example, `plot()` has many methods 
implemented to visualize data of different classes. The specific function is specified with a period after the method name. There also exists a default method. For plot 
this method is `plot.default()`. Different methods may have their own manual page,
so specifying the class with the dot can help finding the documentation.

It is often useful to know what information it stored in an object.
`str()` returns a humanly readable string representation of the object. 

```{r}
str(x)
```

There is much to learn about R. It is highly recommended to go through
[An Introduction to R](https://cran.r-project.org/doc/manuals/r-release/R-intro.html).

## Tidyverse


`tidyverse` [@R-tidyverse] is a collection of many useful packages that
work well together by sharing design principles and data structures.
`tidyverse` also includes `ggplot2` [@R-ggplot2] for visualization.

In this book, we will use

-   often tidyverse tibbles to replace R's built-in data.frames,
-   the pipe operator `|>` to chain functions together, and
-   data transformation functions like `filter()`, `arrange()`,
    `select()`, `group_by()`, and `mutate()` provided by the tidyverse
    package `dplyr`.

A good introduction can be found in the [Section on Data
Transformation](https://r4ds.hadley.nz/data-transform) [@Wickham2023].

Load the tidyverse packages.

```{r}
library(tidyverse)
```

### Tibbles

Here is a short example That analyses the vitamin C content of different
fruits. that will get you familiar with the basic syntax. Data tables in
R are called data.frames. Tidyverse introduces its own version called
tibbles. We create a tibble with the price in dollars per pound and the
vitamin C content in milligrams (mg) per pound for four different types
of fruit.

```{r }
fruit <- tibble(
  name = c("apple", "banana", "mango", "orange"), 
  price = c(2.5, 2.0, 4.0, 3.5), 
  vitamin_c = c(20, 45, 130, 250))
fruit
```

### Transformations

Now, we can add a column with the vitamin C (in mg) that a dollar buys
you using `mutate()` to change the table. Then we filter only rows with
fruit that provides more than 20 mg, and finally we arrange the data
rows by the vitamin C per dollar from largest to smallest.

```{r }
affordable_vitamin_c_sources <- fruit |>
  mutate(vitamin_c_per_dollar = vitamin_c / price) |> 
  filter(vitamin_c_per_dollar > 20) |>
  arrange(desc(vitamin_c_per_dollar))

affordable_vitamin_c_sources 
```

The pipes operator `|>` lets you pass the value to the left (often the
result of a function) as the first argument to the function to the
right. This makes composing a sequence of function calls to transform
data much easier to write and read. The code above starts with the fruit
data and pipes it through three transformation functions. The final
result is assigned with `<-` to a variable.

We can create summary statistics of price using `summarize()`.

```{r }
affordable_vitamin_c_sources |> 
  summarize(min = min(price), 
            mean = mean(price), 
            max = max(price))
```

Often, we want to apply the same function to multiple columns. This can
be achieved using `across()`.

```{r }
affordable_vitamin_c_sources |> 
  summarize(across(c(price, vitamin_c), mean))
```

`dplyr` syntax and evaluation is slightly different from standard R
which may lead to confusion. One example is that column names can be
references without quotation marks. A very useful reference resource
when working with `dplyr` is the [RStudio Data Transformation Cheat
Sheet](https://github.com/rstudio/cheatsheets/raw/master/data-transformation.pdf)
which covers on two pages almost everything you will need and also
contains contains simple example code that you can take and modify for
your use case.

## ggplot2

For visualization, we will use mainly `ggplot2`. The *gg* in `ggplot2`
stands for **Grammar of Graphics** introduced by @Wilkinson2005. The
main idea is that every graph is built from the same basic components:

-   the data,
-   a coordinate system, and
-   visual marks representing the data (geoms).

In `ggplot2`, the components are combined using the `+` operator.

```{r eval=FALSE}
ggplot(data, mapping = aes(x = ..., y = ..., color = ...)) +
  geom_point()
```

Since we typically use a Cartesian coordinate system, `ggplot` uses that
by default. Each `geom_` function uses a `stat_` function to calculate
what is visualizes. For example, `geom_bar` uses `stat_count` to create
a bar chart by counting how often each value appears in the data (see
`? geom_bar`). `geom_point` just uses the stat `"identity"` to display
the points using the coordinates as they are. A great introduction can
be found in the [Section on Data
Visualization](https://r4ds.hadley.nz/data-visualize) [@Wickham2023],
and very useful is [RStudio's Data Visualization Cheat
Sheet](https://rstudio.github.io/cheatsheets/html/data-visualization.html).

We can visualize our fruit data as a scatter plot.

```{r }
ggplot(fruit, aes(x = price, y = vitamin_c)) + 
  geom_point()
```

It is easy to add more geoms. For example, we can add a regression line
using `geom_smooth` with the method `"lm"` (linear model). We suppress
the confidence interval since we only have 3 data points.

```{r }
ggplot(fruit, aes(x = price, y = vitamin_c)) + 
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)
```

Alternatively, we can visualize each fruit's vitamin C content per
dollar using a bar chart.

```{r }
ggplot(fruit, aes(x = name, y = vitamin_c)) + 
  geom_bar(stat = "identity")
```

Note that `geom_bar` by default uses the `stat_count` function to
aggregate data by counting, but we just want to visualize the value
in the tibble, so we specify the identity function instead.
