---
editor_options: 
  markdown: 
    wrap: 72
---

# Introduction

## Used Software

This companion book assumes that you have
[R](https://cran.r-project.org/) and [RStudio
Desktop](https://www.rstudio.com/products/rstudio/) installed and that
you are familiar with the basics of R, how to run R code and install
packages.

If you are new to R, then working through the official R manual [An
Introduction to
R](https://cran.r-project.org/doc/manuals/r-release/R-intro.html)
[@RCore2021] will get you started. There are many introduction videos
for RStudio and a basic video that shows you how to run code and how to
install packages will suffice.

Each book chapter will use a set of packages which need to be installed. 
The installation code can be found at the beginning of each chapter. 
Here is the code to install the packages used in this chapter:

```{r setup_01, message=FALSE}
pkgs <- sort(c('tidyverse', 'ggplot2'))

pkgs_install <- pkgs[!(pkgs %in% installed.packages()[,"Package"])]
if(length(pkgs_install)) install.packages(pkgs_install)
```

The packages used for this chapter are: `r format_pkgs(pkgs)`

```{r setup_01-2, include=FALSE}
all_pkgs <- union(all_pkgs, pkgs)
```

The code in this book uses `tidyverse` to manipulate data and `ggplot2`
for visualization. A great introduction to these very useful tools can
be found in the freely available web book [R for Data
Science](https://r4ds.had.co.nz) by @Wickham2017.

## Tidyverse

```{r}
library(tidyverse)
```

`tidyverse` [@R-tidyverse] is a collection of many very useful packages
that work well together by sharing design principles and data
structures. `tidyverse` also includes `ggplot2` [@R-ggplot2] for
visualization.

In this book, we will use

-   tidyverse tibbles to replace R's built-in data.frames,
-   the pipe operator `%>%` to chain functions together, and
-   data transformation functions like `filter()`, `arrange()`,
    `select()`, `group_by()`, and `mutate()` provided by the tidyverse
    package `dplyr`.

A good introduction can be found in the [Section on Data
Wrangling](https://r4ds.had.co.nz/wrangle-intro.html) [@Wickham2017],
and a very useful reference resource is the [RStudio Data Transformation
Cheat
Sheet](https://github.com/rstudio/cheatsheets/raw/master/data-transformation.pdf).

Here is a short example. We create a tibble with the price in dollars
per pound and the vitamin C content in milligrams (mg) per pound for
three fruit.

```{r }
fruit <- tibble(
  name = c("apple", "banana", "orange"), 
  price = c(2.5, 2.0, 3.5), 
  vitamin_c = c(20, 45, 250))
fruit
```

Now we add a column with the vitamin C (in mg) that a dollar buys you,
filter only fruit that provides more than 20 mg and then order (arrange)
the data by the vitamin C per dollar from largest to smallest.

```{r }
affordable_vitamin_c_sources <- fruit %>% 
  mutate(vitamin_c_per_dollar = vitamin_c / price) %>% 
  filter(vitamin_c_per_dollar > 20) %>% 
  arrange(desc(vitamin_c_per_dollar))

affordable_vitamin_c_sources 
```

The pipes operator `%>%` lets you compose a sequence of function calls
in a more readable way by passing the value to the left on as the first
argument to the function to the right.

## ggplot2

For visualization, we will use mainly `ggplot2`. The *gg* in `ggplot2`
stands for **The Grammar of Graphics** introduced in @Wilkinson2005. The
main idea is that every graph is built from the same basic components:

-   the data,
-   a coordinate system, and
-   visual marks representing the data (geoms).

In `ggplot2`, the components are combined using the `+` operator.

> `ggplot(data, mapping = aes(x = ..., y = ..., color = ...)) +`
> `geom_point()`

Since we typically use a Cartesian coordinate system, `ggplot` uses that
by default. Each `geom_` function uses a `stat_` function to calculate
what is visualizes. For example, `geom_bar` uses `stat_count` to create
a bar chart by counting how often each value appears in the data (see
`? geom_bar`). `geom_point` just uses the stat `"identity"` to display
the points using the coordinates as they are. A great introduction can
be found in the [Chapter on Data
Visualization](https://r4ds.had.co.nz/data-visualisation.html)
[@Wickham2017], and very useful is [RStudio's Data Visualization Cheat
Sheet](https://github.com/rstudio/cheatsheets/raw/master/data-visualization-2.1.pdf).

We can visualize our fruit data as a scatter plot.

```{r }
ggplot(fruit, aes(x = price, y = vitamin_c)) + 
  geom_point()
```
It is easy to add more geoms. For example, we can add a regression line
using `geom_smooth` with the method `"lm"` (linear model). We suppress the
confidence interval since we only have 3 data points.

```{r }
ggplot(fruit, aes(x = price, y = vitamin_c)) + 
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)
```


Alternatively, we can visualize the vitamin C content of each fruit
using a bar chart.

```{r }
ggplot(fruit, aes(x = name, y = vitamin_c)) + 
  geom_bar(stat = "identity")
```
Using an additional geom, we can easily add a smoothing layer with a
regression line (`"lm"` stands for linear model).
```{r }
ggplot(fruit, aes(x = name, y = vitamin_c)) + 
  geom_bar(stat = "identity") +
  geom_smooth(method = "lm")
```


Note that `geom_bar` by default uses the `stat_count` statistics to 
aggregate data by counting, but
we just want to visualize the value already available in the tibble, so
we specify the identity statistic instead.
