---
editor_options:
  markdown:
    wrap: 72
---

# Data

Data for data mining is typically organized in tabular form, with rows containing
the objects of interest and columns representing attributes describing the objects.
We will discuss topics like data quality, sampling, feature selection,
and how to measure similarities between objects and features. 

After this chapter, you can read Appendix A [Data Exploration and Visualization] 
to learn more about data exploration and visualization in R.

### Packages Used in this Chapter {.unnumbered}

```{r setup_02}
pkgs <- c("arules", "caret", "factoextra", "GGally", 
          "palmerpenguins", "plotly", 
          "proxy", "Rtsne", "tidyverse")

pkgs_install <- pkgs[!(pkgs %in% installed.packages()[,"Package"])]
if(length(pkgs_install)) install.packages(pkgs_install)
```

The packages used for this chapter are: 

`r format_pkgs(pkgs)`

```{r setup_02-2, include=FALSE}
source("_common.R")
source("format_pkgs.R")
all_pkgs <- union(all_pkgs, pkgs)
```


## Types of Data

### Attributes and Measurement

The values of features can be measured on [several scales](https://en.wikipedia.org/wiki/Level_of_measurement) 
ranging from 
simple labels all the way to numbers. The scales come in four levels.

| Scale Name | Description | Operations | Statistics | R |
| ---------- | ------------| --------- | -------| ------- |
| Nominal | just a label (e.g., red, green) | $==, !=$ | counts | `factor` |
| Ordinal | label with order (e.g., small, med., large) | $<, >$ | median | `ordered factor` | 
| Interval | difference between two values is meaningful (regular number) | $+, -$ | mean, sd | `numeric` |
| Ratio | has a natural zero (e.g., count, distance) | $/, *$ | percent | `numeric` |

The scales build on each other meaning that an ordinal variable also has the characteristics of 
a nominal variable with the added order information.
We often do not differentiate between interval and ratio scale because we rarely 
not need to calculate percentages or other statistics that require a meaningful zero value.

Nominal data is created using `factor()`. If the factor levels
are not specified, then they are created in alphabetical order.

```{r}
factor(c("red", "green", "green", "blue"))
```

Ordinal data is created using `ordered()`. The levels specify the order. 

```{r}
ordered(c("S", "L", "M", "S"), 
       levels = c("S", "M", "L"))
```


Ratio/interval data is created as a simple vector.

```{r}
c(1, 2, 3, 4, 3, 3)
```



### The Iris Dataset

We will use a toy dataset that comes with R. [Fisher's iris
dataset](https://en.wikipedia.org/wiki/Iris_flower_data_set) gives the
measurements in centimeters of the variables sepal length, sepal width
petal length, and petal width representing the features for 150 flowers (the objects). 
The dataset contains 50
flowers from each of 3 species of iris. The species are Iris Setosa,
Iris Versicolor, and Iris Virginica. For more details see `? iris`.

We load the iris data set. Datasets that come with R or R packages can
be loaded with `data()`. The standard format for data in R is a
data.frame. We convert the data.frame into a tidyverse tibble.

```{r }
library(tidyverse)
data(iris)
iris <- as_tibble(iris)
iris
```

We see that the data contains 150 rows (flowers) and 5 features. tibbles
only show the first few rows and do not show all features, if they do
not fit the screen width. We can call `print` and define how many rows
to show using parameter `n` and force print to show all features by
changing the `width` to infinity.

```{r }
print(iris, n = 3, width = Inf)
```

## Data Quality

Assessing the quality of the available data is crucial before we start
using the data. Start with summary statistics for each column to
identify outliers and missing values. The easiest way is to use the base R
function `summary()`.

```{r }
summary(iris)
```


Feature names will be used in plots and analysis so having understandable 
feature names is part of data quality.
Iris's names are good but for most datasets, you may need to change 
the names using either `rownames()`
or `rename()`.

You can also summarize 
individual columns using tidyverse's dplyr functions.

```{r }
iris |> 
  summarize(mean = mean(Sepal.Length))
```

Using `across()`, multiple columns can be summarized. Un the following,
we calculate all numeric columns using the `mean` function.

```{r }
iris |> 
  summarize(across(where(is.numeric), mean))
```

To find outliers or data problems, you need to look for very small
values (often a suspicious large number of zeros) using min and for
extremely large values using max. Comparing median and mean tells us if
the distribution is symmetric.

A visual method to inspect the data is to use a scatterplot matrix (we
use here `ggpairs()` from package `GGally`). In this plot, we can
visually identify noise data points and outliers (points that are far
from the majority of other points).

```{r }
library(GGally)
ggpairs(iris, aes(color = Species), progress = FALSE)
```

This useful visualization combines many visualizations used to understand the
data and check for quality issues. Rows and columns are the features in the data.
We have also specified the aesthetic that we want to group each species using a different color.

* The visualizations in the diagonal panels show the
smoothed histograms with the distribution for each feature. 
The plot tries to pick a good number of bins for the histogram 
(see messages above). The distribution can be checked if it is close to normal,
unimodal or highly skewed. Also, we can see if the different groups are overlapping 
or separable for each feature. For example, the three distributions for `Sepal.Width`
are almost identical meaning that it is hard to distinguish between 
the different species using this feature alone. `Petal.Lenght` and `Petal.Width`
are much better.

* The lower-left triangle panels contain scatterplots for all pairs features. These 
are useful to see if there if features are correlated (the pearson correlation 
coefficient if printed in the upper-right triangle). For example,
`Petal.Length` and `Petal.Width` are highly correlated overall 
This makes sense since larger plants will have both longer and wider petals.
Inside the Setosa group this correlation it is a lot weaker.
We can also see if groups are 
well separated using projections on two variables. Almost all panels show that
Setosa forms a point cloud well separated from the other two classes while 
Versicolor and Virginica overlap. 
We can also see outliers that are far from the other data points in its group.
See if you can spot the one red dot that is far away from all others.

* The last row/column represents in this data set the class label Species. 
It is a nominal variable so the plots are different. The bottom row panels 
show (regular) histograms. The last column shows boxplots to represent
the distribution of the different features by group. Dots represent 
outliers. Finally, the bottom-right panel contains the counts for the different 
groups as a barplot. In this data set, each group has the same number of observations.

Many data mining methods require complete data, that is the data cannot
contain missing values (`NA`). To remove missing values and duplicates
(identical data points which might be a mistake in the data), we often
do this:

```{r }
clean.data <- iris |> 
  drop_na() |> 
  unique()

summary(clean.data)
```

The iris dataset has no missing values, but one non-unique case is gone 
leaving only 149 flowers. Since only 1 out of 150 flowers in the dataset was 
affected, results on the remaining data will give very similar results to the 
complete data.

Typically, you should spend a lot more time on data cleaning.
It is important to always describe how you clean the data, 
and how many objects are removed. You need to argue that conclusions based on
only the remaining data are still valid.

## Data Preprocessing

### Aggregation

Data often contains groups and we want to compare these groups. We group
the iris dataset by species and then calculate a summary statistic for
each group.

```{r }
iris |> 
  group_by(Species) |> 
  summarize(across(everything(), mean))

iris |> 
  group_by(Species) |> 
  summarize(across(everything(), median))
```

Using this information, we can compare how features differ between
groups.

### Sampling

[Sampling](https://en.wikipedia.org/wiki/Sampling_(statistics)) is often
used in data mining to reduce the dataset size before modeling or
visualization. Another application is to split data randomly into training and testing 
data for working with supervised models (e.g., classification models).

#### Random Sampling

The built-in sample function can sample from a vector. Here we sample
with replacement.

```{r }
sample(c("A", "B", "C"), size = 10, replace = TRUE)
```

We often want to sample rows from a dataset. This can be done by
sampling without replacement from a vector with row indices (using the
functions `seq()` and `nrow()`). The sample vector is then used to
subset the rows of the dataset.

```{r }
take <- sample(seq(nrow(iris)), size = 15)
take

iris[take, ]
```

`dplyr` from tidyverse lets us sample rows from tibbles directly using
`slice_sample()`. I set the random number generator seed to make the
results reproducible.

```{r }
set.seed(1000)
s <- iris |> 
  slice_sample(n = 15)

library(GGally)
ggpairs(s, aes(color = Species), progress = FALSE)
```
Instead of `n` you can also specify the proportion of rows to select using
`prob`. 

#### Stratified Sampling

[Stratified sampling](https://en.wikipedia.org/wiki/Stratified_sampling)
is a method of sampling from a population which can be partitioned into
subpopulations, while controlling the proportions of the subpopulation
in the resulting sample.

In the following, the subpopulations are the different types of species
and we want to make sure to sample the same number (5) flowers from
each. This can be achieved by first grouping the data by species and 
then sampling a number of flowers from each group.


```{r }
set.seed(1000)

s2 <- iris |> 
  group_by(Species) |>
  slice_sample(n = 5) |>
  ungroup()

library(GGally)
ggpairs(s2, aes(color = Species), progress = FALSE)
```

More sophisticated sampling procedures are implemented in the package 
`sampling`.

### Dimensionality Reduction

The number of features is often called the dimensional of the data following the 
idea that each feature (at least the numeric features) can be seen as an axis of the data. 
High-dimensional data is harder to analyze by the user (e.g., visualize). 
It also is problematic for many data mining algorithms since it requires more memory 
and computational resources. 

[Dimensionality reduction](https://en.wikipedia.org/wiki/Dimensionality_reduction)
 tries to represent high-dimensional data in a low-dimensional space so that the 
low-dimensional representation retains some meaningful properties (e.g., information about similarity or distances) of the original data. Dimensionality reduction is 
used for visualization and as a prepossessing technique before using other 
data mining methods like clustering and classification. 

Recently, data embeddings using artificial neural networks have become very popular.
These approaches can not only reduce the dimensionality of the data, but 
learn a better representation of various kinds of data (e.g., text). As such
these approaches can be seen as automatically engineering features from the 
high-dimensional original data.

#### Principal Components Analysis (PCA)

[PCA](https://en.wikipedia.org/wiki/Principal_component_analysis)
calculates principal components (a set of new orthonormal basis vectors
in the data space) from data points such that the first principal
component explains the most variability in the data, the second the next
most and so on. In data analysis, PCA is used to project
high-dimensional data points onto the first few (typically two)
principal components for visualization as a scatter plot and as
preprocessing for modeling (e.g., before k-means clustering). Points
that are closer together in the high-dimensional original space, tend
also be closer together when projected into the lower-dimensional space,

We can use an interactive 3-d plot (from package `plotly`) to look at
three of the four dimensions of the iris dataset. Note that it is hard
to visualize more than 3 dimensions.

```{r eval = knitr::is_html_output()}
plotly::plot_ly(iris, 
                x = ~Sepal.Length, 
                y = ~Petal.Length, 
                z = ~Sepal.Width, 
      color = ~Species, size = 1) |> 
  plotly::add_markers()
```

```{r eval = !knitr::is_html_output(), echo=FALSE, results='asis'}
cat('_The resulting interactive plot can be seen in the [online version of this book.](https://mhahsler.github.io/Introduction_to_Data_Mining_R_Examples/book/data.html#dimensionality-reduction)_')
```

The principal components can be calculated from a matrix using the
function `prcomp()`. We select all numeric columns (by unselecting the
species column) and convert the tibble into a matrix before the
calculation.

```{r }
pc <- iris |> 
  select(-Species) |> 
  as.matrix() |> 
  prcomp()
summary(pc)
```

How important is each principal component can also be seen using a
[scree plot](https://en.wikipedia.org/wiki/Scree_plot). The plot
function for the result of the `prcomp` function visualizes how much
variability in the data is explained by each additional principal
component.

```{r }
plot(pc, type = "line")
```

Note that the first principal component (PC1) explains most of the
variability in the iris dataset.

To find out what information is stored in the object `pc`, we can
inspect the raw object (display *str*ucture).

```{r }
str(pc)
```

The object `pc` (like most objects in R) is a list with a class
attribute. The list element `x` contains the data points projected on
the principal components. We can convert the matrix into a tibble and
add the species column from the original dataset back (since the rows
are in the same order), and then display the data projected on the first
two principal components.

```{r }
iris_projected <- as_tibble(pc$x) |> 
  add_column(Species = iris$Species)

ggplot(iris_projected, aes(x = PC1, y = PC2, color = Species)) + 
  geom_point()
```

Flowers that are displayed close together in this projection are also
close together in the original 4-dimensional space. Since the first
principal component represents most of the variability, we can also show
the data projected only on PC1.

```{r }
ggplot(iris_projected, 
  aes(x = PC1, y = 0, color = Species)) + 
  geom_point() +
  scale_y_continuous(expand=c(0,0)) +
  theme(axis.text.y = element_blank(),
      axis.title.y = element_blank()
  )
```

We see that we can perfectly separate the species Setosa using just the
first principal component. The other two species are harder to separate.

A plot of the projected data with the original axes added as arrows is
called a [biplot](https://en.wikipedia.org/wiki/Biplot). If the arrows
(original axes) align roughly with the axes of the projection, then they
are correlated (linearly dependent).

```{r message=FALSE}
library(factoextra)
fviz_pca(pc)
```

We can also display only the old and new axes.

```{r }
fviz_pca_var(pc)
```

We see Petal.Width and Petal.Length point in the same direction which
indicates that they are highly correlated. They are also roughly aligned
with PC1 (called Dim1 in the plot) which means that PC1 represents most
of the variability of these two variables. Sepal.Width is almost aligned
with the y-axis and therefore it is represented by PC2 (Dim2).
Petal.Width/Petal.Length and Sepal.Width are almost at 90 degrees,
indicating that they are close to uncorrelated. Sepal.Length is
correlated to all other variables and represented by both, PC1 and PC2
in the projection.


#### Multi-Dimensional Scaling (MDS)

[MDS](https://en.wikipedia.org/wiki/Multidimensional_scaling) is similar
to PCA. Instead of data points, it starts with pairwise distances (i.e.,
a distance matrix) and produces a space where points are placed to
represent these distances as well as possible. The axes in this space
are called components and are similar to the principal components in
PCA.

First, we calculate a distance matrix (Euclidean distances) from the 4-d
space of the iris dataset.

```{r }
d <- iris |> 
  select(-Species) |> 
  dist()
```

Metric (classic) MDS tries to construct a space where points with lower
distances are placed closer together. We project the data represented by
a distance matrix on `k = 2` dimensions.

```{r }
fit <- cmdscale(d, k = 2)
colnames(fit) <- c("comp1", "comp2")
fit <- as_tibble(fit) |> 
  add_column(Species = iris$Species)

ggplot(fit, aes(x = comp1, y = comp2, color = Species)) + 
  geom_point()
```

The resulting projection is similar (except for rotation and reflection)
to the result of the projection using PCA.

#### Non-Parametric Multidimensional Scaling

Non-parametric multidimensional scaling performs MDS while relaxing the
need of linear relationships. Methods are available in package `MASS` as
functions `isoMDS()` (implements [isoMAP](https://en.wikipedia.org/wiki/Isomap)) and `sammon()`.

#### Embeddings: Nonlinear Dimensionality Reduction Methods

[Nonlinear dimensionality reduction](https://en.wikipedia.org/wiki/Nonlinear_dimensionality_reduction)
is also called manifold learning or creating a low-dimensional embedding.
These methods have become very popular to support visualizing high-dimensional data,
mine text by converting words into numeric vectors that can be used as features in models, 
and as a method to automatically create features for data mining models by finding an efficient 
representation. Popular methods are:

* Visualization: [t-distributed stochastic neighbor embedding](https://en.wikipedia.org/wiki/T-distributed_stochastic_neighbor_embedding) 
  (`Rtsne()` in package `Rtsne`) and uniform manifold approximation and projection
  (`umap()` in package `umap`) are used 
  for projecting data to 2 dimensions for visualization. Here is an example of 
  embedding the 4-dimensional iris data using tsne. The embedding algorithm 
  requires unique data points.
  
  ```{r }
  iris_distinct <- iris |> distinct(Sepal.Length, Sepal.Width, 
                                    Petal.Length, Petal.Width, 
                                    .keep_all = TRUE)
  
  tsne <- Rtsne::Rtsne(iris_distinct |> select(-Species))
  
  emb <- data.frame(tsne$Y, Species = iris_distinct$Species)
  ggplot(emb, aes(X1, X2, color = Species)) + geom_point()
  ```

  We see that the embedding separates the class Setosa very well from the other 
  two species indicating that the flowers are very different.

* Text mining: [Word2vec](https://en.wikipedia.org/wiki/Word2vec) (`word2vec()` in 
  `word2vec`) is used in [natural language processing](https://cran.r-project.org/web/views/NaturalLanguageProcessing.html) 
  to convert words to numeric vectors that represent similarities between words. 
  These vectors can be used as features for data mining models for clustering
  and classification. Other popular methods for text embedding are 
  [GloVe](https://nlp.stanford.edu/projects/glove/) and 
  [BERT](https://en.wikipedia.org/wiki/BERT_(language_model)). 

* Representation learning: [Autoencoders](https://en.wikipedia.org/wiki/Autoencoder) 
  are artificial neural networks used to learn an efficient representation (encoding) 
  for a set of data by minimizing the reconstruction error. This new representation
  can be seen as a way to automatically creating features that describe the 
  important characteristics of the data while reducing noise. This representation
  often helps models to learn better.
  In R, autoencoders are typically created using 
  the [`keras` package](https://keras3.posit.co/). Creating an
  autoencoder requires some work. You will 
  need to decide on the network topology and have sufficient training data.

### Feature Subset Selection

Feature selection is the process of identifying the features that are
used to create a model. We will talk about feature selection when we
discuss classification models in Chapter 3 in [Feature Selection*].

### Discretization

Some data mining methods require discrete data. Discretization converts
continuous features into discrete features. As an example, we will
discretize the continuous feature Petal.Width. Before we perform
discretization, we should look at the distribution and see if it gives
us an idea how we should group the continuous values into a set of
discrete values. A histogram visualizes the distribution of a single
continuous feature.

```{r }
ggplot(iris, aes(x = Petal.Width)) + 
  geom_histogram(binwidth = .2)
```

The bins in the histogram represent a discretization using a fixed bin
width. The R function `cut()` performs equal interval width
discretization creating a vector of type `factor` where each level
represents an interval.

```{r }
iris |> 
  pull(Sepal.Width) |> 
  cut(breaks = 3)
```

Other discretization methods include equal frequency discretization or
using k-means clustering. These methods are implemented by several R
packages. We use here the implementation in package `arules` and
visualize the results as histograms with blue lines to separate
intervals assigned to each discrete value.

```{r }
library(arules)
iris |> pull(Petal.Width) |> 
  discretize(method = "interval", breaks = 3)
```

To show the differences between the methods, we use the three
discretization methods and draw blue lines in the histogram to show
how they cut the data.

```{r }
ggplot(iris, aes(Petal.Width)) + geom_histogram(binwidth = .2) +
  geom_vline(
      xintercept = iris |> 
      pull(Petal.Width) |> 
      discretize(method = "interval", breaks = 3, onlycuts = TRUE),
    color = "blue"
  ) +
  labs(title = "Discretization: interval", 
       subtitle = "Blue lines are discretization boundaries")

ggplot(iris, aes(Petal.Width)) + geom_histogram(binwidth = .2) +
  geom_vline(
    xintercept = iris |> 
    pull(Petal.Width) |> 
    discretize(method = "frequency", breaks = 3, onlycuts = TRUE),
   color = "blue"
  ) +
  labs(title = "Discretization: frequency", 
       subtitle = "Blue lines are discretization boundaries")

ggplot(iris, aes(Petal.Width)) + geom_histogram(binwidth = .2) +
  geom_vline(
    xintercept = iris |> 
    pull(Petal.Width) |> 
    discretize(method = "cluster", breaks = 3, onlycuts = TRUE),
   color = "blue"
  ) +
  labs(title = "Discretization: cluster", 
       subtitle = "Blue lines are discretization boundaries")
```

The user needs to decide on the number of intervals and the used method.

### Variable Transformation: Standardization

Standardizing (scaling, normalizing) the range of features values is
important to make them comparable. The most popular method is to convert
the values of each feature to
[z-scores](https://en.wikipedia.org/wiki/Standard_score). by subtracting
the mean (centering) and dividing by the standard deviation (scaling).
The standardized feature will have a mean of zero and are measured in
standard deviations from the mean. Positive values indicate how many
standard deviation the original feature value was above the average.
Negative standardized values indicate below-average values.

R-base provides the function `scale()` to standardize the columns in a data.frame. Tidyverse currently does not have a simple scale function, so we
make one It mutates all numeric columns using an anonymous function
that calculates the z-score. 

```{r }
scale_numeric <- function(x) 
  x |> 
  mutate(across(where(is.numeric), 
                function(y) (y - mean(y, na.rm = TRUE)) / sd(y, na.rm = TRUE)))
```

```{r }
iris.scaled <- iris |> 
  scale_numeric()
iris.scaled
summary(iris.scaled)
```

The standardized feature has a mean of zero and most "normal" values
will fall in the range $[-3,3]$ and is measured in standard deviations from the average.
Negative values mean smaller than the average and positive values mean larger than the average.

## Measures of Similarity and Dissimilarity

Proximities help with quantifying how similar two objects are.
Remember, objects are rows in our data. We call one row the feature vector for the object.

[Similariy](https://en.wikipedia.org/wiki/Similarity_(geometry)) is a concept from geometry.
It is the opposite of dissimilarity. The best-known 
way to measure dissimilarity is [Euclidean distance](https://en.wikipedia.org/wiki/Euclidean_distance), but [dissimilarity and similarity](https://en.wikipedia.org/wiki/Similarity_measure) can be measured in many different ways depending on the information we have about the objects.

R stores similarity information always as a dissimilarities/distances matrices. 
Similarities are converted to dissimilarities. 
Distances are symmetric, i.e.,
the distance from A to B is the same as the distance from B to A. R
therefore stores only a triangle (typically the lower triangle) of the
distance matrix.

### Minkowsky Distances

The [Minkowsky
distance](https://en.wikipedia.org/wiki/Minkowski_distance) is a family
of distances including Euclidean and Manhattan distance. It is defined
between two feature vectors $\mathbf{p} = (p_1, p_2, ..., p_n)$ and 
$\mathbf{y} = (q_1, q_2, ..., q_n)$ as

$$
d(\mathbf{p},\mathbf{q}) = \left(\sum_{i=1}^n |p_i - q_i|^r\right)^{\frac{1}{r}} = ||\mathbf{q}-\mathbf{q}||_r. 
$$
Where the power $r$ is a positive integer. This type of distance is also called 
a $r$-norm written as $L^r$. Special values for $r$ are:

* $r = 1$: Manhattan distance
* $r = 2$: Euclidean distance
* $r = \infty$: Maximum norm (only the largest component distance counts)

To avoid
one feature to dominate the distance calculation, scaled data is
typically used. We select the first 5 flowers for this example.

```{r }
iris_sample <- iris.scaled |> 
  select(-Species) |> 
  slice(1:5)
iris_sample
```

Different types of Minkowsky distance matrices between the first 5
flowers can be calculated using `dist()`.

```{r }
dist(iris_sample, method = "euclidean")
dist(iris_sample, method = "manhattan")
dist(iris_sample, method = "maximum")
```

We see that only the lower triangle of the distance matrices are stored
(note that rows start with row 2).

### Distances for Binary Data

Binary data can be encodes as `0` and `1` (numeric) or `TRUE` and
`FALSE` (logical).

```{r }
b <- rbind(
  c(0,0,0,1,1,1,1,0,0,1),
  c(0,0,1,1,1,0,0,1,0,0)
  )
b

b_logical <- apply(b, MARGIN = 2, as.logical)
b_logical
```

#### Hamming Distance

The [Hamming distance](https://en.wikipedia.org/wiki/Hamming_distance)
is the number of mismatches between two binary vectors. For 0-1 data
this is equivalent to the Manhattan distance and also the squared
Euclidean distance.

```{r }
dist(b, method = "manhattan")
dist(b, method = "euclidean")^2
```

#### Jaccard Index

The [Jaccard index](https://en.wikipedia.org/wiki/Jaccard_index) is a
similarity measure that focuses on matching 1s. R converts the
similarity into a dissimilarity using $d_{J} = 1 - s_{J}$.

```{r }
dist(b, method = "binary")
```

### Distances for Mixed Data

Most distance measures work only on numeric data. Often, we have a
mixture of numbers and nominal or ordinal features like this data:

```{r }
people <- tibble(
  height = c(      160,    185,    170),
  weight = c(       52,     90,     75),
  sex    = c( "female", "male", "male")
)
people
```

It is important that nominal features are stored as factors and not
character (`<chr>`).

```{r }
people <- people |> 
  mutate(across(where(is.character), factor))
people
```

#### Gower's Coefficient

The Gower's coefficient of similarity works with mixed data by
calculating the appropriate similarity for each feature and then
aggregating them into a single measure. The package `proxy` implements
Gower's coefficient converted into a distance.

```{r }
library(proxy)
d_Gower <- dist(people, method = "Gower")
d_Gower
```

Gower's coefficient calculation implicitly scales the data because it
calculates distances on each feature individually, so there is no need
to scale the data first.

#### Using Euclidean Distance with Mixed Data

Sometimes methods (e.g., k-means) can only use Euclidean distance. In
this case, nominal features can be converted into 0-1 dummy variables.
After scaling, Euclidean distance will result in a usable distance
measure.

We use package `caret` to create dummy variables.

```{r }
library(caret)
data_dummy <- dummyVars(~., people) |> 
  predict(people)
data_dummy
```

Note that feature sex has now two columns. If we want that height,
weight and sex have the same influence on the distance measure, then we
need to weight the sex columns by 1/2 after scaling.

```{r }
weight_matrix <- matrix(c(1, 1, 1/2, 1/2), 
                        ncol = 4, 
                        nrow = nrow(data_dummy), 
                        byrow = TRUE)
data_dummy_scaled <- scale(data_dummy) * weight_matrix

d_dummy <- dist(data_dummy_scaled)
d_dummy
```

The distance using dummy variables is consistent with Gower's distance.
However, note that Gower's distance is scaled between 0 and 1 while the
Euclidean distance is not.

```{r }
ggplot(tibble(d_dummy, d_Gower), aes(x = d_dummy, y = d_Gower)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)
```

### More Proximity Measures

The package `proxy` implements a wide array of proximity measures
(similarity measures are converted into distances).

```{r }
library(proxy)
pr_DB$get_entry_names()
```

Note that loading the package `proxy` overwrites the default `dist()` function
in R.
You can specify which dist function to use by specifying the package in
the call. For example `stats::dist()` calls the default function in R
(the package `stats` is part of R) while `proxy::dist()` calls the
version in the package `proxy`.


## Exercises*

The R package **palmerpenguins** contains measurements for penguin of different
species from the Palmer Archipelago, Antarctica. Install the package.
It provides a CSV file which can be read in the following way:

```{r }
library("palmerpenguins")
penguins <- read_csv(path_to_file("penguins.csv"))
head(penguins)
```

Create in RStudio a new R Markdown document.
Apply the code in the sections of this chapter to the data set to answer the
following questions. 

1. What is the scale of measurement for each column?
2. Are there missing values in the data? How much data is missing?
3. Compute and discuss basic statistics.
4. Calculate the similarity between five randomly chosen penguins. Do you need to scale the data? 
   Discuss what measures are appropriate for the data.

Make sure your markdown document contains now a well formatted report.
Use the `Knit` button to create a HTML document.



