---
editor_options: 
  markdown: 
    wrap: 72
---

# Classification: Alternative Techniques

This chapter introduces different types of classifiers. It also 
discusses the important problem of class imbalance in data and options 
to deal with it. In addition, this chapter compares visually the 
decision boundaries used by different algorithms. This will provide a 
better understanding of the model bias that different algorithms have.

## Learning Objectives

After this chapter, you should be able to:

* describe the basic behavior of several common classification algorithms;
* compare models using a baseline and a consistent evaluation procedure;
* explain how class imbalance affects predictions and metrics; and
* compare decision boundaries and adjust a classifier's decision threshold.

## Prerequisites

Read [Classification: Basic Concepts]. Familiarity with model training,
cross-validation, confusion matrices, and probability is assumed.

### Packages Used in this Chapter {.unnumbered}

```{r setup_04}
pkgs <- c("basemodels", "C50", "caret", "e1071", "klaR", 
          "lattice", "MASS", "mlbench", "nnet", "palmerpenguins", 
          "randomForest", "rpart", "RWeka", "scales", "tidyverse", 
          "xgboost")

pak::pkg_install(pkgs)
```

```{r setup_04-2, include=FALSE}
source("_common.R")
source("format_pkgs.R")
all_pkgs <- union(all_pkgs, pkgs)
set.seed(1234)
```

The packages used for this chapter are: 

`r format_pkgs(pkgs)`


## Types of Classifiers

Many different [classification algorithms](https://en.wikipedia.org/wiki/Supervised_learning) 
have been proposed in the literature. 
In this chapter, we will apply some of the more popular methods.

### Set up the Training and Test Data

We will use the Zoo dataset which is included in the R package `mlbench`
(you may have to install it). The Zoo dataset containing 17 (mostly
logical) variables on different 101 animals as a data frame with 17
columns (hair, feathers, eggs, milk, airborne, aquatic, predator,
toothed, backbone, breathes, venomous, fins, legs, tail, domestic,
catsize, type). We convert the data frame into a tidyverse tibble
(optional).


```{r message=FALSE, warning=FALSE}
library(tidyverse)
data(Zoo, package = "mlbench")

Zoo <- as_tibble(Zoo)
Zoo
```

The examples use [**caret**](https://topepo.github.io/caret/) for data splitting,
model fitting, and tuning, while specialist packages provide the algorithms.
This book uses the `caret` interface; it does not implement the same examples
with `tidymodels`.

```{r message=FALSE, warning=FALSE}
library(caret)
```

Multi-core support can be used for cross-validation. **Note:** It is
commented out here because it does not work with rJava used by the RWeka-based
classifiers
below.

```{r }
##library(doMC, quietly = TRUE)
##registerDoMC(cores = 4)
##getDoParWorkers()
```

Test data is not used in the model building process and needs to be set
aside purely for testing the model after it is completely built. Here I
use 80% for training.

```{r }
inTrain <- createDataPartition(y = Zoo$type, p = .8)[[1]]
Zoo_train <- Zoo |> slice(inTrain)
Zoo_test <- Zoo |> slice(-inTrain)
```

For hyperparameter tuning, we will use 10-fold cross-validation.

**Note:** Be careful if you have many `NA` values in your data.
`train()` and cross-validation many fail in some cases. If that is the
case then you can remove features (columns) which have many `NA`s, omit
`NA`s using `na.omit()` or use imputation to replace them with
reasonable values (e.g., by the feature mean or via kNN). Highly
imbalanced datasets are also problematic since there is a chance that a
fold does not contain examples of each class leading to a hard to
understand error message.

## Rule-based classifier: PART

```{r }
rulesFit <- Zoo_train |> train(type ~ .,
  method = "PART",
  data = _,
  tuneLength = 5,
  trControl = trainControl(method = "cv"))
rulesFit
```
The model selection results are shown in the table. This is the selected model:

```{r }
rulesFit$finalModel
```

PART returns a decision list, i.e., an ordered rule set. For example, 
the first rule shows that an animal with no feathers but milk is a mammal.
I ordered rule sets, the decision of the first matching rule is used.

## Nearest Neighbor Classifier

K-Nearest neighbor classifiers classify a new data point by looking at the 
majority class labels of its k nearest neighbors in the training data set.
The used kNN implementation uses Euclidean distance to determine what data points 
are near by, so data needs be standardized
(scaled) first. Here legs are measured between 0 and 6 while all other
variables are between 0 and 1. Scaling to z-scores can be directly performed as
preprocessing in `train` using the parameter `preProcess = "scale"`.

The $k$ value is typically choose as an odd number so we get a clear majority.

```{r }
knnFit <- Zoo_train |> train(type ~ .,
  method = "knn",
  data = _,
  preProcess = "scale",
  tuneGrid = data.frame(k = c(1, 3, 5, 7, 9)),
	trControl = trainControl(method = "cv"))
knnFit
```

```{r }
knnFit$finalModel
```

kNN classifiers are lazy models meaning that instead of learning, they just
keep the complete dataset. This is why the final model just gives us a summary statistic for the class labels in the training data.


## Naive Bayes Classifier

Caret's train formula interface translates logicals and factors into dummy 
variables which the classifier interprets as numbers so it would used a [Gaussian naive Bayes](https://en.wikipedia.org/wiki/Naive_Bayes_classifier#Gaussian_naive_Bayes) estimation. To avoid this, I directly specify `x` and `y`.

```{r }
NBFit <- train(x = as.data.frame(Zoo_train[, -ncol(Zoo_train)]), 
               y = pull(Zoo_train, "type"),
               method = "nb",
               tuneGrid = data.frame(fL = c(.2, .5, 1, 5), 
                                     usekernel = TRUE, adjust = 1),
               trControl = trainControl(method = "cv"))
NBFit
```

The final model contains the prior probabilities for each class.
```{r }
NBFit$finalModel$apriori
```

And the conditional probabilities as a table for each feature. For brevity, 
we only show the tables for the first three features. For example, 
the condition probability 
$P(\text{hair} = \text{TRUE} | \text{class} = \text{mammal})$ is 
`r NBFit$finalModel$tables$hair["mammal", "TRUE"]`.

```{r }
NBFit$finalModel$tables[1:3]
```

## Bayesian Network

Bayesian networks are not covered here. R has very good 
support for modeling with Bayesian Networks. An example is
the package [bnlearn](https://www.bnlearn.com/).

## Logistic Regression {#logreg}

Logistic regression is a very powerful classification method 
and should always be tried as 
one of the first models. 
A detailed discussion with more code is available in section 
[Logistic Regression](#appendix_logistic_regression) in the Appendix.

Regular logistic regression predicts only one outcome coded as a 
binary variable. Since we have data with several
classes, we use [multinomial logistic regression](https://en.wikipedia.org/wiki/Multinomial_logistic_regression), 
also called a log-linear model
which is an extension of logistic regresses for multi-class problems.
Caret uses `nnet::multinom()` which implements penalized multinomial regression.

```{r }
logRegFit <- Zoo_train |> train(type ~ .,
  method = "multinom",
  data = _,
  trace = FALSE, # suppress some output
	trControl = trainControl(method = "cv"))
logRegFit
```

```{r }
logRegFit$finalModel
```
The coefficients are log odds ratios
measured against the default class (here mammal).
A negative log odds ratio means that the odds go down with an increase in
the value of the predictor.  A predictor with a
positive log-odds ratio increases the odds. For example,
in the model above, `hair=TRUE` has a negative coefficient for 
bird but `feathers=TRUE` has a large positive coefficient.

## Artificial Neural Network (ANN)

Standard networks have an input layer, an output layer and in between
a single hidden layer.

```{r }
nnetFit <- Zoo_train |> train(type ~ .,
  method = "nnet",
  data = _,
	tuneLength = 5,
	trControl = trainControl(method = "cv"),
  trace = FALSE # no progress output
  )
nnetFit
```

The input layer has a size of 16, one for each input feature and 
the output layer has a size of 7 representing the 7 classes.
Model selection chose a network architecture with a hidden
layer with `r nnetFit$bestTune$size` units
resulting in `r length(nnetFit$finalModel$wts)` learned weights.
Since the model is considered a black-box model only the network architecture
and the used variables are shown as a summary.

```{r }
nnetFit$finalModel
```

For deep Learning, R offers packages for using [tensorflow](https://tensorflow.rstudio.com/) 
and [Keras](https://keras3.posit.co/).


## Support Vector Machines

```{r }
svmFit <- Zoo_train |> train(type ~.,
  method = "svmLinear",
  data = _,
	tuneLength = 5,
	trControl = trainControl(method = "cv"))
svmFit
```

We use a linear support vector machine.
Support vector machines can use kernels to create non-linear decision boundaries.
`method` above can be changed to `"svmPoly"` or `"svmRadial"` to 
use kernels. The choice of kernel is typically make by experimentation.

The support vectors determining the decision boundary are stored in the model.
```{r }
svmFit$finalModel
```

## Ensemble Methods

Many ensemble methods are available in R. We only cover here code for two 
popular methods.

### Random Forest

```{r }
randomForestFit <- Zoo_train |> train(type ~ .,
  method = "rf",
  data = _,
	tuneLength = 5,
	trControl = trainControl(method = "cv"))
randomForestFit
```

The default number of trees is 500 and 
`mtry` determines the number of variables randomly sampled as candidates 
at each split. This number is a tradeoff where a larger number allows each tree 
to pick better splits, but a smaller
number increases the independence between trees.

```{r }
randomForestFit$finalModel
```

The model is a set of 500 trees and the prediction is made by applying all trees and 
then using the majority vote.

Since random forests use bagging (bootstrap sampling to train trees),
the remaining data can be used like a test set. The resulting error 
is called out-of-bag (OOB) error and gives an estimate for the 
generalization error. The model above also shows the confusion matrix based on
the OOB error.


### Gradient Boosted Decision Trees (xgboost)

The idea of gradient boosting is to learn a base model and then to learn
successive models to predict and correct the error of all previous models.
Typically, tree models are used.

```{r }
xgboostFit <- Zoo_train |> train(type ~ .,
  method = "xgbTree",
  data = _,
  tuneLength = 5,
  trControl = trainControl(method = "cv"),
  tuneGrid = expand.grid(
    nrounds = 20,
    max_depth = 3,
    colsample_bytree = .6,
    eta = 0.1,
    gamma=0,
    min_child_weight = 1,
    subsample = .5
  ))
xgboostFit
```

The final model is a complicated set of trees, so only some summary information
is shown.

```{r }
xgboostFit$finalModel
```


## Model Comparison

We first create a simple baseline model that always predicts the majority
class mammal.

```{r }
baselineFit <- Zoo_train |> train(type ~ .,
  method = basemodels::dummyClassifier,
  data = _,
  strategy = "constant",
  constant = "mammal",
  trControl = trainControl(method = "cv" 
                           ))
baselineFit
```

The kappa of 0 clearly indicates that the baseline model has no power.


We collect the performance metrics from the models trained on the same
data.

```{r }
resamps <- resamples(list(
  baseline = baselineFit,
  PART = rulesFit,
	kNearestNeighbors = knnFit,
  NBayes = NBFit,
  logReg = logRegFit,
  ANN = nnetFit,
  SVM = svmFit,
  RandomForest = randomForestFit,
  XGBoost = xgboostFit
    ))
resamps
```

The summary statistics show performance across resamples. All methods perform
well on this relatively simple dataset; the baseline performs poorly, as
expected.

```{r }
summary(resamps)

library(lattice)
bwplot(resamps, layout = c(3, 1))
```

Compare paired resampling results for each metric. The output shows estimated
differences and adjusted p-values. Because resampling folds are not independent,
use these comparisons as descriptive evidence rather than definitive tests of
which model is best. Bonferroni correction is applied for multiple comparisons.

```{r }
difs <- diff(resamps)
summary(difs)
```

The models have similar resampling performance except for the baseline.
We select the random forest for illustration and evaluate it once on the
held-out test set. The test set remains the final performance assessment.

```{r }
pr <- predict(randomForestFit, Zoo_test)
pr
```

Calculate the confusion matrix for the held-out test data.

```{r }
confusionMatrix(pr, reference = Zoo_test$type)
```

## Class Imbalance

Class imbalance occurs when one class has substantially more observations than
another. It can reduce performance on the minority class, especially when
identifying that class is important (for example, detecting a rare disease).
In a binary problem, the class of interest is often called the positive class.
The examples below compare resampling, threshold adjustment, and
cost-sensitive fitting.

For the examples, we will use the Zoo dataset.

```{r }
library(rpart)
library(rpart.plot)
data(Zoo, package = "mlbench")
```

Check the class distribution before fitting a classifier.

```{r }
ggplot(Zoo, aes(y = type)) + geom_bar()
```

The Zoo data contain many mammals and few amphibians and reptiles. A
classifier that favors the majority class may therefore have poor recall for
the rarer classes, as the simple decision tree below illustrates.

```{r }
tree_default <- Zoo |> 
  rpart(type ~ ., data = _)
tree_default

library(rpart.plot)
rpart.plot(tree_default, extra = 2)
```


The resulting tree has no leaf nodes for these rare animal types, so it cannot
predict them.

Predicting whether an animal is a reptile is a highly imbalanced binary
classification problem.
We set up this problem by changing the class variable to make it into a binary
reptile/non-reptile classification problem.

```{r }
Zoo_reptile <- Zoo |> 
  mutate(type = factor(Zoo$type == "reptile", 
                       levels = c(FALSE, TRUE),
                       labels = c("nonreptile", "reptile")))
```

Keep the class variable as a factor (categorical); otherwise `rpart` fits a
regression tree instead of a classification tree.

The class distribution shows that this classification problem is now
highly imbalanced, with 5 reptiles and 96 non-reptiles.


```{r }
summary(Zoo_reptile)
```

```{r }
ggplot(Zoo_reptile, aes(y = type)) + geom_bar()
```

The graph shows that a non-reptile is about 20 times as common as a reptile in
this dataset. We will compare several ways to address this imbalance.

We split the data into training and test sets so we can evaluate each approach
on observations not used for fitting. A 50/50 split increases the chance that
the small test set includes reptiles. Estimates remain highly uncertain
because there are only five reptiles in total.

```{r }
set.seed(1234)

inTrain <- createDataPartition(y = Zoo_reptile$type, p = .5)[[1]]
training_reptile <- Zoo_reptile |> slice(inTrain)
testing_reptile <- Zoo_reptile |> slice(-inTrain)
```

### Option 1: Use the Original Class Distribution

```{r }
fit <- training_reptile |>
  train(type ~ .,
        data = _,
        method = "rpart",
        trControl = trainControl(method = "cv"))
```

**Warnings:** "There were missing values in resampled performance
measures." means that some test folds used in hyper parameter tuning 
did not contain examples of both
classes. This is very likely with strong class imbalance and small datasets.

```{r }
fit
rpart.plot(fit$finalModel, extra = 2)
```

The tree predicts everything as non-reptile. Have a look at the error on
the test set.

```{r }
confusionMatrix(data = predict(fit, testing_reptile),
                ref = testing_reptile$type, positive = "reptile")
```

Accuracy equals the no-information rate and kappa is zero. Sensitivity is also
zero: the model identifies no reptiles. When missing a positive case is more
costly than misclassifying a negative, accuracy alone is not an adequate
metric. Sensitivity, specificity, and the costs of each type of error should
also be considered.

**Note:** The positive class value (the one that you want to detect) is
set manually to reptile using `positive = "reptile"`. Otherwise
sensitivity/specificity will not be correctly calculated.

### Option 2: Upsample Within Resampling

A common approach is to upsample the minority class by drawing observations
with replacement. Upsampling must happen within each resampling iteration;
otherwise, duplicate observations can appear in both the fold used for fitting
and the fold used for validation. `caret` supports this through the `sampling`
argument to `trainControl()` (see the [caret subsampling
guide](https://topepo.github.io/caret/subsampling-for-class-imbalances.html)).

```{r }
set.seed(1000) # for repeatability

fit <- training_reptile |>
  train(type ~ .,
        data = _,
        method = "rpart",
        trControl = trainControl(method = "cv", sampling = "up"),
        control = rpart.control(minsplit = 5))

fit
rpart.plot(fit$finalModel, extra = 2)
```

Upsampling is applied only to each analysis fold; validation folds and the
held-out test set keep the observed class distribution. Evaluate the fitted
model on that untouched test set:

```{r }
confusionMatrix(data = predict(fit, testing_reptile),
                ref = testing_reptile$type, positive = "reptile")
```

Compare accuracy, sensitivity, and specificity with the unresampled model.
Because the test set contains very few reptiles, each metric can change
substantially if even one observation is classified differently.

Upsampling changes the class balance seen during fitting, but it does not
directly specify the relative costs of false positives and false negatives.
Choose a decision threshold or use a cost-sensitive method when you need to
express that tradeoff; see Options 3 and 4.

### Option 3: Build A Larger Tree and use Predicted Probabilities

We increase tree complexity and require fewer observations to split a node.
Here we tune using the area under the ROC curve (ROC AUC). `twoClassSummary`
requires class probabilities, which we also need for the threshold example
below. The tree is still fit using its standard splitting criterion; ROC AUC
is the metric used to select the tuning parameter.

```{r }
fit <- training_reptile |> 
  train(type ~ .,
        data = _,
        method = "rpart",
        tuneLength = 10,
        trControl = trainControl(
                        method = "cv",
                        classProbs = TRUE,  ## for predict with type="prob"
                        summaryFunction=twoClassSummary),  ## for ROC
        metric = "ROC",
        control = rpart.control(minsplit = 3))
fit

rpart.plot(fit$finalModel, extra = 2)

confusionMatrix(data = predict(fit, testing_reptile),
                ref = testing_reptile$type, positive = "reptile")
```

**Note:** Accuracy is close to the no-information rate on this small dataset.

#### Create A Biased Classifier

We can lower the classification threshold to identify more reptiles at the
cost of misclassifying more non-reptiles. For a binary classifier, the default
threshold is often 0.5. Lowering it predicts more positives, typically
increasing sensitivity at the cost of specificity. The 0.01 threshold below
is illustrative; a real threshold should be selected using validation data
and application-specific costs.

```{r }
prob <- predict(fit, testing_reptile, type = "prob")
tail(prob)
pred <- ifelse(prob[,"reptile"] >= 0.01, "reptile", "nonreptile") |> 
  as.factor()

confusionMatrix(data = pred,
                ref = testing_reptile$type, positive = "reptile")
```

This lower threshold trades accuracy and specificity for sensitivity on this
split. Accuracy treats all errors equally and does not capture the cost of
missing a reptile.

#### Plot the ROC Curve

Since we have a binary classification problem and a classifier that
predicts a probability for an observation to be a reptile, we can also
use a [receiver operating characteristic
(ROC)](https://en.wikipedia.org/wiki/Receiver_operating_characteristic)
curve. The ROC curve shows sensitivity against the false-positive rate over
different probability thresholds. Its area (ROC AUC) summarizes ranking
performance: 0.5 corresponds to chance ranking and 1 to perfect ranking. It
does not select an operating threshold or account for application-specific
error costs.

```{r }
library("pROC")
r <- roc(testing_reptile$type == "reptile", prob[,"reptile"])
r

ggroc(r) + geom_abline(intercept = 1, slope = 1, color = "darkgrey")
```

This ROC curve is calculated on the test set for illustration. Do not choose a
threshold from this curve and then report performance on the same test set; use
validation data to choose the threshold and reserve the test set for the final
evaluation.

### Option 4: Use a Cost-Sensitive Classifier

The CART implementation in `rpart` accepts a loss matrix through the `loss`
parameter. Rows represent actual classes and columns represent predicted
classes. The diagonal entries (correct predictions) have zero cost. We assign
a false negative a cost of 100 and a false positive a cost of 1:

```
                 predicted
actual       nonreptile reptile
nonreptile       0         1
reptile         100        0
```

```{r }
cost <- matrix(c(
  0,   1,
  100, 0
), byrow = TRUE, nrow = 2)
cost

fit <- training_reptile |> 
  train(type ~ .,
        data = _,
        method = "rpart",
        parms = list(loss = cost),
        trControl = trainControl(method = "cv"))
```

The warning *"There were missing values in resampled performance
measures"* means that some validation folds did not contain both classes.
With only a few reptiles, this can make cross-validation metrics unstable or
undefined.

```{r }
fit

rpart.plot(fit$finalModel, extra = 2)

confusionMatrix(data = predict(fit, testing_reptile),
                ref = testing_reptile$type, positive = "reptile")
```

In this split, the high false-negative cost results in a classifier that
identifies all reptiles. This result is specific to this small split; it does
not guarantee perfect sensitivity on new data.

Cost-sensitive fitting is useful when error costs are known and the chosen
algorithm supports them. Other approaches include adjusting a decision
threshold or using weighted loss functions; availability depends on the model
and software implementation.


## Comparing Decision Boundaries of Popular Classification Techniques*

Classifiers create decision boundaries to discriminate between classes.
Different classifiers are able to create different shapes of decision
boundaries (e.g., some are strictly linear) and thus some classifiers
may perform better for certain datasets. In this section, we visualize the
decision boundaries found by several popular classification methods.

The following function defines a plot that adds the decision boundary 
(black lines) and the difference between the classification probability 
of the two best classes
(color intensity; indifference is 0 shown as white) by evaluating 
the classifier
at evenly spaced grid points. Note that low resolution will make
evaluation faster but it also will make the decision boundary look like it has
small steps even if it is a straight line.

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(ggplot2)
library(scales)
library(caret)
```

```{r }
decisionplot <- function(model, data, class_var, 
  predict_type = c("class", "prob"), resolution = 3 * 72) {
  # resolution is set to 72 dpi for 3 inches wide images. 
  
  y <- data |> pull(class_var)
  x <- data |> dplyr::select(-all_of(class_var))
  
  # resubstitution accuracy
  prediction <- predict(model, x, type = predict_type[1])
  
  # LDA returns a list
  if(is.list(prediction)) prediction <- prediction$class
  
  prediction <- factor(prediction, levels = levels(y))
  cm <- confusionMatrix(data = prediction, 
                        reference = y)
  acc <- cm$overall["Accuracy"]
  
  # evaluate model on a grid
  r <- sapply(x[, 1:2], range, na.rm = TRUE)
  xs <- seq(r[1,1], r[2,1], length.out = resolution)
  ys <- seq(r[1,2], r[2,2], length.out = resolution)
  g <- cbind(rep(xs, each = resolution), rep(ys, 
                                             time = resolution))
  colnames(g) <- colnames(r)
  g <- as_tibble(g)
  
  # guess how to get class labels from predict
  # (unfortunately not very consistent between models)
  cl <- predict(model, g, type = predict_type[1])
    
  prob <- NULL
  if(is.list(cl)) { # LDA returns a list
    prob <- cl$posterior
    cl <- cl$class
  } else if (inherits(model, "svm")) 
    prob <- attr(predict(model, g, probability = TRUE), "probabilities")
  else 
    if(!is.na(predict_type[2]))
      try(prob <- predict(model, g, type = predict_type[2]))
  
  # We visualize the difference in probability/score between
  # the winning class and the second best class. We only use
  # probability if the classifier's predict function supports it.
  delta_prob <- 1
  if(!is.null(prob))
    
    try({
      if (any(rowSums(prob) != 1))
        prob <- cbind(prob, 1 - rowSums(prob))
    
      max_prob <- t(apply(prob, MARGIN = 1, sort, decreasing = TRUE))
      delta_prob <- max_prob[,1] - max_prob[,2]
    }, silent = TRUE) 
  
  cl <- factor(cl, levels = levels(y))
  
  g <- g |> add_column(prediction = cl, 
                       delta_prob = delta_prob)
  
  ggplot(g, mapping = aes(
    x = .data[[colnames(g)[1]]], 
    y = .data[[colnames(g)[2]]])) +
    geom_raster(mapping = aes(fill = prediction, 
                              alpha = delta_prob)) +
    geom_contour(mapping = aes(z = as.numeric(prediction)), 
      bins = length(levels(cl)), 
      linewidth = .5, 
      color = "black") +
    geom_point(data = data, mapping =  aes(
      x = .data[[colnames(data)[1]]], 
      y = .data[[colnames(data)[2]]],
      shape = .data[[class_var]]), 
      alpha = .7) + 
    scale_alpha_continuous(range = c(0,1), 
                           limits = c(0,1)) + 
    labs(subtitle = paste("Training accuracy:", round(acc, 2)))
}
```

### Iris Dataset

For easier visualization, we use two dimensions of the Iris dataset.

```{r }
set.seed(1000)
data(iris)
iris <- as_tibble(iris)

x <- iris |> dplyr::select(Sepal.Length, Sepal.Width, Species)
# Note: package MASS overwrites the select function.

x

ggplot(x, aes(x = Sepal.Length, 
              y = Sepal.Width, 
              fill = Species)) +  
  stat_density_2d(geom = "polygon", 
                  aes(alpha = after_stat(level))) +
  geom_point()
```

This is the original data.
Color is used to show the density.
Note that there is some overplotting with several points in the same position. 
You could use `geom_jitter()` instead of `geom_point()`.

#### Nearest Neighbor Classifier

We try several values for $k$.

```{r }
model <- x |> caret::knn3(Species ~ ., data = _, k = 1)
decisionplot(model, x, class_var = "Species") + 
  labs(title = "kNN (1 neighbor)")

model <- x |> caret::knn3(Species ~ ., data = _, k = 3)
decisionplot(model, x, class_var = "Species") + 
  labs(title = "kNN (3 neighbors)")

model <- x |> caret::knn3(Species ~ ., data = _, k = 9)
decisionplot(model, x, class_var = "Species") + 
  labs(title = "kNN (9 neighbors)")
```

Increasing $k$ smooths the decision boundary. At $k=1$, we see white
areas around points where flowers of two classes are in the same spot.
Here, the algorithm randomly chooses a class during prediction resulting
in the meandering decision boundary. The predictions in that area are
not stable and every time we ask for a class, we may get a different
class.

Note: The crazy lines in white areas are an artifact of the visualization. Here the
classifier randomly selects a class.

#### Naive Bayes Classifier

Use a Gaussian naive Bayes classifier.

```{r }
model <- x |> e1071::naiveBayes(Species ~ ., data = _)
decisionplot(model, x, class_var = "Species", 
             predict_type = c("class", "raw")) + 
  labs(title = "Naive Bayes")
```

The GBN finds a good model with the advantage that no hyperparameters are needed.

#### Linear Discriminant Analysis

LDA finds linear decision boundaries.

```{r }
model <- x |> MASS::lda(Species ~ ., data = _)
decisionplot(model, x, class_var = "Species") + 
  labs(title = "LDA")
```

Linear decision boundaries work for this dataset so LDA works well.

#### Multinomial Logistic Regression

Multinomial logistic regression is an extension of logistic regression
to problems with more than two classes.

```{r }
model <- x |> nnet::multinom(Species ~., data = _)
decisionplot(model, x, class_var = "Species") + 
  labs(titel = "Multinomial Logistic Regression")
```


#### Decision Trees

Compare different types of decision trees.

```{r }
model <- x |> rpart::rpart(Species ~ ., data = _)
decisionplot(model, x, class_var = "Species") + 
  labs(title = "CART")

model <- x |> rpart::rpart(Species ~ ., data = _,
  control = rpart::rpart.control(cp = 0.001, minsplit = 1))
decisionplot(model, x, class_var = "Species") + 
  labs(title = "CART (overfitting)")

model <- x |> C50::C5.0(Species ~ ., data = _)
decisionplot(model, x, class_var = "Species") + 
  labs(title = "C5.0")
```

#### Ensemble: Random Forest

Use an ensemble method.

```{r }
model <- x |> randomForest::randomForest(Species ~ ., data = _)
decisionplot(model, x, class_var = "Species") + 
  labs(title = "Random Forest")
```

For the default settings for Random forest, the model seems to overfit the 
training data. More data would probably alleviate this issue.

#### Support Vector Machine

Compare SVMs with different kernel functions. 

```{r }
model <- x |> e1071::svm(Species ~ ., data = _, 
                         kernel = "linear", probability = TRUE)
decisionplot(model, x, class_var = "Species") + 
  labs(title = "SVM (linear kernel)")

model <- x |> e1071::svm(Species ~ ., data = _, 
                         kernel = "radial", probability = TRUE)
decisionplot(model, x, class_var = "Species") + 
  labs(title = "SVM (radial kernel)")

model <- x |> e1071::svm(Species ~ ., data = _, 
                         kernel = "polynomial", probability = TRUE)
decisionplot(model, x, class_var = "Species") + 
  labs(title = "SVM (polynomial kernel)")

model <- x |> e1071::svm(Species ~ ., data = _, 
                         kernel = "sigmoid", probability = TRUE)
decisionplot(model, x, class_var = "Species") + 
  labs(title = "SVM (sigmoid kernel)")
```


The linear SVM (without a kernel) produces straight lines and works well on the iris data.
Most kernels do also well, only the sigmoid kernel seems to find a
very strange decision boundary which indicates that the data does not have a
linear decision boundary 
in the projected space.



#### Single Layer Feed-forward Neural Networks

Use a simple network with one hidden layer. We will try a 
different number of 
neurons for the hidden layer.

```{r }
set.seed(1234)

model <- x |> nnet::nnet(Species ~ ., data = _, 
                        size = 1, trace = FALSE)
decisionplot(model, x, class_var  = "Species", 
             predict_type = c("class", "raw")) + 
  labs(title = "ANN (1 neuron)")

model <- x |> nnet::nnet(Species ~ ., data = _, 
                        size = 2, trace = FALSE)
decisionplot(model, x, class_var  = "Species", 
             predict_type = c("class", "raw")) + 
  labs(title = "ANN (2 neurons)")

model <- x |> nnet::nnet(Species ~ ., data = _, 
                        size = 4, trace = FALSE)
decisionplot(model, x, class_var  = "Species", 
             predict_type = c("class", "raw")) + 
  labs(title = "ANN (4 neurons)")

model <- x |> nnet::nnet(Species ~ ., data = _, 
                        size = 6, trace = FALSE)
decisionplot(model, x, class_var  = "Species", 
             predict_type = c("class", "raw")) + 
  labs(title = "ANN (6 neurons)")

model <- x |> nnet::nnet(Species ~ ., data = _, 
                        size = 20, trace = FALSE)
decisionplot(model, x, class_var  = "Species", 
             predict_type = c("class", "raw")) + 
  labs(title = "ANN (20 neurons)")

model <- x |> nnet::nnet(Species ~ ., data = _, 
                        size = 100, trace = FALSE)
decisionplot(model, x, class_var  = "Species", 
             predict_type = c("class", "raw")) + 
  labs(title = "ANN (100 neurons)")

model <- x |> nnet::nnet(Species ~ ., data = _, 
                        size = 100, trace = FALSE)
decisionplot(model, x, class_var  = "Species", 
             predict_type = c("class", "raw")) + 
  labs(title = "ANN 2nd try (100 neurons)")
```

For this simple data set, 2 neurons produce the best classifier. 
The model starts to overfit with 6 or more neurons. I ran the ANN twice with 100 
neurons, the two different decision boundaries indicate that the variability of
the ANN with so many neurons is high.

### Circle Dataset


```{r }
set.seed(1000)

x <- mlbench::mlbench.circle(500)

# You can also experiment with the following datasets.
#x <- mlbench::mlbench.cassini(500)
#x <- mlbench::mlbench.spirals(500, sd = .1)
#x <- mlbench::mlbench.smiley(500)

x <- cbind(as.data.frame(x$x), factor(x$classes))
colnames(x) <- c("x", "y", "class")
x <- as_tibble(x)
x

ggplot(x, aes(x = x, y = y, color = class)) + 
  geom_point()
```

This dataset is challenging for some classification algorithms since 
the optimal decision boundary is a circle around the class in the center.

#### Nearest Neighbor Classifier

Compare kNN classifiers with different values for $k$.


```{r }
model <- x |> caret::knn3(class ~ ., data = _, k = 1)
decisionplot(model, x, class_var = "class") + 
  labs(title = "kNN (1 neighbor)")

model <- x |> caret::knn3(class ~ ., data = _, k = 10)
decisionplot(model, x, class_var = "class") + 
  labs(title = "kNN (10 neighbors)")
```

k-Nearest does not find a smooth decision boundary, but tends to 
overfit the training data at low values for $k$.  

#### Naive Bayes Classifier

The Gaussian naive Bayes classifier works very well on the data. 


```{r }
model <- x |> e1071::naiveBayes(class ~ ., data = _)
decisionplot(model, x, class_var = "class", 
  predict_type = c("class", "raw")) + 
  labs(title = "naive Bayes")
```


#### Linear Discriminant Analysis


LDA cannot find a good model since the true decision boundary is not
linear.

```{r }
model <- x |> MASS::lda(class ~ ., data = _)
decisionplot(model, x, class_var = "class") + labs(title = "LDA")
```



#### Multinomial Logistic Regression

Multinomial logistic regression is an extension of logistic regression
to problems with more than two classes. 

```{r }
model <- x |> nnet::multinom(class ~., data = _)
decisionplot(model, x, class_var = "class") + 
  labs(titel = "Multinomial Logistic Regression")
```

Logistic regression also tries to find a linear
decision boundary and fails.

#### Decision Trees

Compare different decision tree algorithms.

```{r }
model <- x |> rpart::rpart(class ~ ., data = _)
decisionplot(model, x, class_var = "class") + 
  labs(title = "CART")

model <- x |> rpart::rpart(class ~ ., data = _,
  control = rpart::rpart.control(cp = 0, minsplit = 1))
decisionplot(model, x, class_var = "class") + 
  labs(title = "CART (overfitting)")

model <- x |> C50::C5.0(class ~ ., data = _)
decisionplot(model, x, class_var = "class") + 
  labs(title = "C5.0")
```

Decision trees do well with the restriction that they can only create
cuts parallel to the axes.

#### Ensemble: Random Forest

Try random forest on the dataset.

```{r}
library(randomForest)
model <- x |> randomForest(class ~ ., data = _)
decisionplot(model, x, class_var = "class") + 
  labs(title = "Random Forest")
```


#### Support Vector Machine

Compare SVMs with different kernels.

```{r }
model <- x |> e1071::svm(class ~ ., data = _, 
                         kernel = "linear", probability = TRUE)
decisionplot(model, x, class_var = "class") + 
  labs(title = "SVM (linear kernel)")

model <- x |> e1071::svm(class ~ ., data = _, 
                         kernel = "radial", probability = TRUE)
decisionplot(model, x, class_var = "class") + 
  labs(title = "SVM (radial kernel)")

model <- x |> e1071::svm(class ~ ., data = _, 
                         kernel = "polynomial", probability = TRUE)
decisionplot(model, x, class_var = "class") + 
  labs(title = "SVM (polynomial kernel)")

model <- x |> e1071::svm(class ~ ., data = _, 
                         kernel = "sigmoid", probability = TRUE)
decisionplot(model, x, class_var = "class") + 
  labs(title = "SVM (sigmoid kernel)")
```

A linear SVM does not work on this data set. An SMV with a radial kernel performs 
well, the other kernels have issues with finding a linear decision boundary in the 
projected space.

#### Single Layer Feed-forward Neural Networks

Use a simple network with one hidden layer. We will try a 
different number of 
neurons for the hidden layer.

```{r }
model <- x |> nnet::nnet(class ~ ., data = _, size = 1, trace = FALSE)
decisionplot(model, x, class_var = "class", 
  predict_type = c("class", "raw")) + labs(title = "ANN (1 neuron)")

model <- x |> nnet::nnet(class ~ ., data = _, size = 2, trace = FALSE)
decisionplot(model, x, class_var = "class", 
  predict_type = c("class", "raw")) + labs(title = "ANN (2 neurons)")

model <- x |> nnet::nnet(class ~ ., data = _, size = 4, trace = FALSE)
decisionplot(model, x, class_var = "class", 
  predict_type = c("class", "raw")) + labs(title = "ANN (4 neurons)")

model <- x |> nnet::nnet(class ~ ., data = _, size = 6, trace = FALSE)
decisionplot(model, x, class_var = "class", 
  predict_type = c("class", "raw")) + labs(title = "ANN (6 neurons)")
model <- x |> nnet::nnet(class ~ ., data = _, size = 20, trace = FALSE)
decisionplot(model, x, class_var = "class", 
  predict_type = c("class", "raw")) + labs(title = "ANN (20 neurons)")
```

The plots show that a network with 4 and 6 neurons performs well, while
a larger number of neurons leads to overfitting the training data.

## More Information on Classification with R

-   The `caret` framework used in this book: <https://topepo.github.io/caret/>
-   Alternative modeling framework (not used in these examples):
    <https://www.tidymodels.org/>
-   R taskview on machine learning:
    <http://cran.r-project.org/web/views/MachineLearning.html>

## Exercises*

We will again use the Palmer penguin data for the exercises.

```{r }
library(palmerpenguins)
head(penguins)
```

Create a R markdown file with the code and do the following below.

1. Apply at least 3 different classification models to the data.
2. Compare the models and a simple baseline model. Which model 
  performs the best? Does it perform significantly better than the other 
  models?

## Chapter Summary

Different classifiers make different assumptions and produce different
decision boundaries. Compare them against a simple baseline using the same
training and evaluation splits. With imbalanced classes, accuracy alone can be
misleading; resampling, costs, and probability thresholds change the balance
between detecting positives and avoiding false alarms.

## Key Terms

* **Baseline classifier:** a simple reference model used to contextualize
  performance.
* **Class imbalance:** a large difference in the number of examples per class.
* **Decision boundary:** the feature-space boundary separating predicted
  classes.
* **Sensitivity:** the fraction of actual positive cases correctly identified.
* **Threshold:** a probability cutoff used to turn predicted probabilities
  into class labels.
