---
editor_options: 
  markdown: 
    wrap: 72
---

# Classification: Alternative Techniques

Install the packages used in this chapter:

```{r setup_04}
pkgs <- sort(c('tidyverse', 'caret', 'RWeka', 'lattice', 'scales',
  'e1071', 'MASS', 'nnet', 'rpart', 'C50', 'randomForest', 'keras',
  'mlbench'
  ))

pkgs_install <- pkgs[!(pkgs %in% installed.packages()[,"Package"])]
if(length(pkgs_install)) install.packages(pkgs_install)
```

The packages used for this chapter are: `r format_pkgs(pkgs)`

```{r setup_04-2, include=FALSE}
all_pkgs <- union(all_pkgs, pkgs)
```

We will use tidyverse to prepare the data.

```{r message=FALSE, warning=FALSE}
library(tidyverse)
```

Show fewer digits

```{r }
options(digits=3)
```

## Introduction

Many different [classification algorithms](https://en.wikipedia.org/wiki/Supervised_learning) 
have been proposed in the literature. 
In this chapter, we will apply some of the more popular methods.

## Training and Test Data

We will use the Zoo dataset which is included in the R package `mlbench`
(you may have to install it). The Zoo dataset containing 17 (mostly
logical) variables on different 101 animals as a data frame with 17
columns (hair, feathers, eggs, milk, airborne, aquatic, predator,
toothed, backbone, breathes, venomous, fins, legs, tail, domestic,
catsize, type). We convert the data frame into a tidyverse tibble
(optional).

```{r }
data(Zoo, package="mlbench")
Zoo <- as_tibble(Zoo)
Zoo
```

We will use the package [**caret**](https://topepo.github.io/caret/) to
make preparing training sets and building classification (and
regression) models easier. A great cheat sheet can be found
[here](https://ugoproto.github.io/ugo_r_doc/pdf/caret.pdf).

```{r message=FALSE, warning=FALSE}
library(caret)
```

Multi-core support can be used for cross-validation. **Note:** It is
commented out here because it does not work with rJava used in RWeka
below.

```{r }
##library(doMC, quietly = TRUE)
##registerDoMC(cores = 4)
##getDoParWorkers()
```

Test data is not used in the model building process and needs to be set
aside purely for testing the model after it is completely built. Here I
use 80% for training.

```{r }
inTrain <- createDataPartition(y = Zoo$type, p = .8)[[1]]
Zoo_train <- Zoo |> slice(inTrain)
Zoo_test <- Zoo |> slice(-inTrain)
```

## Fitting Different Classification Models to the Training Data

Create a fixed sampling scheme (10-folds) so we can compare the fitted
models later.

```{r }
train_index <- createFolds(Zoo_train$type, k = 10)
```

The fixed folds are used in `train()` with the argument
`trControl = trainControl(method = "cv", indexOut = train_index))`. If
you don't need fixed folds, then remove `indexOut = train_index` in the
code below.

For help with building models in caret see: `? train`

**Note:** Be careful if you have many `NA` values in your data.
`train()` and cross-validation many fail in some cases. If that is the
case then you can remove features (columns) which have many `NA`s, omit
`NA`s using `na.omit()` or use imputation to replace them with
reasonable values (e.g., by the feature mean or via kNN). Highly
imbalanced datasets are also problematic since there is a chance that a
fold does not contain examples of each class leading to a hard to
understand error message.

### Conditional Inference Tree (Decision Tree)

```{r }
ctreeFit <- Zoo_train |> train(type ~ .,
  method = "ctree",
  data = _,
	tuneLength = 5,
	trControl = trainControl(method = "cv", indexOut = train_index))
ctreeFit
plot(ctreeFit$finalModel)
```

### C 4.5 Decision Tree

```{r }
C45Fit <- Zoo_train |> train(type ~ .,
  method = "J48",
  data = _,
	tuneLength = 5,
	trControl = trainControl(method = "cv", indexOut = train_index))
C45Fit
C45Fit$finalModel
```

### K-Nearest Neighbors

**Note:** kNN uses Euclidean distance, so data should be standardized
(scaled) first. Here legs are measured between 0 and 6 while all other
variables are between 0 and 1. Scaling can be directly performed as
preprocessing in `train` using the parameter `preProcess = "scale"`.

```{r }
knnFit <- Zoo_train |> train(type ~ .,
  method = "knn",
  data = _,
  preProcess = "scale",
	tuneLength = 5,
  tuneGrid=data.frame(k = 1:10),
	trControl = trainControl(method = "cv", indexOut = train_index))
knnFit

knnFit$finalModel
```

### PART (Rule-based classifier)

```{r }
rulesFit <- Zoo_train |> train(type ~ .,
  method = "PART",
  data = _,
  tuneLength = 5,
  trControl = trainControl(method = "cv", indexOut = train_index))
rulesFit

rulesFit$finalModel
```

### Linear Support Vector Machines

```{r }
svmFit <- Zoo_train |> train(type ~.,
  method = "svmLinear",
  data = _,
	tuneLength = 5,
	trControl = trainControl(method = "cv", indexOut = train_index))
svmFit

svmFit$finalModel
```

### Random Forest

```{r }
randomForestFit <- Zoo_train |> train(type ~ .,
  method = "rf",
  data = _,
	tuneLength = 5,
	trControl = trainControl(method = "cv", indexOut = train_index))
randomForestFit

randomForestFit$finalModel
```

### Gradient Boosted Decision Trees (xgboost)

```{r }
xgboostFit <- Zoo_train |> train(type ~ .,
  method = "xgbTree",
  data = _,
  tuneLength = 5,
  trControl = trainControl(method = "cv", indexOut = train_index),
  tuneGrid = expand.grid(
    nrounds = 20,
    max_depth = 3,
    colsample_bytree = .6,
    eta = 0.1,
    gamma=0,
    min_child_weight = 1,
    subsample = .5
  ))
xgboostFit

xgboostFit$finalModel
```

### Artificial Neural Network

```{r }
nnetFit <- Zoo_train |> train(type ~ .,
  method = "nnet",
  data = _,
	tuneLength = 5,
	trControl = trainControl(method = "cv", indexOut = train_index),
  trace = FALSE)
nnetFit

nnetFit$finalModel
```

## Comparing Models

Collect the performance metrics from the models trained on the same
data.

```{r }
resamps <- resamples(list(
  ctree = ctreeFit,
  C45 = C45Fit,
  SVM = svmFit,
  KNN = knnFit,
  rules = rulesFit,
  randomForest = randomForestFit,
  xgboost = xgboostFit,
  NeuralNet = nnetFit
    ))
resamps
```

Calculate summary statistics

```{r }
summary(resamps)

library(lattice)
bwplot(resamps, layout = c(3, 1))
```

Perform inference about differences between models. For each metric, all
pair-wise differences are computed and tested to assess if the
difference is equal to zero. By default Bonferroni correction for
multiple comparison is used. Differences are shown in the upper triangle
and p-values are in the lower triangle.

```{r }
difs <- diff(resamps)
difs
summary(difs)
```

All perform similarly well except ctree (differences in the first row
are negative and the p-values in the first column are \<.05 indicating
that the null-hypothesis of a difference of 0 can be rejected).

## Applying the Chosen Model to the Test Data

Most models do similarly well on the data. We choose here the random
forest model.

```{r }
pr <- predict(randomForestFit, Zoo_test)
pr
```

Calculate the confusion matrix for the held-out test data.

```{r }
confusionMatrix(pr, reference = Zoo_test$type)
```

## Deep Learning with keras/tensorflow

The keras package needs the packages `reticulate` and `tensorflow`. To
install keras you need to

1.  have a working Python installation,
2.  install the keras R package `install.packages("keras")`, and
3.  install the tensorflow/keras Python modules with
    `library(keras); install_keras()`

```{r message=FALSE, warning=FALSE}
library(keras)
```

Prepare the data. All data needs to be in a matrix of all
numeric/integer values. The class variable needs to be one-hot encodes
with the `keras` function `to_categorical()`.

```{r warning=FALSE}
X <- Zoo_train |> 
  select(!type) |> 
  mutate(across(everything(), as.integer)) |> 
  as.matrix()
head(X)

y <- Zoo_train |> 
  pull("type") |> 
  as.integer() |> 
  {\(x)(x - 1L)}() |> 
  to_categorical()
head(y)

X_test <- Zoo_test |> 
  select(!type) |> 
  mutate(across(everything(), as.integer)) |> 
  as.matrix()
y_test <- Zoo_test |> 
  pull("type") |> 
  as.integer() |> 
  {\(x)(x - 1L)}() |>
  to_categorical()
```

Deep learning uses a large set of hyper-parameters. Choices are the
activation function, number of layers, number of units per layer, the
optimizer, and the type of regularization used. 

We define here a simple network with a single dense hidden layer.
Note that in keras, we have to specify the input size in this first dense layer. 
The output of the classifier is a categorical class value, therefore
we use for the output layer the softmax activation function, and for
training we specify 
categorical cross-entropy as the loss function, and accuracy
as an additional metric.
Also, L2 regularization is used for the weights of the hidden layer to
reduce overfitting. 

```{r message=FALSE, warning=FALSE}
model <- keras_model_sequential(input_shape = c(ncol(X)), 
              name = "single_hidden_layer_classifier") |>
  layer_dense(units = 10, activation = 'relu', 
              kernel_regularizer=regularizer_l2(l=0.01), name = "hidden1") |>
  layer_dense(units = ncol(y), activation = 'softmax', name = "output")

model
```


```{r message=FALSE, warning=FALSE}
model <- model |>  
  compile(loss = 'categorical_crossentropy', 
          optimizer = 'adam', 
          metrics = 'accuracy')
```

For model training, we need to specify the batch size and the number of
training epochs. The fitting process can also use a fraction of the
training data for validation to provide generalization loss/accuracy.

```{r}
history <- model |>
  fit(
    X, y,
    batch_size = 10,
    epochs = 100,
    validation_split = .2
  )

plot(history)
```

To create predictions from the model, we have to convert the one-hot
encoding back to class labels.

```{r}
class_labels <- levels(Zoo_train |> pull(type))

pr <- predict(model, X_test) |> apply(MARGIN = 1, FUN = which.max)
pr <- factor(pr, labels = class_labels, levels = seq_along(class_labels))

pr
```

```{r }
confusionMatrix(pr, reference = Zoo_test$type)
```

## Comparing Decision Boundaries of Popular Classification Techniques

Classifiers create decision boundaries to discriminate between classes.
Different classifiers are able to create different shapes of decision
boundaries (e.g., some are strictly linear) and thus some classifiers
may perform better for certain datasets. This page visualizes the
decision boundaries found by several popular classification methods.

The following plot adds the decision boundary (black lines) and
classification confidence (color intensity) by evaluating the classifier
at evenly spaced grid points. Note that low resolution (to make
evaluation faster) will make the decision boundary look like it has
small steps even if it is a (straight) line.

```{r message=FALSE, warning=FALSE}
library(scales)
library(tidyverse)
library(ggplot2)
library(caret)

decisionplot <- function(model, data, class_var, 
  predict_type = c("class", "prob"), resolution = 3 * 72) {
  # resolution is set to 72 dpi if the image is rendered  3 inches wide. 
  
  y <- data |> pull(class_var)
  x <- data |> dplyr::select(-all_of(class_var))
  
  # resubstitution accuracy
  prediction <- predict(model, x, type = predict_type[1])
  # LDA returns a list
  if(is.list(prediction)) prediction <- prediction$class
  prediction <- factor(prediction, levels = levels(y))
  
  cm <- confusionMatrix(data = prediction, reference = y)
  acc <- cm$overall["Accuracy"]
  
  # evaluate model on a grid
  r <- sapply(x[, 1:2], range, na.rm = TRUE)
  xs <- seq(r[1,1], r[2,1], length.out = resolution)
  ys <- seq(r[1,2], r[2,2], length.out = resolution)
  g <- cbind(rep(xs, each = resolution), rep(ys, time = resolution))
  colnames(g) <- colnames(r)
  g <- as_tibble(g)
  
  ### guess how to get class labels from predict
  ### (unfortunately not very consistent between models)
  cl <- predict(model, g, type = predict_type[1])
  
  # LDA returns a list
  prob <- NULL
  if(is.list(cl)) { 
    prob <- cl$posterior
    cl <- cl$class
  } else
    if(!is.na(predict_type[2]))
      try(prob <- predict(model, g, type = predict_type[2]))
  
  # we visualize the difference in probability/score between the 
  # winning class and the second best class.
  # don't use probability if predict for the classifier does not support it.
  max_prob <- 1
  if(!is.null(prob))
    try({
      max_prob <- t(apply(prob, MARGIN = 1, sort, decreasing = TRUE))
      max_prob <- max_prob[,1] - max_prob[,2]
    }, silent = TRUE) 
  
  cl <- factor(cl, levels = levels(y))
  
  g <- g |> add_column(prediction = cl, probability = max_prob)
  
  ggplot(g, mapping = aes(
    x = .data[[colnames(g)[1]]], y = .data[[colnames(g)[2]]])) +
    geom_raster(mapping = aes(fill = prediction, alpha = probability)) +
    geom_contour(mapping = aes(z = as.numeric(prediction)), 
      bins = length(levels(cl)), size = .5, color = "black") +
    geom_point(data = data, mapping =  aes(
      x = .data[[colnames(data)[1]]], 
      y = .data[[colnames(data)[2]]],
      shape = .data[[class_var]]), alpha = .7) + 
    scale_alpha_continuous(range = c(0,1), limits = c(0,1), guide = "none") +  
    labs(subtitle = paste("Training accuracy:", round(acc, 2)))
}
```

### Iris Dataset

For easier visualization, we use two dimensions of the Iris dataset.
Contour lines visualize the density like mountains on a map.

```{r }
set.seed(1000)
data(iris)
iris <- as_tibble(iris)

### Three classes (MASS also has a select function)
x <- iris |> dplyr::select(Sepal.Length, Sepal.Width, Species)
x

ggplot(x, aes(x = Sepal.Length, y = Sepal.Width, fill = Species)) +  
  stat_density_2d(geom = "polygon", aes(alpha = ..level..)) +
  geom_point()
```

*Note:* There is some overplotting and you could use `geom_jitter()`
instead of `geom_point()`.

#### K-Nearest Neighbors Classifier

```{r }
model <- x |> caret::knn3(Species ~ ., data = _, k = 1)
decisionplot(model, x, class_var = "Species") + 
  labs(title = "kNN (1 neighbor)")

model <- x |> caret::knn3(Species ~ ., data = _, k = 3)
decisionplot(model, x, class_var = "Species") + 
  labs(title = "kNN (3 neighbor)")

model <- x |> caret::knn3(Species ~ ., data = _, k = 9)
decisionplot(model, x, class_var = "Species") + 
  labs(title = "kNN (9 neighbor)")
```

Increasing $k$ smooths the decision boundary. At $k=1$, we see white
areas around points where flowers of two classes are in the same spot.
Here, the algorithm randomly chooses a class during prediction resulting
in the meandering decision boundary. The predictions in that area are
not stable and every time we ask for a class, we may get a different
class.

#### Naive Bayes Classifier

```{r }
model <- x |> e1071::naiveBayes(Species ~ ., data = _)
decisionplot(model, x, class_var = "Species", 
             predict_type = c("class", "raw")) + 
  labs(title = "Naive Bayes")
```

#### Linear Discriminant Analysis

```{r }
model <- x |> MASS::lda(Species ~ ., data = _)
decisionplot(model, x, class_var = "Species") + 
  labs(title = "LDA")
```

#### Multinomial Logistic Regression (implemented in nnet)

Multinomial logistic regression is an extension of logistic regression
to problems with more than two classes.

```{r }
model <- x |> nnet::multinom(Species ~., data = _)
decisionplot(model, x, class_var = "Species") + 
  labs(titel = "Multinomial Logistic Regression")
```

#### Decision Trees

```{r }
model <- x |> rpart::rpart(Species ~ ., data = _)
decisionplot(model, x, class_var = "Species") + 
  labs(title = "CART")

model <- x |> rpart::rpart(Species ~ ., data = _,
  control = rpart.control(cp = 0.001, minsplit = 1))
decisionplot(model, x, class_var = "Species") + 
  labs(title = "CART (overfitting)")

model <- x |> C50::C5.0(Species ~ ., data = _)
decisionplot(model, x, class_var = "Species") + 
  labs(title = "C5.0")

model <- x |> randomForest::randomForest(Species ~ ., data = _)
decisionplot(model, x, class_var = "Species") + 
  labs(title = "Random Forest")
```

#### SVM

```{r }
model <- x |> e1071::svm(Species ~ ., data = _, kernel = "linear")
decisionplot(model, x, class_var = "Species") + 
  labs(title = "SVM (linear kernel)")

model <- x |> e1071::svm(Species ~ ., data = _, kernel = "radial")
decisionplot(model, x, class_var = "Species") + 
  labs(title = "SVM (radial kernel)")

model <- x |> e1071::svm(Species ~ ., data = _, kernel = "polynomial")
decisionplot(model, x, class_var = "Species") + 
  labs(title = "SVM (polynomial kernel)")

model <- x |> e1071::svm(Species ~ ., data = _, kernel = "sigmoid")
decisionplot(model, x, class_var = "Species") + 
  labs(title = "SVM (sigmoid kernel)")
```

#### Single Layer Feed-forward Neural Networks

```{r }
model <-x |> nnet::nnet(Species ~ ., data = _, size = 1, trace = FALSE)
decisionplot(model, x, class_var  = "Species", 
  predict_type = c("class", "raw")) + labs(title = "NN (1 neuron)")

model <-x |> nnet::nnet(Species ~ ., data = _, size = 2, trace = FALSE)
decisionplot(model, x, class_var  = "Species", 
  predict_type = c("class", "raw")) + labs(title = "NN (2 neurons)")

model <-x |> nnet::nnet(Species ~ ., data = _, size = 4, trace = FALSE)
decisionplot(model, x, class_var  = "Species", 
  predict_type = c("class", "raw")) + labs(title = "NN (4 neurons)")

model <-x |> nnet::nnet(Species ~ ., data = _, size = 10, trace = FALSE)
decisionplot(model, x, class_var  = "Species", 
  predict_type = c("class", "raw")) + labs(title = "NN (10 neurons)")
```

### Circle Dataset

This set is not linearly separable!

```{r }
set.seed(1000)

x <- mlbench::mlbench.circle(500)
###x <- mlbench::mlbench.cassini(500)
###x <- mlbench::mlbench.spirals(500, sd = .1)
###x <- mlbench::mlbench.smiley(500)
x <- cbind(as.data.frame(x$x), factor(x$classes))
colnames(x) <- c("x", "y", "class")
x <- as_tibble(x)
x

ggplot(x, aes(x = x, y = y, color = class)) + 
  geom_point()
```

#### K-Nearest Neighbors Classifier

```{r }
model <- x |> caret::knn3(class ~ ., data = _, k = 1)
decisionplot(model, x, class_var = "class") + 
  labs(title = "kNN (1 neighbor)")

model <- x |> caret::knn3(class ~ ., data = _, k = 10)
decisionplot(model, x, class_var = "class") + 
  labs(title = "kNN (10 neighbor)")
```

#### Naive Bayes Classifier

```{r }
model <- x |> e1071::naiveBayes(class ~ ., data = _)
decisionplot(model, x, class_var = "class", 
  predict_type = c("class", "raw")) + 
  labs(title = "naive Bayes")
```

#### Linear Discriminant Analysis

LDA cannot find a good model since the true decision boundary is not
linear.

```{r }
model <- x |> MASS::lda(class ~ ., data = _)
decisionplot(model, x, class_var = "class") + labs(title = "LDA")
```

#### Logistic Regression (implemented in nnet)

Multinomial logistic regression is an extension of logistic regression
to problems with more than two classes. It also tries to find a linear
decision boundary.

```{r }
model <- x |> nnet::multinom(class ~., data = _)
decisionplot(model, x, class_var = "class") + 
  labs(titel = "Multinomial Logistic Regression")
```

#### Decision Trees

```{r }
model <- x |> rpart::rpart(class ~ ., data = _)
decisionplot(model, x, class_var = "class") + 
  labs(title = "CART")

model <- x |> rpart::rpart(class ~ ., data = _,
  control = rpart.control(cp = 0.001, minsplit = 1))
decisionplot(model, x, class_var = "class") + 
  labs(title = "CART (overfitting)")

model <- x |> C50::C5.0(class ~ ., data = _)
decisionplot(model, x, class_var = "class") + 
  labs(title = "C5.0")

library(randomForest)
model <- x |> randomForest(class ~ ., data = _)
decisionplot(model, x, class_var = "class") + 
  labs(title = "Random Forest")
```

#### SVM

Linear SVM does not work for this data.

```{r }
model <- x |> e1071::svm(class ~ ., data = _, kernel = "linear")
decisionplot(model, x, class_var = "class") + 
  labs(title = "SVM (linear kernel)")

model <- x |> e1071::svm(class ~ ., data = _, kernel = "radial")
decisionplot(model, x, class_var = "class") + 
  labs(title = "SVM (radial kernel)")

model <- x |> e1071::svm(class ~ ., data = _, kernel = "polynomial")
decisionplot(model, x, class_var = "class") + 
  labs(title = "SVM (polynomial kernel)")

model <- x |> e1071::svm(class ~ ., data = _, kernel = "sigmoid")
decisionplot(model, x, class_var = "class") + 
  labs(title = "SVM (sigmoid kernel)")
```

#### Single Layer Feed-forward Neural Networks

```{r }
model <-x |> nnet::nnet(class ~ ., data = _, size = 1, trace = FALSE)
decisionplot(model, x, class_var = "class", 
  predict_type = c("class")) + labs(title = "NN (1 neuron)")

model <-x |> nnet::nnet(class ~ ., data = _, size = 2, trace = FALSE)
decisionplot(model, x, class_var = "class", 
  predict_type = c("class")) + labs(title = "NN (2 neurons)")

model <-x |> nnet::nnet(class ~ ., data = _, size = 4, trace = FALSE)
decisionplot(model, x, class_var = "class", 
  predict_type = c("class")) + labs(title = "NN (4 neurons)")

model <-x |> nnet::nnet(class ~ ., data = _, size = 10, trace = FALSE)
decisionplot(model, x, class_var = "class", 
  predict_type = c("class")) + labs(title = "NN (10 neurons)")
```

## More Information on Classification with R

-   Package caret: <http://topepo.github.io/caret/index.html>
-   Tidymodels (machine learning with tidyverse):
    <https://www.tidymodels.org/>
-   R taskview on machine learning:
    <http://cran.r-project.org/web/views/MachineLearning.html>
