---
editor_options: 
  markdown: 
    wrap: 72
---

# Association Analysis: Basic Concepts and Algorithms

Install the packages used in this chapter:

```{r setup_05}
pkgs <- c("arules", "arulesViz", "mlbench", "palmerpenguins", "tidyverse")

pkgs_install <- pkgs[!(pkgs %in% installed.packages()[,"Package"])]
if(length(pkgs_install)) install.packages(pkgs_install)
```

```{r setup_05-2, include=FALSE}
source("_common.R")
source("format_pkgs.R")
all_pkgs <- union(all_pkgs, pkgs)
```

The packages used for this chapter are: `r format_pkgs(pkgs)`


## Preliminaries

[Association rule mining](https://en.wikipedia.org/wiki/Association_rule_learning) 
plays a vital role in discovering hidden patterns and relationships within large
transactional datasets. Applications range from exploratory data analysis in marketing to building rule-based
classifiers.
@arules:Agrawal+Imielinski+Swami:1993 introduced the problem
of mining association rules from transaction data as follows (the definition is taken from @arules:Hahsler+Gruen+Hornik:2005b):

Let $I = \{i_1,i_2,...,i_n\}$ be a set of $n$ binary attributes called items. Let $D = \{t_1,t_2,...,t_m\}$ be
a set of transactions called the database. Each transaction in $D$ has a unique transaction ID and
contains a subset of the items in $I$. A rule is defined as an implication of the form $X \Rightarrow Y$ where
$X,Y \subseteq I$ and $X \cap Y = \emptyset$ are called itemsets. On itemsets and rules several quality measures can
be defined. The most important measures are support and confidence. The support $supp(X)$ of
an itemset $X$ is defined as the proportion of transactions in the data set which contain the itemset.
Itemsets with a support which surpasses a user-defined threshold $\sigma$ are called frequent itemsets. The
confidence of a rule is defined as $conf(X \Rightarrow Y) = supp(X \cup Y)/supp(X)$. Association rules are rules
with $supp(X \cup Y) \ge \sigma$ and $conf(X) \ge \delta$ where $\sigma$ and $\delta$ are user-defined thresholds.
The found set of association rules is then used reason about the data.

You can read the free sample chapter from the textbook [@Tan2005]:
[Chapter 5. Association Analysis: Basic Concepts and
Algorithms](https://www-users.cs.umn.edu/~kumar001/dmbook/ch5_association_analysis.pdf)

### The arules Package

Association rule mining in R is implemented in the package `arules`.

```{r }
library(tidyverse)
library(arules)
library(arulesViz)
```

For information about the `arules` package try: `help(package="arules")`
and `vignette("arules")` (also available at
[CRAN](http://cran.r-project.org/web/packages/arules/vignettes/arules.pdf))

arules uses the S4 object system to implement classes and methods.
Standard R objects use the [S3 object
system](http://adv-r.had.co.nz/S3.html) which do not use formal class
definitions and are usually implemented as a list with a class
attribute. `arules` and many other R packages use the [S4 object
system](http://adv-r.had.co.nz/S4.html) which is based on formal class
definitions with member variables and methods (similar to
object-oriented programming languages like Java and C++). Some important
differences of using S4 objects compared to the usual S3 objects are:

-   coercion (casting): `as(from, "class_name")`
-   help for classes: `class? class_name`

### Transactions

#### Create Transactions

We will use the Zoo dataset from `mlbench`.

```{r }
data(Zoo, package = "mlbench")
head(Zoo)
```

The data in the data.frame need to be converted into a set of
transactions where each row represents a transaction and each column is
translated into items. This is done using the constructor
`transactions()`. For the Zoo data set this means that we consider
animals as transactions and the different traits (features) will become
items that each animal has. For example the animal *antelope* has the
item *hair* in its transaction.

```{r }
trans <- transactions(Zoo)
```

The conversion gives a warning because only discrete features (`factor`
and `logical`) can be directly translated into items. Continuous
features need to be discretized first.

What is column 13?

```{r }
summary(Zoo[13])
ggplot(Zoo, aes(legs)) + geom_bar()
Zoo$legs |> table()
```

Possible solution: Make legs into has/does not have legs

```{r }
Zoo_has_legs <- Zoo |> mutate(legs = legs > 0)
ggplot(Zoo_has_legs, aes(legs)) + geom_bar()
Zoo_has_legs$legs |> table()
```

**Alternatives:**

-   use each unique value as an item:

```{r }
Zoo_unique_leg_values <- Zoo |> mutate(legs = factor(legs))
Zoo_unique_leg_values$legs |> head()
```

-   discretize (see
    [`? discretize`](https://www.rdocumentation.org/packages/arules/topics/discretize)
    and [discretization in the code for Chapter
    2](chap2.html#discretize-features)):

```{r }
Zoo_discretized_legs <- Zoo |> mutate(
  legs = discretize(legs, breaks = 2, method="interval")
)
table(Zoo_discretized_legs$legs)
```

Convert data into a set of transactions

```{r }
trans <- transactions(Zoo_has_legs)
trans
```

#### Inspect Transactions

```{r }
summary(trans)
```

Look at created items. They are still called column names since the
transactions are actually stored as a large sparse logical matrix (see
below).

```{r }
colnames(trans)
```

Compare with the original features (column names) from Zoo

```{r }
colnames(Zoo)
```

Look at a (first) few transactions as a matrix. 1 indicates the presence
of an item.

```{r }
as(trans, "matrix")[1:3,]
```

Look at the transactions as sets of items

```{r }
inspect(trans[1:3])
```

Plot the binary matrix. Dark dots represent 1s.

```{r }
image(trans)
```

Look at the relative frequency (=support) of items in the data set. Here
we look at the 10 most frequent items.

```{r }
itemFrequencyPlot(trans,topN = 20)

ggplot(
  tibble(
    Support = sort(itemFrequency(trans, type = "absolute"), decreasing = TRUE),
    Item = seq_len(ncol(trans))
  ), aes(x = Item, y = Support)) + geom_line()
```

**Alternative encoding:** Also create items for FALSE (use factor)

```{r }
sapply(Zoo_has_legs, class)
Zoo_factors <- Zoo_has_legs |> mutate(across(where(is.logical), factor))
sapply(Zoo_factors, class)
summary(Zoo_factors)

trans_factors <- transactions(Zoo_factors)
trans_factors

itemFrequencyPlot(trans_factors, topN = 20)

## Select transactions that contain a certain item
trans_insects <- trans_factors[trans %in% "type=insect"]
trans_insects
inspect(trans_insects)
```

#### Vertical Layout (Transaction ID Lists)

The default layout for transactions is horizontal layout (i.e. each
transaction is a row). The vertical layout represents transaction data
as a list of transaction IDs for each item (= transaction ID lists).

```{r }
vertical <- as(trans, "tidLists")
as(vertical, "matrix")[1:10, 1:5]
```

## Frequent Itemset Generation

For this dataset we have already a huge number of possible itemsets

```{r }
2^ncol(trans)
```

Find frequent itemsets (target="frequent") with the default settings.

```{r }
its <- apriori(trans, parameter=list(target = "frequent"))
its
```

Default minimum support is .1 (10%). **Note:** We use here a very small
data set. For larger datasets the default minimum support might be to
low and you may run out of memory. You probably want to start out with a
higher minimum support like .5 (50%) and then work your way down.

```{r }
5/nrow(trans)
```

In order to find itemsets that effect 5 animals I need to go down to a
support of about 5%.

```{r }
its <- apriori(trans, parameter=list(target = "frequent", support = 0.05))
its
```

Sort by support

```{r }
its <- sort(its, by = "support")
its |> head(n = 10) |> inspect()
```

Look at frequent itemsets with many items (set breaks manually since
Automatically chosen breaks look bad)

```{r }
ggplot(tibble(`Itemset Size` = factor(size(its))), aes(`Itemset Size`)) + 
  geom_bar()
its[size(its) > 8] |> inspect()
```


## Rules Generation

We use the APRIORI algorithm (see
[`? apriori`](https://www.rdocumentation.org/packages/arules/topics/apriori))

```{r }
rules <- apriori(trans, parameter = list(support = 0.05, confidence = 0.9))
length(rules)

rules |> head() |> inspect()
rules |> head() |> quality()
```

Look at rules with highest lift

```{r }
rules <- sort(rules, by = "lift")
rules |> head(n = 10) |> inspect()
```

Create rules using the alternative encoding (with "FALSE" item)

```{r }
r <- apriori(trans_factors)
r
print(object.size(r), unit = "Mb")

inspect(r[1:10])
r |> head(n = 10, by = "lift") |> inspect()
```

### Calculate Additional Interest Measures

```{r }
interestMeasure(rules[1:10], measure = c("phi", "gini"),
  trans = trans)
```

Add measures to the rules

```{r }
quality(rules) <- cbind(quality(rules),
  interestMeasure(rules, measure = c("phi", "gini"),
    trans = trans))
```

Find rules which score high for Phi correlation

```{r }
rules |> head(by = "phi") |> inspect()
```

### Mine Using Templates

Sometimes it is beneficial to specify what items should be where in the
rule. For apriori we can use the parameter appearance to specify this
(see
[`? APappearance`](https://www.rdocumentation.org/packages/arules/topics/APappearance)).
In the following we restrict rules to an animal `type` in the RHS and
any item in the LHS.

```{r }
type <- grep("type=", itemLabels(trans), value = TRUE)
type

rules_type <- apriori(trans, appearance= list(rhs = type))

rules_type |> sort(by = "lift") |> head() |> inspect()
```

Saving rules as a CSV-file to be opened with Excel or other tools.

`write(rules, file = "rules.csv", quote = TRUE)`

## Compact Representation of Frequent Itemsets

Find maximal frequent itemsets (no superset if frequent)

```{r }
its_max <- its[is.maximal(its)]
its_max
its_max |> head(by = "support") |> inspect()
```

Find closed frequent itemsets (no superset if frequent)

```{r }
its_closed <- its[is.closed(its)]
its_closed

its_closed |> head(by = "support") |> inspect()

counts <- c(
  frequent=length(its),
  closed=length(its_closed),
  maximal=length(its_max)
)

ggplot(as_tibble(counts, rownames = "Itemsets"),
  aes(Itemsets, counts)) + geom_bar(stat = "identity")
```


## Association Rule Visualization*

Visualization is a very powerful approach to analyse large sets of 
mined association rules and frequent itemsets. We present here some options
to create static visualizations and inspect rule sets interactively.

### Static Visualizations

```{r }
library(arulesViz)
```

Default scatterplot

```{r }
plot(rules)
```

Note that some jitter (randomly move points) was added to show how many
rules have the same confidence and support value. Without jitter:

```{r }
plot(rules, control = list(jitter = 0))

plot(rules, shading = "order")
##plot(rules, interactive = TRUE)
```

Grouped plot

```{r }
plot(rules, method = "grouped")
##plot(rules, method = "grouped", engine = "interactive")
```

As a graph

```{r }
plot(rules, method = "graph")
plot(rules |> head(by = "phi", n = 100), method = "graph")
```

### Interactive Visualizations

We will use the association rules mined from the Iris dataset for the
following examples.

```{r }
data(iris)
summary(iris)
```

Convert the data to transactions. 

```{r }
iris_trans <- transactions(iris)
```
Note that this conversion gives a warning to indicate that some potentially 
unwanted conversion happens. Some features are numeric and
need to be discretized. The conversion automatically applies
frequency-based discretization with 3 classes to each numeric feature,
however, the use may want to use a different discretization strategy.

```{r}
iris_trans |> head() |> inspect()
```

Next, we mine association rules.

```{r }
rules <- apriori(iris_trans, parameter = list(support = 0.1, confidence = 0.8))
rules
```

#### Interactive Inspect With Sorting, Filtering and Paging

```{r }
inspectDT(rules,options = list(scrollX = TRUE))
```

#### Scatter Plot

Plot rules as a scatter plot using an interactive html widget. To avoid
overplotting, jitter is added automatically. Set `jitter = 0` to disable
jitter. Hovering over rules shows rule information. *Note:*
plotly/javascript does not do well with too many points, so plot selects
the top 1000 rules with a warning if more rules are supplied.

```{r }
plot(rules, engine = "html")
```

#### Matrix Visualization

Plot rules as a matrix using an interactive html widget.

```{r }
plot(rules, method = "matrix", engine = "html") 
```

#### Visualization as Graph

Plot rules as a graph using an interactive html widget. *Note:* the used
javascript library does not do well with too many graph nodes, so plot
selects the top 100 rules only (with a warning).

```{r }
plot(rules, method = "graph", engine = "html")
```

#### Interactive Rule Explorer

You can specify a rule set or a dataset. To explore rules that can be
mined from iris, use: `ruleExplorer(iris)`

The rule explorer creates an interactive Shiny application that can be
used locally or deployed on a server for sharing. A deployed version of
the ruleExplorer is available
[here](https://mhahsler-apps.shinyapps.io/ruleExplorer_demo/) (using
[shinyapps.io](https://www.shinyapps.io/)).

## Exercises*

We will again use the Palmer penguin data for the exercises.

```{r }
library(palmerpenguins)
head(penguins)
```

1. Translate the penguin data into transaction data with:
  ```{r }
  trans <- transactions(penguins)
  trans
  ```
  Why does the conversion report warnings? 
  
2. What do the following first three transactions mean?
  ```{r }
  inspect(trans[1:3])
  ```

Next, use the `ruleExplorer()` function to analyze association rules
created for the transaction data set.

1. Use the default settings for the parameters. 
  Using the _Data Table_, what is the association rule with the highest lift. What does its LHS, RHS,
  support, confidence and lift mean?
  
2. Use the _Graph_ visualization. Use select by id to highlight different species and
  different islands and then hover over some of the rules. What do you see?

