# Association Analysis: Advanced Concepts

This chapter discusses a few advanced concepts of association analysis.
First, we look at how categorical and continuous attributes are converted
into items. The we look at integrating item hierarchies into the analysis.
Finally, sequence pattern mining is introduced.

### Packages Used in this Chapter {.unnumbered}

```{r setup_06}
pkgs <- c("arules", "arulesSequences", "tidyverse")

pkgs_install <- pkgs[!(pkgs %in% installed.packages()[,"Package"])]
if(length(pkgs_install)) install.packages(pkgs_install)
```

```{r setup_06-2, include=FALSE}
source("_common.R")
source("format_pkgs.R")
all_pkgs <- union(all_pkgs, pkgs)
```

The packages used for this chapter are: 

`r format_pkgs(pkgs)`

## Handling Categorical Attributes

Categorical attributes are nominal or ordinal variables. 
In R they are `factors` or `ordinal`. They are
translated into a series of binary items (one for each level constructed as ⁠variable
`name = level`⁠). Items cannot represent order and this ordered factors lose
the order information. Note that nominal variables need to be encoded as 
factors (and not characters or numbers) before converting them into transactions.

For the special case of Boolean variables (`logical`), the `TRUE` value is 
converted into an item with the name of the variable and for the `FALSE` 
values no item is created.

We will give an example in the next section.

## Handling Continuous Attributes

Continuous variables cannot directly be represented as items and need to be 
discretized first (see [Discretization] in Chapter 2). 
An item resulting from discretization might be `age>18` and 
the column contains only `TRUE` or `FALSE`. Alternatively, it can be a factor 
with levels `age<=18`, `⁠50=>age>18`⁠ and `age>50`. These will be automatically
converted into 3 items, one for each level. Discretization is described in functions 
`discretize()` and `discretizeDF()` to discretize all columns in a data.frame.

We give a short example using the iris dataset. We add an extra `logical` column
to show how Boolean attributes are converted in to items.

```{r }
data(iris)

## add a Boolean attribute
iris$Versicolor <- iris$Species == "versicolor"
head(iris)
```

The first step is to
discretize continuous attributes (marked as `<dbl>` in the table above).
We discretize the two Petal features.

```{r }
library(tidyverse)
library(arules)

iris_disc <- iris %>% 
  mutate(Petal.Length = discretize(Petal.Length, 
                          method = "frequency", 
                          breaks = 3, 
                          labels = c("short", "medium", "long")),
         Petal.Width = discretize(Petal.Width,
                          method = "frequency", 
                          breaks = 2, 
                          labels = c("narrow", "wide"))
         )
  

head(iris_disc)
```

Next, we convert the dataset into transactions.

```{r }
trans <- transactions(iris_disc)
trans
```

The conversion creates a warning because there are still two undiscretized columns
in the data. The warning indicates that the default discretization is used 
automatically.

```{r }
itemLabels(trans)
```

We see that all continuous variables are discretized and the different ranges 
create an item. For example `Petal.Width` has the two items `Petal.Width=narrow`
and `Petal.Width=wide`. The automatically discretized variables show intervals.
`Sepal.Length=[4.3,5.4)` means that this item used for flowers with 
a sepal length between 4.3 and 5.4 cm.

The species is converted into three items, one for each class. The logical 
variable `Versicolor` created only a single item that is used when 
the variable is `TRUE`.

## Handling Concept Hierarchies

Often an item hierarchy is available for transactions used for association rule mining. For example in a supermarket dataset items like "bread" and "beagle" might belong to the item group (category) "baked goods."
Transactions can store item hierarchies as additional columns in the itemInfo data.frame.

### Aggregation

To perform analysis at a group level of the item hierarchy, `aggregate()` produces a new object with items aggregated to a given group level. A group-level item is present if one or more of the items in the group are present in the original object. If rules are aggregated, and the aggregation would lead to the same aggregated group item in the lhs and in the rhs, then that group item is removed from the lhs. Rules or itemsets, which are not unique after the aggregation, are also removed. Note also that the quality measures are not applicable to the new rules and thus are removed. If these measures are required, then aggregate the transactions before mining rules.

We use the Groceries data set in this example. It contains 1 month (30 days) of real-world point-of-sale transaction data from a typical local grocery outlet. The items 
are 169 products categories.

```{r }
data("Groceries")
Groceries
```

The dataset also contains two aggregation levels.

```{r }
head(itemInfo(Groceries))
```

We aggregate to level1 stored in Groceries. All items with the same level2 label
will become a single item with that name. This reduces the number of items to
the 55 level2 categories

```{r }
Groceries_level2 <- aggregate(Groceries, by = "level2")
Groceries_level2
head(itemInfo(Groceries_level2)) ## labels are alphabetically sorted!
```
We can now compare an original transaction with the aggregated transaction.

```{r }
inspect(head(Groceries, 3))
inspect(head(Groceries_level2, 3))
```
For example, citrus fruit in the first transaction was translated to the category fruit.
Note that the order of items in a transaction is not important, so it might change
during aggregation.

It is now easy to mine rules on the aggregated data.
```{r }
rules <- apriori(Groceries_level2, support = 0.005)
rules |> head(3, by = "support") |> inspect()
```

You can add your own aggregation to an existing dataset by constructing
the and iteminfo data.frame and adding it to the transactions. See `? hierarchy` 
for details.

### Multi-level Analysis

To analyze relationships between individual items and item groups at the same time, `addAggregate()` can be used to create a new transactions object which contains both, the original items and group-level items.

```{r}
Groceries_multilevel <- addAggregate(Groceries, "level2")
Groceries_multilevel |> head(n=3) |> inspect()
```

The added group-level items are marked with an \* after the name. Now we can mine rules 
including items from multiple levels.

```{r }
rules <- apriori(Groceries_multilevel,
  parameter = list(support = 0.005))
rules
```

Mining rules with group-level items added will 
create many spurious rules of the type

```{r eval=FALSE}
⁠item A => group of item A⁠
```

with a confidence of 1. 
This will also happen if you mine itemsets. `filterAggregate()` 
can be used to filter these spurious rules or itemsets.

```{r }
rules <- filterAggregate(rules)
rules

rules |> head(n = 3, by = "lift") |> inspect()
```

Using multi-level mining can reduce the number of rules and help to analyze if
customers differentiate between products in a group.

## Sequential Patterns

The frequent sequential pattern mining algorithm cSPADE [@Zaki2000] is implemented in 
the `arules` extension package `arulesSequences`. 

Sequential pattern mining starts with sequences of events. 
Each sequence is identified by a sequence ID and 
each event is a set of items that happen together. 
The order of events is specified using
event IDs. The goal is to find subsequences of items in events that
follow each other frequently. These are called frequent sequential pattern.

We will look at a small example dataset that comes with the package
arulesSequences.

```{r}
library(arulesSequences)
data(zaki)

inspect(zaki)
```

The dataset contains four sequences (see `sequenceID`) and the event IDs
are integer numbers to provide the order events in a sequence.
In arulesSequences, this set of sequences is implemented as a regular transaction
set, where each transaction is an event. The temporal information is added
as extra columns to the transaction's `transactionInfo()` data.frame.

Mine frequent sequence patterns
using cspade is very similar to using apriori.
Here we set support so we will find patterns that occur in 50% of the 
sequences.

```{r}
fsp <- cspade(zaki, parameter = list(support = .5))
fsp |> inspect()
```
For example, pattern 17 shows that D in an event, it is often followed by 
an event by containing B and F which in turn is followed by an event
containing A.

The cspade algorithm supports many additional parameters to control gaps
and windows. Details can be found in the manual page for `cspade`.

Rules, similar to regular association rules can be generated 
from frequent sequence patterns using `ruleInduction()`.
```{r }
rules <- ruleInduction(fsp, confidence = .8)
rules |> inspect()
```

The usual measures of confidence and lift are used.