---
editor_options:
  markdown:
    wrap: 72
---

# (APPENDIX) Appendix {-} 

# Data Exploration and Visualization

The following code covers the important part of 
data exploration. For space reasons, this chapter was moved from the
printed textbook to this 
[Data Exploration Web Chapter.](https://mhahsler.github.io/Introduction_to_Data_Mining_R_Examples/DM_chapters/data_exploration_1st_edition.pdf)


### Packages Used in this Chapter {.unnumbered}

```{r setup_90}
pkgs <- c("arules", "GGally", 
          "ggcorrplot", "hexbin", "palmerpenguins", "plotly", "seriation", "tidyverse")

pkgs_install <- pkgs[!(pkgs %in% installed.packages()[,"Package"])]
if(length(pkgs_install)) install.packages(pkgs_install)
```

The packages used for this chapter are: 

`r format_pkgs(pkgs)`

```{r setup_90-2, include=FALSE}
source("_common.R")
source("format_pkgs.R")
all_pkgs <- union(all_pkgs, pkgs)
```


We will use again the iris dataset.

```{r }
library(tidyverse)
data(iris)
iris <- as_tibble(iris)
iris
```

## Exploring Data

### Basic statistics

Get summary statistics (using base R)

```{r }
summary(iris)
```

Get mean and standard deviation for sepal length.

```{r }
iris |> 
  summarize(avg_Sepal.Length = mean(Sepal.Length), 
            sd_Sepal.Length = sd(Sepal.Length))
```

Data with missing values will result in statistics of `NA`. Adding the
parameter `na.rm = TRUE` can be used in most statistics functions to
ignore missing values.

```{r }
mean(c(1, 2, NA, 3, 4, 5))
mean(c(1, 2, NA, 3, 4, 5),  na.rm = TRUE)
```

Outliers are typically the smallest or the largest values of a feature.
To make the mean more robust against outliers, we can trim 10% of
observations from each end of the distribution.

```{r }
iris |>
  summarize(
    avg_Sepal.Length = mean(Sepal.Length),
    trimmed_avg_Sepal.Length = mean(Sepal.Length, trim = .1)
  )
```

Sepal length does not have outliers, so the trimmed mean is almost
identical.

To calculate a summary for a set of features (e.g., all numeric
features), tidyverse provides `across(where(is.numeric), fun)`.

```{r }
iris |> summarize(across(where(is.numeric), mean))
iris |> summarize(across(where(is.numeric), sd))

iris |> summarize(across(where(is.numeric), 
            list(min = min, 
                 median = median, 
                 max = max)))
```

The median absolute deviation (MAD) is another measure of dispersion.

```{r }
iris |> summarize(across(where(is.numeric), mad))
```

### Grouped Operations and Calculations

We can use the nominal feature to form groups and then calculate
group-wise statistics for the continuous features. We often use
group-wise averages to see if they differ between groups.

```{r }
iris |> 
  group_by(Species) |> 
  summarize(across(Sepal.Length, mean))
iris |> 
  group_by(Species) |> 
  summarize(across(where(is.numeric), mean))
```

We see that the species Virginica has the highest average for all, but
Sepal.Width.

The statistical difference between the groups can be tested using [ANOVA
(analysis of
variance)](http://www.sthda.com/english/wiki/one-way-anova-test-in-r).

```{r }
res.aov <- aov(Sepal.Length ~ Species, data = iris)
summary(res.aov)
TukeyHSD(res.aov)
```

The summary shows that there is a significant difference for
Sepal.Length between the groups. `TukeyHDS` evaluates differences
between pairs of groups. In this case, all are significantly different.
If the data only contains two groups, the `t.test` can be used.

### Tabulate data

We can count the number of flowers for each species.

```{r }
iris |> 
  group_by(Species) |> 
  summarize(n())
```

In base R, this can be also done using `count(iris$Species)`.

For the following examples, we discretize the data using cut.

```{r }
iris_ord <- iris |> 
  mutate(across(where(is.numeric),  
    function(x) cut(x, 3, labels = c("short", "medium", "long"), 
                    ordered = TRUE)))

iris_ord
summary(iris_ord)
```

Cross tabulation is used to find out if two discrete features are
related.

```{r }
tbl <- iris_ord |> 
  select(Sepal.Length, Species) |> 
  table()
tbl
```

The table contains the number of rows that contain the combination of
values (e.g., the number of flowers with a short Sepal.Length and
species Setosa is `r tbl[1, 1]`). If a few cells have very large counts
and most others have very low counts, then there might be a
relationship. For the iris data, we see that species Setosa has mostly a
short Sepal.Length, while Versicolor and Virginica have longer sepals.

Creating a cross table with tidyverse is a little more involved and uses
pivot operations and grouping.

```{r }
iris_ord |>
  select(Species, Sepal.Length) |>
### Relationship Between Nominal and Ordinal Features
  pivot_longer(cols = Sepal.Length) |>
  group_by(Species, value) |> 
  count() |> 
  ungroup() |>
  pivot_wider(names_from = Species, values_from = n)
```

We can use a statistical test to determine if there is a significant
relationship between the two features. [Pearson's chi-squared
test](https://en.wikipedia.org/wiki/Chi-squared_test) for independence
is performed with the null hypothesis that the joint distribution of the
cell counts in a 2-dimensional contingency table is the product of the
row and column marginals. The null hypothesis h0 is independence between
rows and columns.

```{r }
tbl |> 
  chisq.test()
```

The small p-value indicates that the null hypothesis of independence
needs to be rejected. For small counts (cells with counts \<5),
[Fisher's exact
test](https://en.wikipedia.org/wiki/Fisher%27s_exact_test) is better.

```{r }
fisher.test(tbl)
```

### Percentiles (Quantiles)

[Quantiles](https://en.wikipedia.org/wiki/Quantile) are cutting points
dividing the range of a probability distribution into continuous
intervals with equal probability. For example, the median is the
empirical 50% quantile dividing the observations into 50% of the
observations being smaller than the median and the other 50% being
larger than the median.

By default quartiles are calculated. 25% is typically called Q1, 50% is
called Q2 or the median and 75% is called Q3.

```{r }
iris |> 
  pull(Petal.Length) |> 
  quantile()
```

The interquartile range is a measure for variability that is robust
against outliers. It is defined the length Q3 - Q2 which covers the 50%
of the data in the middle.

```{r }
iris |> 
  summarize(IQR = 
  quantile(Petal.Length, probs = 0.75) - 
    quantile(Petal.Length, probs = 0.25))
```

### Correlation

#### Pearson Correlation

Correlation can be used for ratio/interval scaled features. We typically
think of the [Pearson correlation
coefficient](https://en.wikipedia.org/wiki/Pearson_correlation_coefficient)
between features (columns).

```{r }
cc <- iris |> 
  select(-Species) |> 
  cor()
cc
```

`cor` calculates a correlation matrix with pairwise correlations between
features. Correlation matrices are symmetric, but different to
distances, the whole matrix is stored.

The correlation between Petal.Length and Petal.Width can be visualized
using a scatter plot.

```{r }
ggplot(iris, aes(Petal.Length, Petal.Width)) + 
  geom_point() +
  geom_smooth(method = "lm")
```

`geom_smooth` adds a regression line by fitting a linear model (`lm`).
Most points are close to this line indicating strong linear dependence
(i.e., high correlation).

We can calculate individual correlations by specifying two vectors.

```{r }
with(iris, cor(Petal.Length, Petal.Width))
```

*Note:* `with` lets you use columns using just their names and
`with(iris, cor(Petal.Length, Petal.Width))` is the same as
`cor(iris$Petal.Length, iris$Petal.Width)`.

Finally, we can test if a correlation is significantly different from
zero.

```{r }
with(iris, cor.test(Petal.Length, Petal.Width))
```

A small p-value (less than 0.05) indicates that the observed correlation
is significantly different from zero. This can also be seen by the fact
that the 95% confidence interval does not span zero.

Sepal.Length and Sepal.Width show little correlation:

```{r } 
ggplot(iris, aes(Sepal.Length, Sepal.Width)) + 
  geom_point() +   
  geom_smooth(method = "lm") 

with(iris, cor(Sepal.Length, Sepal.Width)) 
with(iris, cor.test(Sepal.Length, Sepal.Width))
```

#### Rank Correlation

Rank correlation is used for ordinal features or if the correlation is
not linear. To show this, we first convert the continuous features in
the Iris dataset into ordered factors (ordinal) with three levels using
the function `cut`.

```{r }
iris_ord <- iris |> 
  mutate(across(where(is.numeric), 
    function(x) cut(x, 3, 
                    labels = c("short", "medium", "long"), 
                    ordered = TRUE)))

iris_ord
summary(iris_ord)
iris_ord |> 
  pull(Sepal.Length)
```

Two measures for rank correlation are Kendall's Tau and Spearman's Rho.

[Kendall's Tau Rank Correlation
Coefficient](https://en.wikipedia.org/wiki/Kendall_rank_correlation_coefficient)
measures the agreement between two rankings (i.e., ordinal features).

```{r }
iris_ord |> 
  select(-Species) |> 
  sapply(xtfrm) |> 
  cor(method = "kendall")
```

**Note:** We have to use `xtfrm` to transform the ordered factors into
ranks, i.e., numbers representing the order.

[Spearman's
Rho](https://en.wikipedia.org/wiki/Spearman%27s_rank_correlation_coefficient)
is equal to the Pearson correlation between the rank values of those two
features.

```{r }
iris_ord |> 
  select(-Species) |> 
  sapply(xtfrm) |> 
  cor(method = "spearman")
```

Spearman's Rho is much faster to compute on large datasets then
Kendall's Tau.

Comparing the rank correlation results with the Pearson correlation on
the original data shows that they are very similar. This indicates that
discretizing data does not result in the loss of too much information.

```{r }
iris |> 
  select(-Species) |> 
  cor()
```

### Density 

[Density estimation](https://en.wikipedia.org/wiki/Density_estimation)
estimate the probability density function
(distribution) of a continuous variable from observed data.

Just plotting the data using points is not very helpful for a single
feature.

```{r }
ggplot(iris, aes(x = Petal.Length, y = 0)) + geom_point()
```

#### Histograms

A [histograms](https://en.wikipedia.org/wiki/Histogram) shows more about
the distribution by counting how many values fall within a bin and
visualizing the counts as a bar chart. We use `geom_rug` to place marks
for the original data points at the bottom of the histogram.

```{r }
ggplot(iris, aes(x = Petal.Length)) +
  geom_histogram() +
  geom_rug(alpha = 1/2)
```

Two-dimensional distributions can be visualized using 2-d binning or
hexagonal bins.

```{r }
ggplot(iris, aes(Sepal.Length, Sepal.Width)) +
  geom_bin2d(bins = 10) +
  geom_jitter(color = "red")

ggplot(iris, aes(Sepal.Length, Sepal.Width)) +
  geom_hex(bins = 10) +
  geom_jitter(color = "red")
```

#### Kernel Density Estimate (KDE)

[Kernel density
estimation](https://en.wikipedia.org/wiki/Kernel_density_estimation) is
used to estimate the probability density function (distribution) of a
feature. It works by replacing each value with a kernel function (often
a Gaussian) and then adding them up. The result is an estimated
probability density function that looks like a smoothed version of the
histogram. The bandwidth (`bw`) of the kernel controls the amount of
smoothing.

```{r }
ggplot(iris, aes(Petal.Length)) +
  geom_density(bw = .2) +
  geom_rug(alpha = 1/2)
```

Kernel density estimates can also be done in two dimensions.

```{r }
ggplot(iris, aes(Sepal.Length, Sepal.Width)) +
  geom_density_2d_filled() +
  geom_jitter()
```

## Visualization

Visualization uses several components to convey information:

* Symbols: Circles, dots, lines, bars, ...
* Position: Axes (labels and units) and the origin (0/0) are important.
* Length, Size and Area: Should only be used if it faithfully represents
    information.
* Color: A color should not overpower another. Limit to 3 or 4 colors.
* Angle: Human eyes are not good at comparing angles!

All components of the plot need to convey information. E.g., do not use
color just to make it more colorful. 
All good visualizations show the important patterns clearly.

For space reasons, the chapter on data exploration and visualization 
was moved from the printed textbook and can now be found in the 
[Data Exploration Web Chapter.](https://www-users.cse.umn.edu/~kumar001/dmbook/data_exploration_1st_edition.pdf)

For visualization in R, we will use mainly `ggplot2`. The *gg* in `ggplot2`
stands for **Grammar of Graphics** introduced by @Wilkinson2005. The
main idea is that every graph is built from the same basic components:

-   the data,
-   a coordinate system, and
-   visual marks representing the data (geoms).

In `ggplot2`, the components are combined using the `+` operator.

```{r eval=FALSE}
ggplot(data, mapping = aes(x = ..., y = ..., color = ...)) +
  geom_point()
```

Since we typically use a Cartesian coordinate system, `ggplot` uses that
by default. Each `geom_` function uses a `stat_` function to calculate
what is visualizes. For example, `geom_bar` uses `stat_count` to create
a bar chart by counting how often each value appears in the data (see
`? geom_bar`). `geom_point` just uses the stat `"identity"` to display
the points using the coordinates as they are. 

Additional components like the main title, axis labels, and different scales can
be also added. A great introduction can
be found in the [Section on Data
Visualization](https://r4ds.hadley.nz/data-visualize) [@Wickham2023],
and very useful is [RStudio's Data Visualization Cheatsheet](https://rstudio.github.io/cheatsheets/html/data-visualization.html).

Next, we go through the basic visualizations used when working with data.

### Histogram

Histograms show the distribution of a single continuous feature.

```{r }
ggplot(iris, aes(Petal.Width)) + geom_histogram(bins = 20)
```

If the data contains groups, then the group information can be easily 
added as an aesthetic to the histogram.

```{r }
ggplot(iris, aes(Petal.Width)) + 
         geom_histogram(bins = 20, aes(fill = Species))
```

The bars appear stacked with a green blocks on top pf blue blocks.
To display the three distributions behind each other, we change `position`
for placement and make the bars slightly translucent using `alpha`.
  

```{r }
ggplot(iris, aes(Petal.Width)) + 
         geom_histogram(bins = 20, aes(fill = Species), alpha = .5, position = 'identity')
```

### Boxplot

Boxplots are used to compare the distribution of a feature between
different groups. The horizontal line in the middle of the boxes are the
group-wise medians, the boxes span the interquartile range. The whiskers
(vertical lines) span typically 1.4 times the interquartile range.
Points that fall outside that range are typically outliers shown as
dots.

```{r }
ggplot(iris, aes(Species, Sepal.Length)) + 
  geom_boxplot()
```

The group-wise medians can also be calculated directly.

```{r }
iris |> group_by(Species) |> 
  summarize(across(where(is.numeric), median))
```

To compare the distribution of the four features using a ggplot boxplot,
we first have to transform the data into long format (i.e., all feature
values are combined into a single column).

```{r }
library(tidyr)
iris_long <- iris |> 
  mutate(id = row_number()) |> 
  pivot_longer(1:4)

ggplot(iris_long, aes(name, value)) + 
  geom_boxplot() +
  labs(y = "Original value")
```

This visualization is only useful if all features have roughly the same
range. The data can be scaled first to compare the distributions.

```{r }
library(tidyr)
iris_long_scaled <- iris |> 
  scale_numeric() |> 
  mutate(id = row_number()) |> pivot_longer(1:4)

ggplot(iris_long_scaled, aes(name, value)) + 
  geom_boxplot() +
  labs(y = "Scaled value")
```

### Scatter plot

Scatter plots show the relationship between two continuous features.

```{r }
ggplot(iris, aes(x = Petal.Length, y = Sepal.Length)) + 
  geom_point(aes(color = Species))
```

We can add a regression using `geom_smooth` with the 
linear model method to show that there is a 
linear relationship between the two variables. A confidence interval 
for the regression is also shown. 
This can be suppressed using `se = FALSE`.

```{r }
ggplot(iris, aes(x = Petal.Length, y = Sepal.Length)) + 
  geom_point(aes(color = Species)) +  
  geom_smooth(method = "lm")
```

We can also perform group-wise linear regression by adding the color
aesthetic also to `geom_smooth`.

```{r }
ggplot(iris, aes(x = Petal.Length, y = Sepal.Length)) + 
  geom_point(aes(color = Species)) +  
  geom_smooth(method = "lm", aes(color = Species))
```

The same can be achieved by using the color aesthetic in the `qqplot` call,
then it applies to all geoms.

### Scatter Plot Matrix

A scatter plot matrix show the relationship between all pairs of features
by arranging panels in a matrix. First, lets look at a regular
R-base plot.

```{r }
pairs(iris, col = iris$Species)
```

The package `GGally` provides a way more sophisticated visualization.

```{r }
library("GGally")
ggpairs(iris,  aes(color = Species), progress = FALSE)
```

Additional plots
(histograms, density estimates and box plots) and correlation
coefficients are shown in different panels. See the [Data Quality]
section for a description of how to interpret the different panels.

### Matrix Visualization

Matrix visualization shows the values in the matrix using a color scale.

```{r }
iris_matrix <- iris |> select(-Species) |> as.matrix()
```

We need the long format for tidyverse.

```{r }
iris_long <- as_tibble(iris_matrix) |> 
  mutate(id = row_number()) |> 
  pivot_longer(1:4)

head(iris_long)

ggplot(iris_long, aes(x = name, y = id)) + 
  geom_tile(aes(fill = value))
```

Smaller values are darker. Package `seriation` provides a simpler
plotting function.

```{r }
library(seriation)
ggpimage(iris_matrix, prop = FALSE)
```

We can scale the features to z-scores to make them better comparable.

```{r }
iris_scaled <- scale(iris_matrix)
ggpimage(iris_scaled, prop = FALSE)
```

This reveals red and blue blocks. Each row is a flower and the flowers
in the Iris dataset are sorted by species. The blue blocks for the top
50 flowers show that these flowers are smaller than average for all but
Sepal.Width and the red blocks show that the bottom 50 flowers are
larger for most features.

Often, reordering data matrices help with visualization. A reordering
technique is called seriation. Ir reorders rows and columns to place
more similar points closer together.

```{r }
ggpimage(iris_scaled, order = seriate(iris_scaled), prop = FALSE)
```

We see that the rows (flowers) are organized from very blue to very red
and the features are reordered to move Sepal.Width all the way to the
right because it is very different from the other features.

### Correlation Matrix

A correlation matrix contains the correlation between features.

```{r }
cm1 <- iris |> 
  select(-Species) |> 
  as.matrix() |> 
  cor()
cm1
```

Package `ggcorrplot` provides a visualization for correlation matrices.

```{r }
library(ggcorrplot)
ggcorrplot(cm1)
```

Package `seriation` provides a reordered version for this plot using a
heatmap.

```{r }
gghmap(cm1, prop = TRUE)
```

Correlations can also be calculates between objects by transposing the
data matrix.

```{r }
cm2 <- iris |> 
  select(-Species) |> 
  as.matrix() |> 
  t() |> 
  cor()

ggcorrplot(cm2)
```

Object-to-object correlations can be used as a measure of similarity.
The dark red blocks indicate different species.

### Parallel Coordinates Plot

Parallel coordinate plots can visualize several features in a single
plot. Lines connect the values for each object (flower).

```{r }
library(GGally)
ggparcoord(iris, columns = 1:4, groupColumn = 5)
```

The plot can be improved by reordering the variables to place correlated
features next to each other.

```{r }
o <- seriate(as.dist(1-cor(iris[,1:4])), method = "BBURCG")
get_order(o)
ggparcoord(iris, 
           columns = as.integer(get_order(o)), 
           groupColumn = 5)
```

### Star Plot

Star plots are a type of [radar chart](https://en.wikipedia.org/wiki/Radar_chart)
to visualize three or more quantitative variables represented on axes starting from 
the plot's origin. R-base offers a simple star plot. We plot the 
first 5 flowers of each species.


```{r }
flowers_5 <- iris[c(1:5, 51:55, 101:105), ]
flowers_5

stars(flowers_5[, 1:4], ncol = 5)
```

### More Visualizations

A well organized collection of visualizations with code can be found at
[The R Graph Gallery](https://www.r-graph-gallery.com/).

## Exercises*

The R package **palmerpenguins** contains measurements for penguin of different
species from the Palmer Archipelago, Antarctica. Install the package.
It provides a CSV file which can be read in the following way:

```{r }
library("palmerpenguins")
penguins <- read_csv(path_to_file("penguins.csv"))
head(penguins)
```

Create in RStudio a new R Markdown document.
Apply the code in the sections of this chapter to the data set to answer the
following questions. 

1. Group the penguins by species, island or sex. What can you find out?
2. Create histograms and boxplots for each continuous variable. Interpret the 
   distributions.
3. Create scatterplots and a scatterplot matrix. Can you identify correlations?
4. Create a reordered correlation matrix visualization. What does the visualizations show?

Make sure your markdown document contains now a well formatted report.
Use the `Knit` button to create a HTML document.



