---
editor_options: 
  markdown: 
    wrap: 72
---

```{r ch91_clean_R, include=FALSE}
source("clean_R.R")
source("_common.R")
source("format_pkgs.R")
```

# Regression {#appendix_regression}

Regression is an important statistical method that is covered 
the freely available [Appendix D](https://mhahsler.github.io/Introduction_to_Data_Mining_R_Examples/DM_chapters/appendices_2ed.pdf)
of the data mining textbook.

In this extra chapter, we introduce the regression problem
and multiple linear regression. In addition,
alternative models like regression trees and regularized regression
are discussed.

## Learning Objectives

After this chapter, you should be able to:

* fit and interpret a multiple linear regression model;
* compare nested models and assess interactions and categorical predictors;
* generate predictions and explain their uncertainty and limitations; and
* compare linear regression with tree-based and regularized alternatives.

## Prerequisites

Read [Classification: Basic Concepts] for the train/test distinction. Basic
knowledge of descriptive statistics, correlation, and algebra is assumed.

### Packages Used in this Chapter {.unnumbered}

```{r ch91_packages_used_in_this_ch_001, results='hide', message=FALSE}
pkgs <- c('lars', 'rpart', 'rpart.plot', 'nnet', 'tidyverse')
  
pak::pkg_install(pkgs)
```

```{r ch91_packages_used_in_this_ch_002, include=FALSE}
all_pkgs <- union(all_pkgs, pkgs)
```

The packages used for this chapter are: 

`r format_pkgs(pkgs)`


## Introduction

Recall that classification predicts one of a small set of discrete (i.e., nominal) labels (e.g., yes or no, 
small, medium or large). 
Regression is also a supervised learning problem, but its goal is to predict
a numeric response from a set of predictors.
We start with the popular linear regression and will later discuss alternative 
regression methods.

Linear regression models a response $y$ as a linear function of predictors
$X_1, X_2, ..., X_p$ (also called explanatory variables or covariates). With
multiple predictors, the model is called multiple linear regression.
The linear regression model is:
$$y_i = \beta_0 + \beta_1 x_{i1} + \dots + \beta_p x_{ip} + \epsilon_i, \qquad
\hat{y}_i = \beta_0 + \beta_1 x_{i1} + \dots + \beta_p x_{ip},$$

where $\beta_0$ is the intercept, $\boldsymbol{\beta}$ is a $p+1$-dimensional parameter
vector learned from the data, and $\epsilon_i$ is the random error term. The
observed residual is $e_i = y_i - \hat{y}_i$. In vector notation, the model is
$$\mathbf{y} = \mathbf{X} \boldsymbol{\beta} + \boldsymbol{\epsilon},$$
and its fitted values are
$$\hat{\mathbf{y}} = \mathbf{X} \boldsymbol{\beta}.$$
Here, $\mathbf{X}$ is the design matrix containing the predictors.

The error that is often minimized in regression problems is the squared error defined
as:

$$SE= \sum_i (y_i - f(\mathbf{x}_i))^2$$

The parameter vector is estimated by minimizing the sum of squared residuals in
the training data.

Linear regression makes several assumptions that should be checked:

* _Linearity:_ The conditional mean of the response is linear in the predictors.
* _Homoscedasticity:_ The error variance is constant across fitted values.
* _Independence:_ Errors are independent across observations, as appropriate
  for the sampling design.
* _No perfect multicollinearity:_ No predictor is an exact linear combination
  of other predictors. Strong correlations can make coefficient estimates
  unstable, though they do not always invalidate predictions.
* _Normal errors:_ A normal error assumption is used for some small-sample
  confidence intervals and tests; it is not required for least-squares
  coefficient estimation itself.

## A First Linear Regression Model

We will use the Iris dataset and try to predict 
the `Petal.Width` using the other variables. We first
load and shuffle data since the flowers in the dataset are in order by species.

```{r ch91_a_first_linear_regressio_003}
data(iris)
set.seed(2000) # make the sampling reproducible

x <- iris[sample(1:nrow(iris)),]
plot(x, col=x$Species)
```

The Iris data is very clean, so we make the data a little messy by adding a random error
to each variable and introduce a useless, completely random feature.

```{r ch91_a_first_linear_regressio_004}
x[,1] <- x[,1] + rnorm(nrow(x))
x[,2] <- x[,2] + rnorm(nrow(x))
x[,3] <- x[,3] + rnorm(nrow(x))
x <- cbind(x[,-5], 
           useless = mean(x[,1]) + rnorm(nrow(x)), 
           Species = x[,5])

plot(x, col=x$Species)
summary(x)
head(x)
```

We split the data into training and test data. Since the data is shuffled, we
effectively perform holdout sampling. This is often not done in 
statistical applications, but we use the machine learning approach here. 

```{r ch91_a_first_linear_regressio_005}
train <- x[1:100,]
test <- x[101:150,]
```

Linear regression is done in R using the `lm()` (linear model) function which is
part of the R core package `stats`.
Like other modeling functions in R, `lm()` uses a formula interface. Here we create a formula
to predict `Petal.Width` by all other variables other than `Species`.

```{r ch91_a_first_linear_regressio_006}
model1 <- lm(Petal.Width ~ Sepal.Length
            + Sepal.Width + Petal.Length + useless,
            data = train)
model1
```

The result is a model with the fitted $\beta$ coefficients. More information can 
be displayed using the summary function. 

```{r ch91_a_first_linear_regressio_007}
summary(model1)
```

The summary reports coefficient estimates and tests of whether each coefficient
is zero, conditional on the other predictors and the model assumptions. Here,
`Petal.Length` has a small p-value and the coefficient for `useless` is close
to zero. Such tests do not establish that a predictor is causally important,
and p-values can be optimistic after model selection.

R-squared (the coefficient of determination) is the proportion of the response
variation explained by the fitted model on these data. Adjusted R-squared
accounts for the number of predictors, but neither statistic is a substitute
for evaluating predictions on held-out data. There is no universal cutoff for
what counts as a good R-squared.

Plotting the model produces diagnostic plots (see `plot.lm()`). For example,
to check that the error term has a mean of 0 and is homoscedastic, 
the residual vs. predicted value scatter plot 
should have a red line that stays close to 0 and
not look like a funnel where they are increasing when the fitted values increase.
To check if the residuals are approximately normally distributed,
the Q-Q plot should be close to the straight diagonal line.

```{r ch91_a_first_linear_regressio_008}
plot(model1, which = 1:2)
```

In this case, the two plots look fine.

## Comparing Nested Models
Here we compare nested linear models. Models are nested when one model's terms
are a subset of the other's, with the same response and observations. We first
remove the random `useless` predictor.

```{r ch91_comparing_nested_models_009}
model2 <- lm(Petal.Width ~ Sepal.Length + 
               Sepal.Width + Petal.Length,
             data = train)
summary(model2)
```

The next example removes the intercept by adding `-1` to the formula. This
changes the model assumptions and should only be done when a zero response at
zero predictors is scientifically appropriate.

```{r ch91_comparing_nested_models_010}
model3 <- lm(Petal.Width ~ Sepal.Length + 
               Sepal.Width + Petal.Length - 1,
             data = train)
summary(model3)
```

Here is a very simple model.

```{r ch91_comparing_nested_models_011}
model4 <- lm(Petal.Width ~ Petal.Length -1,
             data = train)
summary(model4)
```

For nested linear models, `anova()` performs a partial F-test of whether the
additional terms reduce residual error enough to support the larger model.
Compare only nested models fitted to the same observations. Here, models 1 and
2 form one nested pair, and models 3 and 4 form another. Models 2 and 3 differ
in their intercept structure and are not compared with the same test.

```{r ch91_comparing_nested_models_012}
anova(model1, model2)
anova(model3, model4)
```


Interpret each test in light of its assumptions and the model purpose; a
non-significant result does not prove that the smaller model is correct. See
`?anova.lm` for details.

## Stepwise Variable Selection
Stepwise variable selection performs backward (or forward) model selection for linear models
and uses the smallest AIC ([Akaike information criterion](https://en.wikipedia.org/wiki/Akaike_information_criterion)) to decide 
what variable to remove and when to stop.

```{r ch91_stepwise_variable_select_013}
s1 <- step(lm(Petal.Width ~ . -Species, data = train))

summary(s1)
```

Each table represents one step and shows the AIC when each variable is removed and 
the one with the smallest AIC (in the first table `useless`) is removed. It stops 
with a small model that only uses `Petal.Length`.


## Modeling with Interaction Terms
Linear regression models the effect of each predictor separately
using a $\beta$ coefficient.
What if two variables are only important together? 
This is called an interaction between predictors and is modeled using 
interaction terms.
In R's `formula()` we can use `:` and `*` 
to specify interactions. For linear regression, an interaction 
means that a new predictor is created by multiplying the two original 
predictors.

We can create a model with an interaction terms between `Sepal.Length`,  `Sepal.Width` and 
`Petal.Length`.
```{r ch91_modeling_with_interactio_014}
model5 <- lm(Petal.Width ~ Sepal.Length *
               Sepal.Width * Petal.Length,
             data = train)

summary(model5)

anova(model2, model5)
```
The models are nested, so the ANOVA comparison assesses whether the interaction
terms improve fit over the main-effects model. Check predictive performance on
held-out data before choosing a model for prediction.

## Prediction

We make predictions for the held-out test set.

```{r ch91_prediction_015}
test[1:5,]

test[1:5,]$Petal.Width

predict(model4, test[1:5,])
```

The most used error measure for regression is the 
[RMSE](https://en.wikipedia.org/wiki/Root_mean_square_deviation)
root-mean-square error.

```{r ch91_prediction_016}
RMSE <- function(predicted, true) mean((predicted-true)^2)^.5

RMSE(predict(model4, test), test$Petal.Width)
```

We can also visualize the quality of the prediction using a simple scatter plot of
predicted vs. actual values.

```{r ch91_prediction_017}
plot(test[,"Petal.Width"], predict(model4, test),
  xlim=c(0,3), ylim=c(0,3), 
  xlab = "actual", ylab = "predicted",
  main = "Petal.Width")
abline(0,1, col="red")
cor(test[,"Petal.Width"], predict(model4, test))
```
Perfect predictions would be on the red line, the farther they are away, 
the larger the error.


## Using Nominal Variables

[Dummy variables](https://en.wikipedia.org/wiki/Dummy_variable_(statistics)) also 
called one-hot encoding in machine learning is used for factors
(i.e., levels are translated into individual 0-1 variable).
The first level of factors is automatically used as the reference and
the other levels are presented as 0-1 dummy variables called contrasts.

```{r ch91_using_nominal_variables_018}
levels(train$Species)
```

`model.matrix()` is used internally to create dummy variables when the design matrix 
for the regression is created..

```{r ch91_using_nominal_variables_019}
head(model.matrix(Petal.Width ~ ., data=train))
```

Note that there is no dummy variable for species Setosa, because it is
used as the reference (when the other two dummy variables are 0). 
It is often useful to set the reference level.
A simple way is to use the
function `relevel()` to change which factor is listed first.

Let us perform model selection using AIC on the training data and then evaluate 
the final model on the held out test set to estimate the generalization error. 

```{r ch91_using_nominal_variables_020}
model6 <- step(lm(Petal.Width ~ ., data=train))
model6

summary(model6)
```

```{r ch91_using_nominal_variables_021}
RMSE(predict(model6, test), test$Petal.Width)

plot(test[,"Petal.Width"], predict(model6, test),
  xlim=c(0,3), ylim=c(0,3), 
  xlab = "actual", ylab = "predicted",
  main = "Petal.Width")
abline(0,1, col="red")

cor(test[,"Petal.Width"], predict(model6, test))
```
We see that the `Species` variable provides information to improve the regression model.


## Alternative Regression Models
### Regression Trees

Many models used for classification can also perform regression.
For example CART implemented in rpart performs regression 
by estimating a value for each leaf note.
Regression is always performed by `rpart()` when the response variable 
is not a `factor()`.

```{r ch91_regression_trees_022}
library(rpart)
library(rpart.plot)

model7 <- rpart(Petal.Width ~ ., data = train,
  control = rpart.control(cp = 0.01))
model7

rpart.plot(model7)
```
The regression tree shows the predicted value in as the top number in the node.
Also, remember that tree-based models automatically variable selection
by choosing the splits.

Let's evaluate the regression tree by calculating the RMSE.

```{r ch91_regression_trees_023}
pred <- predict(model7, test)
RMSE(pred, test$Petal.Width)
```

And visualize the quality. 

```{r ch91_regression_trees_024}
plot(test[,"Petal.Width"], pred,
  xlim = c(0,3), ylim = c(0,3), 
  xlab = "actual", ylab = "predicted",
  main = "Petal.Width")
abline(0,1, col = "red")

cor(test[,"Petal.Width"], pred)
```

The plot and the correlation coefficient indicate that the model is very good.
In the plot we see an important property of this method which is that it predicts
exactly the same value when the data falls in the same leaf node.


### Regularized Regression
[LASSO](https://en.wikipedia.org/wiki/Lasso_(statistics)) (least absolute shrinkage and selection operator)
uses L1 regularization to reduce to perform automatic variable selection.
The regularization adds a penalty for the L1 norm of the parameter vector $\boldsymbol{\beta}$ (i.e., the summ of all
parameter values)
to the optimization. Increasing the weight of this penalty term (the weight is called $\lambda$) 
results in more and more weights being pushed to 0 which effectively reduces the number of 
parameters used in the regression.
An implementation called 
the [elastic net](https://en.wikipedia.org/wiki/Elastic_net_regularization) 
is available as the function `lars()` in package `lars`.

```{r ch91_regularized_regression_025}
library(lars)
```

We create a design matrix (with dummy variables and interaction terms).
`lm()` did this automatically for us, but for this `lars()` implementation
we have to do it manually.

```{r ch91_regularized_regression_026}
x <- model.matrix(~ . + Sepal.Length*Sepal.Width*Petal.Length ,
  data = train[, -4])
head(x)
y <- train[, 4]
```

```{r ch91_regularized_regression_027}
model_lars <- lars(x, y)

plot(model_lars)
model_lars
```



The fitted model's plot shows how variables are added (from left to right to the model).
The text output shoes that `Petal.Length` is the most important variable added to the model in step 1.
Then `Speciesvirginica` is added and so on. 
This creates a sequence of nested models where one variable is added at a time.
To select the best model [Mallows's Cp statistic](https://en.wikipedia.org/wiki/Mallows's_Cp)
can be used.

```{r ch91_regularized_regression_028}
plot(model_lars, plottype = "Cp")

best <- which.min(model_lars$Cp)

coef(model_lars, s = best)
```
The variables that are not selected have a $\beta$ coefficient of 0.

To make predictions with this model, we first have to convert the test data 
into a design matrix with the dummy variables and interaction terms.  

```{r ch91_regularized_regression_029}
x_test <- model.matrix(~ . + Sepal.Length*Sepal.Width*Petal.Length,
  data = test[, -4])
head(x_test)
```
Now we can compute the predictions.

```{r ch91_regularized_regression_030}
predict(model_lars, x_test[1:5,], s = best)
```

The prediction is the `fit` element. We can calculate the RMSE.

```{r ch91_regularized_regression_031}
pred <- predict(model_lars, x_test, s = best)$fit
RMSE(pred, test$Petal.Width)
```

And visualize the prediction.

```{r ch91_regularized_regression_032}
plot(test[,"Petal.Width"], 
     pred,
     xlim=c(0,3), ylim=c(0,3), 
     xlab = "actual", ylab = "predicted",
     main = "Petal.Width")
abline(0,1, col = "red")

cor(test[,"Petal.Width"],
    pred)
```
The model shows good predictive power on the test set.


### ANNs

Regression can be performed using [artificial neural networks](https://en.wikipedia.org/wiki/Neural_network_(machine_learning)) 
with typically a linear final layer. 
We will create a network using a single hidden layer 
with 3 neurons (a manually tuned hyper parameter) and a linear output layer (set via `linout`).

```{r ch91_anns_033}
library(nnet)

model_nnet <- nnet(Petal.Width ~ ., data = train, size = 3, linout = TRUE)
model_nnet
```

```{r ch91_anns_034}
pred <- predict(model_nnet, test)
RMSE(pred, test$Petal.Width)
```

And visualize the quality. 

```{r ch91_anns_035}
plot(test[,"Petal.Width"], pred,
  xlim = c(0,3), ylim = c(0,3), 
  xlab = "actual", ylab = "predicted",
  main = "Petal.Width")
abline(0,1, col = "red")

cor(test[,"Petal.Width"], pred)
```
_Note:_ It is often necessary to scale the inputs to the ANN so it can learn effectively. 
It is very popular to scale all inputs 
to the ranges $[0,1]$ or $[-1,1]$. The ranges of the Iris dataset are fine, so
there was no need for scaling.


### Other Types of Regression

* Robust regression: robust against violation of assumptions like heteroscedasticity and outliers 
  (`robustbase::roblm()` and `robustbase::robglm`)
* Generalized linear models (`glm()`). An example is logistic regression discussed in the next chapter.
* Nonlinear least squares (`nlm()`).

## Exercises

We will again use the Palmer penguin data for the exercises.

```{r ch91_exercises_036}
library(palmerpenguins)
head(penguins)
```

Create an R markdown document that performs the following:

1. Create a linear regression model to predict the weight 
   of a penguin (`body_mass_g`). 
2. How high is the R-squared. What does it mean.
3. What variables are significant, what are not?
4. Use stepwise variable selection to remove unnecessary variables. 
5. Predict the weight for the following new penguin:
   ```{r ch91_exercises_037 }
   library("tibble")
   
   new_penguin <- tibble(
     species = factor("Adelie", 
       levels = c("Adelie", "Chinstrap", "Gentoo")),
     island = factor("Dream", 
       levels = c("Biscoe", "Dream", "Torgersen")),
    bill_length_mm = 39.8, 
    bill_depth_mm = 19.1, 
    flipper_length_mm = 184, 
    body_mass_g = NA, 
    sex = factor("male", levels = c("female", "male")), 
    year = 2007
   ) 
   new_penguin
   ```
6. Create a regression tree. Look at the tree and explain what it does. 
  Then use the regression tree to predict the weight for the above penguin. 

**A good answer should** explain the response and predictors, interpret
coefficients with their units and reference levels, and describe what
R-squared and the significance results do and do not establish. Explain the
stepwise selection process, report the prediction for the specified penguin,
and describe the tree's splits and prediction. Where possible, assess
prediction quality on data not used to fit or select the model.

## Chapter Summary

Regression predicts a numeric response from one or more explanatory variables. The chapter builds a linear model, interprets coefficients, compares nested models, and demonstrates stepwise variable selection and interaction terms. It shows how to generate predictions, work with nominal predictors, and assess models on data not used for fitting or selection. When ordinary linear regression is not suitable, regression trees, regularized regression, and artificial neural networks offer alternatives; other regression approaches are also surveyed. Model assumptions, complexity, and evaluation data should guide comparisons.

## Key Terms

* **Response:** the numeric outcome a regression model predicts.
* **Coefficient:** the estimated change in response associated with a predictor,
  holding other model terms fixed.
* **Interaction:** a model term allowing the effect of one predictor to depend
  on another predictor.
* **Residual:** the difference between an observed response and its fitted
  value.
* **Regularization:** a method that constrains model coefficients to reduce
  overfitting.
