---
editor_options: 
  markdown: 
    wrap: 72
---

# Logistic Regression {#appendix_logistic_regression}

This chapter introduces the popular classification method 
logistic regression more in detail. Logistic regression is introduced
as an alternative classification method in Chapter 4 of Introduction to Data Mining.

```{r setup_logreg}
pkgs <- c("glmnet", "caret")

pkgs_install <- pkgs[!(pkgs %in% installed.packages()[,"Package"])]
if(length(pkgs_install)) install.packages(pkgs_install)
```

The packages used for this chapter are: 

`r format_pkgs(pkgs)`

```{r setup_logreg-2, include=FALSE}
source("_common.R")
source("format_pkgs.R")
all_pkgs <- union(all_pkgs, pkgs)
```

## Introduction

Logistic regression contains the word regression, but it is actually a 
statistical classification
model to predict the probability $p$ of a binary outcome given a set of features.
It is a very powerful classification model that can be fit very quickly. It is one of the 
first classification models you should try on new data.

Logistic regression is a [generalized linear model](https://en.wikipedia.org/wiki/Generalized_linear_model) 
with the logit as the link function and a binomial error distribution. 
It can 
be thought of as a linear regression with the
log odds ratio (logit) 
of the binary outcome as the dependent variable:

$$logit(p) = ln\left(\frac{p}{1-p}\right) = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + ...$$

The logit function links the probability p to the linear regression by converting 
a number in the probability range $[0,1]$ to the range $[-\infty,+\infty]$.

```{r}
logit  <- function(p) log(p/(1-p))
p <- seq(0, 1, length.out = 100)
plot(logit(p), p, type = "l")
abline(h = 0.5, lty = 2)
abline(v = 0, lty = 2)
```

The figure above shows actually the inverse of the logit function.
The inverse of the logit function is called [logistic (or sigmoid) function](https://en.wikipedia.org/wiki/Logistic_function) ($\sigma(\cdot)$) which is 
often used in ML, and especially for artificial neural networks, 
to squash the set of real numbers to the $[0,1]$ interval.
Using the inverse function, we see that the probability of the outcome $p$ 
is modeled by the logistic function of the linear regression:

$$ p = \frac{e^{\beta_0 + \beta_1 x_1 + \beta_2 x_2 + ...}}{1 +  e^{\beta_0 + \beta_1 x_1 + \beta_2 x_2 + ...}} = \frac{1}{1+e^{-(\beta_0 + \beta_1 x_1 + \beta_2 x_2 + ...)}} = \sigma(\beta_0 + \beta_1 x_1 + \beta_2 x_2 + ...)$$

After the $\boldsymbol{\beta} = (\beta_0, \beta_1,...)$ parameter vector is fitted using training data by
minimizing the log loss (i.e., cross-entropy loss), 
the equation above can be used to predict
the probability $p$ given a new data point $\mathbf{x} = (x_1, x_2, ...)$. If the predicted $p > .5$
then we predict that the event happens, otherwise we predict that it does not happen.

The outcome itself is binary and therefore has a Bernoulli distribution. Since we have 
multiple examples in our data we draw several times from this distribution resulting
in a Binomial distribution for the number of successful events drawn. Logistic 
regression therefore uses a logit link function to link the probability of the event 
to the linear regression and the distribution family is Binomial.

## Data Preparation
We load and shuffle the data. We also add a useless variable to see if the logistic regression removes it.

```{r}
data(iris)
set.seed(100) # for reproducability

x <- iris[sample(1:nrow(iris)),]
x <- cbind(x, useless = rnorm(nrow(x)))
```

We create a binary classification problem by
asking if a flower is of species Virginica or not.
We create new logical variable called `virginica` and remove the 
`Species` column.

```{r}
x$virginica <- x$Species == "virginica"
x$Species <- NULL
```

We can visualize the data using a scatter plot matrix and use the color red for 
`virginica == TRUE` and black for the other flowers.


```{r}
plot(x, col=x$virginica + 1)
```

## A first Logistic Regression Model
Logistic regression is a generalized linear model (GLM) with logit as the
link function and a binomial distribution. The `glm()` function is provided by 
the R core package `stats` which is installed with R and automatically loads
when R is started.

```{r}
model <- glm(virginica ~ .,
  family = binomial(logit), data = x)
```

*About the warning:* glm.fit: fitted probabilities numerically 0 or 1 occurred means that the data is possibly linearly separable.

```{r}
model
```

Check which features are significant?

```{r}
summary(model)
```

AIC ([Akaike information criterion](https://en.wikipedia.org/wiki/Akaike_information_criterion)) 
is a measure of how good the model is. Smaller is better. It can be used for model selection.

The parameter estimates in the coefficients table are log odds. The `*` and `.`
indicate if the effect of the parameter is significantly different from 0.
Positive numbers 
mean that increasing the variable increases the predicted probability 
and negative numbers mean that the probability decreases. For example,
observing a larger Petal.Length increases the predicted probability for the flower to
be of class Virginica. This effect is significant and you can 
verify it in the scatter plot above. For Petal.Length, the red dots have 
larger values than
the black dots.

## Stepwise Variable Selection

Only two variables were flagged as significant. We can remove insignificant 
variables by trying to remove one variable at a time 
as long as the model does not significantly deteriorate (according to the AIC).
This variable selection process is done automatically by the `step()` function.

```{r}
model2 <- step(model, data = x)
summary(model2)
```

The estimates ($\beta_0, \beta_1,...$ ) are
log-odds and can be converted into odds using $exp(\beta)$.
A negative log-odds ratio means that the odds go down with an increase in
the value of the predictor.  A predictor with a
positive log-odds ratio increases the odds. In this case, the odds of
looking at a Virginica iris goes down with Sepal.Width and increases with the
other two predictors.

## Calculate the Response
**Note:** we do here in-sample testing on the data we learned the data
from. To get a generalization error estimate you should use a test set or
cross-validation!

```{r}
pr <- predict(model2, x, type = "response")
round(pr[1:10], 2)
```

The response is the predicted probability of the flower being of species 
Virginica. The probabilities of the first 10 flowers are shown. Below is 
a histogram of predicted probabilities. The color is used to show the 
examples that have the true class Virginica.

```{r}
hist(pr, breaks = 20, main = "Predicted Probability vs. True Class")
hist(pr[x$virginica == TRUE], col = "red", breaks = 20, add = TRUE)
```

## Check Classification Performance

Here we perform in-sample evaluation on the training set. To get an estimate for
generalization error, we should calculate the performance on a held out test set.

The predicted class is calculated by checking if the predicted probability 
is larger than .5.  

```{r }
pred <- pr > .5
```

Now we can create a confusion table and calculate the accuracy.
```{r}
tbl <- table(predicted = pred, actual = x$virginica)
tbl

sum(diag(tbl))/sum(tbl)
```

We can also use caret's more advanced function `caret::confusionMatrix()`. Our code
above uses `logical` vectors. For caret, we need to make sure that both, 
the reference and the predictions are coded as `factor`.

```{r }
caret::confusionMatrix(
  reference = factor(x$virginica, labels = c("Yes", "No"), levels = c(TRUE, FALSE)), 
  data = factor(pred, labels = c("Yes", "No"), levels = c(TRUE, FALSE)))
```

We see that the model performs well with a very high accuracy and kappa value.

## Regularized Logistic Regression

`glmnet::glmnet()` fits generalized linear models (including logistic regression) 
using regularization via penalized maximum likelihood.
The regularization parameter $\lambda$ is a hyperparameter and 
glmnet can use cross-validation to find an appropriate 
value. glmnet does not have a function interface, so we have
to supply a matrix for `X` and a vector of responses for `y`.

```{r }
library(glmnet)

X <- as.matrix(x[, 1:5])
y <- x$virginica

fit <- cv.glmnet(X, y, family = "binomial")
fit
```

There are several selection rules for lambda, we look at the 
coefficients of the logistic regression using the 
lambda that  gives the most regularized model such that the cross-validated error is within one standard error of the minimum cross-validated error.

```{r}
coef(fit, s = fit$lambda.1se)
```

A dot means 0. We see that the predictors Sepal.Length and 
useless are not used in the prediction giving a models similar to
stepwise variable selection above. 

A predict function is provided. We need to specify
what regularization to use and that we want to predict a class
label.

```{r}
predict(fit, newx = X[1:5,], s = fit$lambda.1se, type = "class")
```

Glmnet provides supports many types of 
generalized linear models. Examples can be found in the
article [An Introduction to glmnet](https://glmnet.stanford.edu/articles/glmnet.html).

## Multinomial Logistic Regression

Regular logistic regression predicts only one outcome of a binary event represented 
by two classes. Extending this model to data with more than two classes
is called [multinomial logistic regression](https://en.wikipedia.org/wiki/Multinomial_logistic_regression), 
(or log-linear model). 
A popular implementation uses simple artificial neural networks.
Regular logistic regression is equivalent to a single neuron with a 
sigmoid (i.e., logistic) activation function optimized with cross-entropy loss.
For multinomial logistic regression, one neuron is used for each class and the 
probability distribution is calculated with the softmax activation.
This extension is implemented in `nnet::multinom()`.

```{r}
set.seed(100)
x <- iris[sample(1:nrow(iris)), ]

model <- nnet::multinom(Species ~., data = x)
model
```
We get a $\boldsymbol{\beta}$ vector with weights for two of the three classes. 
The third class is used as the default class with all weights set to 0. This can
be interpreted as comparing the log odds of each of the two classes with the 
default class. A positive number means that increasing the variable makes the 
class more likely and a negative number means the opposite.

Predict the class for the first 5 flowers in the training data. 

```{r}
x[1:5, ]
predict(model, x[1:5,])
```

The package glmnet implements also multinomial logistic regression using
`glmnet(..., family = "multinomial")`.


## Exercises

We will again use the Palmer penguin data for the exercises.

```{r }
library(palmerpenguins)
head(penguins)
```

Create an R markdown document that performs the following:

1. Create a test and a training data set (see section [Holdout Method] in Chapter 3).
2. Create a logistic regression using the training set to predict the variable sex.
2. Use stepwise variable selection. What variables are selected? 
3. What do the parameters for for each of the selected features tell you?
4. Predict the sex of the penguins in the test set. Create a 
  confusion table and calculate the accuracy and discuss how well the model works.
