---
editor_options: 
  markdown: 
    wrap: 72
---

# Logistic Regression*

This chapter introduces the popular classification method 
logistic regression.

### Packages Used in this Chapter {.unnumbered}
This chapter only uses R's base functionality and does not need extra packages.

```{r setup_log_reg, include=FALSE}
source("_common.R")
```

## Introduction

Logistic regression contains the word regression, but it is actually a 
probabilistic, statistical classification
model to predict the probability $p$ of a binary outcome given a set of features.
It is a very powerful model that can be fit very quickly. It is one of the 
first classification models you should try on new data.

Logistic regression is a [generalized linear model](https://en.wikipedia.org/wiki/Generalized_linear_model) 
with the logit as the link function and a binomial error model. It can 
be thought of as a linear regression with the
log odds ratio (logit) 
of the binary outcome as the dependent variable:

$$logit(p) = ln\left(\frac{p}{1-p}\right) = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + ...$$

The logit function links the probability p to the linear regression by converting 
the logit into a number in the probability range $[0,1]$.
```{r}
logit  <- function(p) log(p/(1-p))
p <- seq(0, 1, length.out = 100)
plot(logit(p), p, type = "l")
abline(h = 0.5, lty = 2)
abline(v = 0, lty = 2)
```


This is equivalent to modeling the probability of the outcome $p$ by:

$$ p = \frac{e^{\beta_0 + \beta_1 x_1 + \beta_2 x_2 + ...}}{1 +  e^{\beta_0 + \beta_1 x_1 + \beta_2 x_2 + ...}} = \frac{1}{1+e^{-(\beta_0 + \beta_1 x_1 + \beta_2 x_2 + ...)}}$$

## Data Preparation
We load and shuffle the data. We also add a useless variable to see if the logistic regression removes it.

```{r}
data(iris)
set.seed(100) # for reproducability

x <- iris[sample(1:nrow(iris)),]
x <- cbind(x, useless = rnorm(nrow(x)))
```

We create a binary classification problem by
classify if a flower is of species Virginica.

```{r}
x$virginica <- x$Species == "virginica"
x$Species <- NULL
plot(x, col=x$virginica + 1)
```

## A first Logistic Regression Model
Logistic regression is a generalized linear model (GLM) with logit as the
link function and a binomial error model.

```{r}
model <- glm(virginica ~ .,
  family = binomial(logit), data = x)
```

*About the warning:* glm.fit: fitted probabilities numerically 0 or 1 occurred means that the data is possibly linearly separable.

```{r}
model
```

Check which features are significant?

```{r}
summary(model)
```

AIC ([Akaike information criterion](https://en.wikipedia.org/wiki/Akaike_information_criterion)) 
is a measure of how good the model is. Smaller is better. It can be used for model selection.

## Stepwise Variable Selection

```{r}
model2 <- step(model, data = x)
summary(model2)
```

The estimates ($\beta_0, \beta_1,...$ ) are
log-odds and can be converted into odds using $exp(\beta)$.
A negative log-odds ratio means that the odds go down with an increase in
the value of the predictor.  A predictor with a
positive log-odds ratio increases the odds. In this case, the odds of
looking at a Virginica iris goes down with Sepal.Width and increases with the
other two predictors.

## Calculate the Response
**Note:** we do here in-sample testing on the data we learned the data
from. To get a generalization error estimate you should use a test set or
cross-validation!

```{r}
pr <- predict(model2, x, type = "response")
round(pr[1:10], 2)
```

The response is the predicted probability of the flower being of species 
virginica. The probabilities of the first 10 flowers are shown. Below is 
a histogram of predicted probabilities. The color is used to show the 
examples that have the true class virginica.

```{r}
hist(pr, breaks = 20, main = "Predicted Probability vs. True Class")
hist(pr[x$virginica == TRUE], col = "red", breaks = 20, add = TRUE)
```

## Check Classification Performance

We calculate the predicted class by checking if the probability is larger than
.5. 

```{r }
pred <- pr > .5
```

Now we can create a confusion table and calculate the accuracy.
```{r}
tbl <- table(predicted = pred, actual = x$virginica)
tbl

sum(diag(tbl))/sum(tbl)
```

We can also use caret's more advanced function `caret::confusionMatrix()`. Our code
above uses `logical` vectors. For caret, we need to make sure that both, 
the reference and the predictions are coded as `factor`.

```{r }
caret::confusionMatrix(
  reference = factor(x$virginica, labels = c("Yes", "No"), levels = c(TRUE, FALSE)), 
  data = factor(pred, labels = c("Yes", "No"), levels = c(TRUE, FALSE)))
```

We see that the model performs well with a very high accuracy and kappa value.

## Regularized Logistic Regression

Glmnet fits generalized linear models (including logistic regression) 
using regularization via penalized maximum likelihood.
The regularization parameter $\lambda$ is a hyperparameter and 
glmnet can use cross-validation to find an appropriate 
value. glmnet does not have a function interface, so we have
to supply a matrix for `X` and a vector of responses for `y`.

```{r }
library(glmnet)

X <- as.matrix(x[, 1:5])
y <- x$virginica

fit <- cv.glmnet(X, y, family = "binomial")
fit
```

There are several selection rules for lambda, we look at the 
coefficients of the logistic regression using the 
lambda that  gives the most regularized model such that the cross-validated error is within one standard error of the minimum cross-validated error.

```{r}
coef(fit, s = fit$lambda.1se)
```

A dot means 0. We see that the predictors Sepal.Length and 
useless are not used in the prediction giving a models similar to
stepwise variable selection above. 

A predict function is provided. We need to specify
what regularization to use and that we want to predict a class
label.

```{r}
predict(fit, newx = X[1:5,], s = fit$lambda.1se, type = "class")
```

Glmnet provides supports many types of 
generalized linear models. Examples can be found in the
article [An Introduction to glmnet](https://glmnet.stanford.edu/articles/glmnet.html).

## Exercises

We will again use the Palmer penguin data for the exercises.

```{r }
library(palmerpenguins)
head(penguins)
```

Create an R markdown document that performs the following:

1. Create a test and a training data set (see section [Holdout Method] in Chapter 3).
2. Create a logistic regression using the training set to predict the variable sex.
2. Use stepwise variable selection. What variables are selected? 
3. What do the parameters for for each of the selected features tell you?
4. Predict the sex of the penguins in the test set. Create a 
  confusion table and calculate the accuracy and discuss how well the model works.
