---
editor_options: 
  markdown: 
    wrap: 72
---

# Logistic Regression {#appendix_logistic_regression}

This chapter examines logistic regression in more detail. Logistic regression is introduced
as an alternative classification method in Chapter 4 of Introduction to Data Mining.

## Learning Objectives

After this chapter, you should be able to:

* fit a binary logistic regression model and interpret its coefficients;
* convert model scores into class probabilities and predicted labels;
* assess classification performance on held-out data; and
* recognize when regularized or multinomial logistic regression is appropriate.

## Prerequisites

Read [Classification: Basic Concepts] and [Regression]. Familiarity with
probability, odds, model formulas, and train/test evaluation is assumed.

```{r setup_logreg}
pkgs <- c("glmnet", "caret")

pak::pkg_install(pkgs)
```

The packages used for this chapter are: 

`r format_pkgs(pkgs)`

```{r setup_logreg-2, include=FALSE}
source("_common.R")
source("format_pkgs.R")
all_pkgs <- union(all_pkgs, pkgs)
```

## Introduction

Despite its name, logistic regression is a classification model for estimating
the probability $p$ of a binary outcome from a set of features. It is a common
baseline because it is interpretable and often fits quickly.

Logistic regression is a [generalized linear model](https://en.wikipedia.org/wiki/Generalized_linear_model)
with the logit link and a binomial response distribution. It models the log
odds of the binary outcome as a linear function of the predictors:

$$logit(p) = ln\left(\frac{p}{1-p}\right) = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + ...$$

The logit function maps a probability $p$ in $(0,1)$ to the real line
$(-\infty,+\infty)$.

```{r}
logit  <- function(p) log(p/(1-p))
p <- seq(0, 1, length.out = 100)
plot(logit(p), p, type = "l")
abline(h = 0.5, lty = 2)
abline(v = 0, lty = 2)
```

The plotted curve is the inverse logit, also called the [logistic or sigmoid
function](https://en.wikipedia.org/wiki/Logistic_function). It maps real-valued
linear predictors to probabilities in $(0,1)$.
Using the inverse function, we see that the probability of the outcome $p$ 
is modeled by the logistic function of the linear regression:

$$ p = \frac{e^{\beta_0 + \beta_1 x_1 + \beta_2 x_2 + ...}}{1 +  e^{\beta_0 + \beta_1 x_1 + \beta_2 x_2 + ...}} = \frac{1}{1+e^{-(\beta_0 + \beta_1 x_1 + \beta_2 x_2 + ...)}} = \sigma(\beta_0 + \beta_1 x_1 + \beta_2 x_2 + ...)$$

After estimating $\boldsymbol{\beta} = (\beta_0, \beta_1,...)$ from training
data by maximizing the binomial likelihood (equivalently, minimizing log loss),
the equation above predicts
the probability $p$ for a new data point $\mathbf{x} = (x_1, x_2, ...)$. With a
default threshold of 0.5, we predict the event when $p > .5$ and otherwise
predict that it does not occur.

Each binary outcome is modeled with a Bernoulli distribution. Across multiple
observations, the number of positive outcomes has a binomial distribution.
Logistic regression uses the logit link to relate the event probability to the
linear predictor.

## Data Preparation
We load and shuffle the data. We also add a useless variable to see if the logistic regression removes it.

```{r}
data(iris)
set.seed(100) # for reproducibility

x <- iris[sample(1:nrow(iris)),]
x <- cbind(x, useless = rnorm(nrow(x)))
```

We create a binary classification problem by
asking if a flower is of species Virginica or not.
We create new logical variable called `virginica` and remove the 
`Species` column.

```{r}
x$virginica <- x$Species == "virginica"
x$Species <- NULL
```

We can visualize the data using a scatter plot matrix and use the color red for 
`virginica == TRUE` and black for the other flowers.


```{r}
plot(x, col=x$virginica + 1)
```

## A first Logistic Regression Model
Logistic regression is a generalized linear model (GLM) with logit as the
link function and a binomial distribution. The `glm()` function is provided by 
the R core package `stats` which is installed with R and automatically loads
when R is started.

```{r}
model <- glm(virginica ~ .,
  family = binomial(logit), data = x)
```

*About the warning:* glm.fit: fitted probabilities numerically 0 or 1 occurred means that the data is possibly linearly separable.

```{r}
model
```

Check which features are significant?

```{r}
summary(model)
```

AIC ([Akaike information criterion](https://en.wikipedia.org/wiki/Akaike_information_criterion)) 
is a measure of how good the model is. Smaller is better. It can be used for model selection.

The parameter estimates in the coefficients table are log odds. The `*` and `.`
indicate if the effect of the parameter is significantly different from 0.
Positive numbers 
mean that increasing the variable increases the predicted probability 
and negative numbers mean that the probability decreases. For example,
observing a larger Petal.Length increases the predicted probability for the flower to
be of class Virginica. This effect is significant and you can 
verify it in the scatter plot above. For Petal.Length, the red dots have 
larger values than
the black dots.

## Stepwise Variable Selection

Only two variables were flagged as significant. We can remove insignificant 
variables by trying to remove one variable at a time 
as long as the model does not significantly deteriorate (according to the AIC).
This variable selection process is done automatically by the `step()` function.

```{r}
model2 <- step(model, data = x)
summary(model2)
```

The estimates ($\beta_0, \beta_1,...$ ) are
log-odds and can be converted into odds using $exp(\beta)$.
A negative log-odds ratio means that the odds go down with an increase in
the value of the predictor.  A predictor with a
positive log-odds ratio increases the odds. In this case, the odds of
looking at a Virginica iris goes down with Sepal.Width and increases with the
other two predictors.

## Calculate the Response
**Note:** we do here in-sample testing on the data we learned the data
from. To get a generalization error estimate you should use a test set or
cross-validation!

```{r}
pr <- predict(model2, x, type = "response")
round(pr[1:10], 2)
```

The response is the predicted probability of the flower being of species 
Virginica. The probabilities of the first 10 flowers are shown. Below is 
a histogram of predicted probabilities. The color is used to show the 
examples that have the true class Virginica.

```{r}
hist(pr, breaks = 20, main = "Predicted Probability vs. True Class")
hist(pr[x$virginica == TRUE], col = "red", breaks = 20, add = TRUE)
```

## Illustrate In-Sample Classification Performance

Here we perform in-sample evaluation on the training set. To get an estimate for
generalization error, we should calculate the performance on a held out test set.

The predicted class is calculated by checking if the predicted probability 
is larger than .5.  

```{r }
pred <- pr > .5
```

Now we can create a confusion table and calculate the accuracy.
```{r}
tbl <- table(predicted = pred, actual = x$virginica)
tbl

sum(diag(tbl))/sum(tbl)
```

We can also use caret's more advanced function `caret::confusionMatrix()`. Our code
above uses `logical` vectors. For caret, we need to make sure that both, 
the reference and the predictions are coded as `factor`.

```{r }
caret::confusionMatrix(
  reference = factor(x$virginica, labels = c("Yes", "No"), levels = c(TRUE, FALSE)), 
  data = factor(pred, labels = c("Yes", "No"), levels = c(TRUE, FALSE)))
```

These in-sample accuracy and kappa values describe performance on the data used
to fit the model and are likely optimistic. Use the held-out test set, as
discussed in the previous chapter, to estimate performance on new observations.

## Regularized Logistic Regression

`glmnet::glmnet()` fits generalized linear models (including logistic regression) 
using regularization via penalized maximum likelihood.
The regularization parameter $\lambda$ is a hyperparameter and 
glmnet can use cross-validation to find an appropriate 
value. glmnet does not have a function interface, so we have
to supply a matrix for `X` and a vector of responses for `y`.

```{r }
library(glmnet)

X <- as.matrix(x[, 1:5])
y <- x$virginica

fit <- cv.glmnet(X, y, family = "binomial")
fit
```

There are several selection rules for lambda, we look at the 
coefficients of the logistic regression using the 
lambda that  gives the most regularized model such that the cross-validated error is within one standard error of the minimum cross-validated error.

```{r}
coef(fit, s = fit$lambda.1se)
```

A dot means 0. We see that the predictors Sepal.Length and 
useless are not used in the prediction giving a models similar to
stepwise variable selection above. 

A predict function is provided. We need to specify
what regularization to use and that we want to predict a class
label.

```{r}
predict(fit, newx = X[1:5,], s = fit$lambda.1se, type = "class")
```

Glmnet provides supports many types of 
generalized linear models. Examples can be found in the
article [An Introduction to glmnet](https://glmnet.stanford.edu/articles/glmnet.html).

## Multinomial Logistic Regression

Regular logistic regression predicts only one outcome of a binary event represented 
by two classes. Extending this model to data with more than two classes
is called [multinomial logistic regression](https://en.wikipedia.org/wiki/Multinomial_logistic_regression), 
(or log-linear model). 
A popular implementation uses simple artificial neural networks.
Regular logistic regression is equivalent to a single neuron with a 
sigmoid (i.e., logistic) activation function optimized with cross-entropy loss.
For multinomial logistic regression, one neuron is used for each class and the 
probability distribution is calculated with the softmax activation.
This extension is implemented in `nnet::multinom()`.

```{r}
set.seed(100)
x <- iris[sample(1:nrow(iris)), ]

model <- nnet::multinom(Species ~., data = x)
model
```
We get a $\boldsymbol{\beta}$ vector with weights for two of the three classes. 
The third class is used as the default class with all weights set to 0. This can
be interpreted as comparing the log odds of each of the two classes with the 
default class. A positive number means that increasing the variable makes the 
class more likely and a negative number means the opposite.

Predict the class for the first 5 flowers in the training data. 

```{r}
x[1:5, ]
predict(model, x[1:5,])
```

The package glmnet implements also multinomial logistic regression using
`glmnet(..., family = "multinomial")`.


## Exercises

We will again use the Palmer penguin data for the exercises.

```{r }
library(palmerpenguins)
head(penguins)
```

Create an R markdown document that performs the following:

1. Create a test and a training data set (see section [Holdout Method] in Chapter 3).
2. Create a logistic regression using the training set to predict the variable sex.
2. Use stepwise variable selection. What variables are selected? 
3. What do the parameters for for each of the selected features tell you?
4. Predict the sex of the penguins in the test set. Create a 
  confusion table and calculate the accuracy and discuss how well the model works.

**A good answer should** state how the training and test sets were formed and
keep variable selection within the training data. Identify the selected
predictors and interpret their coefficients with the outcome coding and
reference levels made clear. Use the test confusion table to discuss both
correct and incorrect predictions; interpret accuracy in light of the class
distribution.

## Chapter Summary

Logistic regression models the probability of a categorical outcome through a
linear predictor and a link function. Coefficients describe changes in log
odds, while predicted probabilities can be converted into labels using a
decision threshold. Evaluate predictions on held-out data, and use
regularization when many or correlated predictors make the fit unstable.

## Key Terms

* **Logit:** the logarithm of the odds of an event.
* **Odds:** the probability of an event divided by the probability it does not
  occur.
* **Logistic function:** a transformation that maps a real-valued score to a
  probability between 0 and 1.
* **Odds ratio:** the multiplicative change in odds associated with a one-unit
  predictor change, holding other terms fixed.
* **Regularization:** a penalty on coefficient size used to improve stability
  and reduce overfitting.
