---
editor_options: 
  markdown: 
    wrap: 72
---

# Logistic Regression*

This chapter only uses R's base functionality and does not need extra packages.

```{r setup_log_reg, include=FALSE}
source("_common.R")
```

## Introduction

Logistic regression contains the word regression, but it is actually a 
probabilistic statistical classification
model to predict a binary outcome (a probability) given a set of features.
It is a very powerful model that can be fit very quickly. It is one of the 
first classification models you should try on new data.


Logistic regression can be thought of as a linear regression with the
log odds ratio (logit) 
of the binary outcome as the dependent variable:

$$logit(p) = ln(\frac{p}{1-p}) = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + ...$$


```{r}
logit  <- function(p) log(p/(1-p))
x <- seq(0, 1, length.out = 100)
plot(x, logit(x), type = "l")
abline(v=0.5, lty = 2)
abline(h=0, lty = 2)
```


This is equivalent to modeling the probability of the outcome $p$ by

$$ p = \frac{e^{\beta_0 + \beta_1 x_1 + \beta_2 x_2 + ...}}{1 +  e^{\beta_0 + \beta_1 x_1 + \beta_2 x_2 + ...}} = \frac{1}{1+e^{-(\beta_0 + \beta_1 x_1 + \beta_2 x_2 + ...)}}$$

## Data Preparation
Load and shuffle data. We also add a useless variable to see if the logistic regression removes it.

```{r}
data(iris)
x <- iris[sample(1:nrow(iris)),]
x <- cbind(x, useless = rnorm(nrow(x)))
```

Make Species into a binary classification problem so we will
classify if a flower is of species Virginica

```{r}
x$virginica <- x$Species == "virginica"
x$Species <- NULL
plot(x, col=x$virginica+1)
```

## Create a Logistic Regression Model
Logistic regression is a generalized linear model (GLM) with logit as the
link function and a binomial error model.

```{r}
model <- glm(virginica ~ .,
  family = binomial(logit), data=x)
```

About the warning: glm.fit: fitted probabilities numerically 0 or 1 occurred means that the data is possibly linearly separable.

```{r}
model
```

Check which features are significant?

```{r}
summary(model)
```

AIC can be used for model selection

## Stepwise Variable Selection

```{r}
model2 <- step(model, data = x)
summary(model2)
```

The estimates ($\beta_0, \beta_1,...$ ) are
log-odds and can be converted into odds using $exp(\beta)$.
A negative log-odds ratio means that the odds go down with an increase in
the value of the predictor.  A predictor with a
positive log-odds ratio increases the odds. In this case, the odds of
looking at a Virginica iris goes down with Sepal.Width and increases with the
other two predictors.

## Calculate the Response
**Note:** we do here in-sample testing on the data we learned the data
from. To get a generalization error estimate you should use a test set or
cross-validation!

```{r}
pr <- predict(model2, x, type="response")
round(pr, 2)
hist(pr, breaks=20)
hist(pr[x$virginica==TRUE], col="red", breaks=20, add=TRUE)
```

## Check Classification Performance

We calculate the predicted class by checking if the probability is larger than
.5. 

```{r }
pred <- pr > .5
```

Now er can create a confusion table and calculate the accuracy.
```{r}
tbl <- table(actual = x$virginica, predicted = pr>.5)
tbl

sum(diag(tbl))/sum(tbl)
```

We can also use caret's more advanced function `confusionMatrix()`. Our code
above uses `logical` vectors.
but foo caret, we need to make sure that both, the reference and the predictions
are coded as `factor`.

```{r }
caret::confusionMatrix(
  reference = factor(x$virginica, levels = c(TRUE, FALSE)), 
  data = factor(pr>.5, levels = c(TRUE, FALSE)))
```

We see that the model performs well with a very high accuracy and kappa value.


## Exercises

We will again use the Palmer penguin data for the exercises.

```{r }
library(palmerpenguins)
head(penguins)
```

Create an R markdown document that performs the following:

1. Create a test and a training data set (see section [Holdout Method] in Chapter 3).
2. Create a logistic regression using the training set to predict the variable sex.
2. Use stepwise variable selection. What variables are selected? 
3. What do the parameters for for each of the selected features tell you?
4. Predict the sex of the penguins in the test set. Create a 
  confusion table and calculate the accuracy and discuss how well the model works.
