# Association Analysis: Basic Concepts and Algorithms


```{r setup_05, include=FALSE}
pkgs <- sort(c('tidyverse', 'arules', 'arulesViz', 'mlbench'))

lapply(pkgs, function(pkg) {
  if (system.file(package = pkg) == '') install.packages(pkg)
})

all_pkgs <- union(all_pkgs, pkgs)
```

**Packages used for this chapter:** `r format_pkgs(pkgs)`

You can read the free sample chapter from the textbook [@Tan2005]:
[Chapter 5. Association Analysis: Basic Concepts and Algorithms](https://www-users.cs.umn.edu/~kumar001/dmbook/ch5_association_analysis.pdf)

## The arules Package

Association rule mining in R is implemented in the package `arules`.

```{r }
library(tidyverse)
library(arules)
library(arulesViz)
```

For information about the `arules` package try:
`help(package="arules")`
and
`vignette("arules")` (also available at [CRAN](http://cran.r-project.org/web/packages/arules/vignettes/arules.pdf))

arules uses the S4 object system to implement classes and methods.
Standard R objects use the [S3 object system](http://adv-r.had.co.nz/S3.html)
which do not use formal class definitions and are usually implemented
as a list with a class attribute.
`arules` and many other R packages use the
[S4 object system](http://adv-r.had.co.nz/S4.html) which is based on
formal class definitions with member variables and methods 
(similar to object-oriented programming languages like Java and C++).
Some important differences of using S4 objects compared to the usual S3
objects are:

* coercion (casting): `as(from, "class_name")`
* help for classes: `class? class_name`



## Transactions
### Create Transactions

We will use the Zoo dataset from `mlbench`.

```{r }
data(Zoo, package = "mlbench")
head(Zoo)
```

The data in the data.frame need to be converted into a set of transactions where each row represents a transaction and each column is translated into items.
This is done using the contructor `transactions()`.
For the Zoo data set this means that we consider animals as transactions
and the different traits (features) will become items that each animal has. For
example the animal _antelope_ has the item _hair_ in its transaction.

```{r }
trans <- transactions(Zoo)
```


The conversion gives a warning because only discrete features (`factor` and `logical`) can be
directly translated into items. Continuous features need to be discretized first.

What is column 13?

```{r }
summary(Zoo[13])
ggplot(Zoo, aes(legs)) + geom_histogram()
table(Zoo$legs)
```

Possible solution: Make legs into has/does not have legs

```{r }
Zoo_has_legs <- Zoo %>% mutate(legs = legs > 0)
ggplot(Zoo_has_legs, aes(legs)) + geom_bar()
table(Zoo_has_legs$legs)
```

__Alternatives:__

* use each unique value as an item:

```{r }
Zoo_unique_leg_values <- Zoo %>% mutate(legs = factor(legs))
head(Zoo_unique_leg_values$legs)
```

* discretize (see [`? discretize`](https://www.rdocumentation.org/packages/arules/topics/discretize) and
[discretization in the code for Chapter 2](chap2.html#discretize-features)):

```{r }
Zoo_discretized_legs <- Zoo %>% mutate(
  legs = discretize(legs, breaks = 2, method="interval")
)
table(Zoo_discretized_legs$legs)
```


 Convert data into a set of transactions

```{r }
trans <- transactions(Zoo_has_legs)
trans
```

### Inspect Transactions

```{r }
summary(trans)
```

Look at created items. They are still called column names since the transactions are actually stored as a large sparse logical matrix (see below).

```{r }
colnames(trans)
```

Compare with the original features (column names) from Zoo

```{r }
colnames(Zoo)
```

Look at a (first) few transactions as a matrix. 1 indicates the presence of an item.

```{r }
as(trans, "matrix")[1:3,]
```

Look at the transactions as sets of items

```{r }
inspect(trans[1:3])
```

Plot the binary matrix. Dark dots represent 1s.

```{r }
image(trans)
```

Look at the relative frequency (=support) of items in the data set. Here we look at the 10 most frequent items.

```{r }
itemFrequencyPlot(trans,topN = 20)

ggplot(
  tibble(
    Support = sort(itemFrequency(trans, type = "absolute"), decreasing = TRUE),
    Item = seq_len(ncol(trans))
  ), aes(x = Item, y = Support)) + geom_line()
```

__Alternative encoding:__ Also create items for FALSE (use factor)

```{r }
sapply(Zoo_has_legs, class)
Zoo_factors <- Zoo_has_legs %>% mutate_if(is.logical, factor)
sapply(Zoo_factors, class)
summary(Zoo_factors)

trans_factors <- transactions(Zoo_factors)
trans_factors

itemFrequencyPlot(trans_factors, topN = 20)

## Select transactions that contain a certain item
trans_insects <- trans_factors[trans %in% "type=insect"]
trans_insects
inspect(trans_insects)
```

### Vertical Layout (Transaction ID Lists)

The default layout for transactions is horizontal layout (i.e. each transaction is a row).
The vertical layout represents transaction data as a list of transaction IDs for each item (= transaction ID lists).

```{r }
vertical <- as(trans, "tidLists")
as(vertical, "matrix")[1:10, 1:5]
```

## Frequent Itemsets
### Mine Frequent Itemsets

For this dataset we have already a huge number of possible itemsets

```{r }
2^ncol(trans)
```

Find frequent itemsets (target="frequent") with the default settings.

```{r }
its <- apriori(trans, parameter=list(target = "frequent"))
its
```

Default minimum support is .1 (10\%).
__Note:__ We use here a very small data set. For larger datasets
the default minimum support might be to low and you may run out of memory. You probably want to start out with a higher minimum support like
.5 (50\%) and then work your way down.

```{r }
5/nrow(trans)
```

In order to find itemsets that effect 5 animals I need to go down to a
support of about 5\%.

```{r }
its <- apriori(trans, parameter=list(target = "frequent", support = 0.05))
its
```

Sort by support

```{r }
its <- sort(its, by = "support")
inspect(head(its, n = 10))
```

Look at frequent itemsets with many items (set breaks manually since
Automatically chosen breaks look bad)

```{r }
ggplot(tibble(`Itemset Size` = factor(size(its))), aes(`Itemset Size`)) + geom_bar()
inspect(its[size(its) > 8])
```

### Concise Representation of Itemsets

Find maximal frequent itemsets (no superset if frequent)

```{r }
its_max <- its[is.maximal(its)]
its_max
inspect(head(its_max, by = "support"))
```

Find closed frequent itemsets (no superset if frequent)

```{r }
its_closed <- its[is.closed(its)]
its_closed
inspect(head(its_closed, by = "support"))

counts <- c(
  frequent=length(its),
  closed=length(its_closed),
  maximal=length(its_max)
)

ggplot(as_tibble(counts, rownames = "Itemsets"),
  aes(Itemsets, counts)) + geom_bar(stat = "identity")
```



## Association Rules
### Mine Association Rules

We use the APRIORI algorithm (see [`? apriori`](https://www.rdocumentation.org/packages/arules/topics/apriori))

```{r }
rules <- apriori(trans, parameter = list(support = 0.05, confidence = 0.9))
length(rules)

inspect(head(rules))
quality(head(rules))
```

Look at rules with highest lift

```{r }
rules <- sort(rules, by = "lift")
inspect(head(rules, n = 10))
```

Create rules using the alternative encoding (with "FALSE" item)

```{r }
r <- apriori(trans_factors)
r
print(object.size(r), unit = "Mb")

inspect(r[1:10])
inspect(head(r, n = 10, by = "lift"))
```

### Calculate Additional Interest Measures

```{r }
interestMeasure(rules[1:10], measure = c("phi", "gini"),
  trans = trans)
```

Add measures to the rules

```{r }
quality(rules) <- cbind(quality(rules),
  interestMeasure(rules, measure = c("phi", "gini"),
    trans = trans))
```

Find rules which score high for Phi correlation

```{r }
inspect(head(rules, by = "phi"))
```

### Mine Using Templates

Sometimes it is beneficial to specify what items should be where in the rule. For apriori we can use the parameter appearance to specify this (see [`? APappearance`](https://www.rdocumentation.org/packages/arules/topics/APappearance)). In
the following we restrict rules to an animal `type` in the RHS and any item in
the LHS.

```{r }
type <- grep("type=", itemLabels(trans), value = TRUE)
type

rules_type <- apriori(trans, appearance= list(rhs = type))

inspect(head(sort(rules_type, by = "lift")))
```

Saving rules as a CSV-file to be opened with Excel or other tools.

`write(rules, file = "rules.csv", quote = TRUE)`

## Association Rule Visualization

```{r }
library(arulesViz)
```

Default scatterplot

```{r }
plot(rules)
```

Note that some jitter (randomly move points) was added to show how many rules have the
same confidence and support value. Without jitter:

```{r }
plot(rules, control = list(jitter = 0))

plot(rules, shading = "order")
##plot(rules, interactive = TRUE)
```

Grouped plot

```{r }
plot(rules, method = "grouped")
##plot(rules, method = "grouped", engine = "interactive")
```

As a graph

```{r }
plot(rules, method = "graph")
plot(head(rules, by = "phi", n = 100), method = "graph")
```


## Interactive Visualizations

We will use the association rules mined from the Iris dataset for the following examples.

```{r }
data(iris)
summary(iris)
```

Convert the data to transactions. Note that the features are numeric and need to be discretized.
The conversion automatically applies frequency-based discretization with 3 classes to each numeric
feature (with a warning).

```{r }
iris_trans <- transactions(iris)
inspect(head(iris_trans))
```

Next, we mine association rules.

```{r }
rules <- apriori(iris_trans, parameter = list(support = 0.1, confidence = 0.8))
rules
```

### Interactive Inspect With Sorting, Filtering and Paging

```{r }
inspectDT(rules)
```

### Scatter Plot 

Plot rules as a scatter plot using an interactive html widget. To avoid overplotting,
jitter is added automatically. Set `jitter = 0` to disable jitter. Hovering over rules shows 
rule information.
_Note:_ plotly/javascript does not do well with too many points, so plot
selects the top 1000 rules with a warning if more rules are supplied.

```{r }
plot(rules, engine = "html")
```

### Matrix Visualization

Plot rules as a matrix using an interactive html widget. 

```{r }
plot(rules, method = "matrix", engine = "html") 
```

### Visualization as Graph
 
Plot rules as a graph using an interactive html widget.
_Note:_  the used javascript library does not do well with too many graph nodes, so plot selects the top 100 rules only (with a warning).

```{r }
plot(rules, method = "graph", engine = "html")
```

### Interactive Rule Explorer

You can specify a rule set or a dataset. To explore rules that can be mined from iris, use:
`ruleExplorer(iris)`

The rule explorer creates an interactive Shiny application that can be used locally or 
deployed on a server for sharing. A deployed version of the ruleExplorer is available 
[here](https://mhahsler-apps.shinyapps.io/ruleExplorer_demo/) (using [shinyapps.io](https://www.shinyapps.io/)).
